### 2018-01-05 Submissao/big.tex
(+) Documentos da submissao do projeto.

\documentclass[12pt,english]{amsart}
%\usepackage[margin=0.5in]{geometry}
\usepackage[a4paper, margin=2cm]{geometry}
\linespread{1.5}

%\usepackage{fontspec}
%\setmainfont{Times New Roman}
\usepackage{mathptmx}

\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{lscape}
\usepackage{rotating}
\usepackage{epstopdf}

\usepackage{verbatim}
\usepackage{url}
\usepackage{xspace}

\usepackage{listings}
\usepackage{color}
    \definecolor{light}{gray}{0.97}
    \definecolor{dark}{gray}{0.30}
\lstset{
%columns=fullflexible,
%basicstyle=\ttfamily,
escapeinside={||},
    %mathescape=true,
    language=C, % choose the language of the code
    basicstyle=\fontfamily{pcr}\selectfont\scriptsize\color{black},
    keywordstyle=\color{black}\bfseries, % style for keywords
    numbers=none, % where to put the line-numbers
    numberstyle=\tiny, % the size of the fonts that are used for the line-numbers
    backgroundcolor=\color{light},
    showspaces=false, % show spaces adding particular underscores
    showstringspaces=false, % underline spaces within strings
    showtabs=false, % show tabs within strings adding particular underscores
    %frame=single, % adds a frame around the code
    tabsize=2, % sets default tabsize to 2 spaces
    %rulesepcolor=\color{gray}
    captionpos=b, % sets the caption-position to bottom
    breaklines=false, % sets automatic line breaking
    %breakatwhitespace=false,
    numbersep=2em,
    % C was used in the blocksworld example to refer to block C and nowhere else
    emph={par,or,hor,do,end,loop,code,await,pause,emit,input,event,call,with,%
          var,and,then,else,return,pure,deterministic,nohold,finalize,%
          class, every, FOREVER, this, spawn, in, pool, watching, until, 
          interface, each, abort, when, signal, PROC, CHAN, SIGNAL, PAR, not,
          bool, tag, escape, traverse,implementation,output,true,false,
          native,@const,@pure,@safe,define,public,private,none},
    emphstyle={\bfseries},
    commentstyle=\color{dark}\scriptsize,
    %xleftmargin=20pt,
    %xrightmargin=20pt,
    framesep=20pt,
    %upquote=true,
    %aboveskip={1.5\baselineskip},
}

\newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
\newcommand{\code}[1] {{\small{\texttt{#1}}}}

\usepackage{enumitem}
\setlist{nolistsep}

% Auto-Standby
\title{Energy Efficiency for IoT Software in the Large
        } %\\ \large{Research project proposal submitted to the Instituto Serrapilheira}}

%\author{Anonymous Author}

\begin{document}

\date{}
\maketitle

\vspace{-0.5cm}
\begin{comment}
\abstract{
The annual power consumption of network-connected devices worldwide is
estimated to be greater than Germany's total electricity consumption.
By 2020, there will be 50 billion ``things'' connected to the \emph{Internet of
Things (IoT)}.
%
Currently, most of the energy to power these devices is wasted in \emph{standby
mode}, when they are not performing their main functionality.
%
Considering this alarming scenario, the International Energy Agency (IEA)
launched an initiative to ensure that IoT devices consume a minimal amount of
energy while in standby mode.
%
IEA estimates that it is possible to save 65\% of the energy used by
network-enabled devices with existing technologies.

This research project aims to address the software challenges, as determined by
IEA, towards an energy-efficient IoT:
    to ensure that devices power down to lowest possible consumption modes,
    and that they remain there for longest possible periods of time.
We propose to investigate transparent mechanisms in a programming language that
ensure deep and long periods of standby, meaning that \emph{all} programs ever
written in this language would benefit from low-power modes automatically,
without extra programming efforts.
}
\end{comment}

%\newpage

\section{INTRODUCTION}
\label{sec.introduction}

According to the International Energy Agency (IEA)~\cite{iea.data}, there were
around 14 billion traditional network-connected devices in 2013 (e.g., mobile
phones and smart TVs).
This number is expected to increase to 50 billion by 2020 with
the proliferation of IoT devices (e.g., smart bulbs and fitness wearables).
%
IoT and traditional network-connected devices already outnumber people on the
planet by a factor of two, and the amount of data traffic is expected to grow
at an exponential rate in the next years.
%By 2050, the energy demands of networked devices should exceed the current
%consumption of Germany and Canada combined, corresponding to 6\% of current
%global electricity consumption.
%
However, most of the energy to power these devices is consumed while they are
in \emph{standby mode}, i.e., when their software is neither transmitting or
processing data.
%
The annual $CO_2$ emissions related to standby in Australia are equivalent to
those of 1 million cars.
%
%This undesirable inefficiency opens opportunities for improvements in energy
%savings closer to 65\% only considering currently available technologies.
%
The projected growth of IoT devices together with the surprising effects of
standby consumption, made network standby efficiency one of the six
pillars of G20's \emph{Energy Efficiency Action Plan}%
\footnote{G20's Energy Efficiency Action Plan: \url{https://www.iea-4e.org/projects/g20}}.

Many other organizations have also reported the importance of energy savings
specifically in the context of the IoT and networked devices.
%
In 2010, the Internet Engineering Task Force (IETF) created a working group on energy
management (EMAN) with the following statement:
``Energy management is becoming an additional requirement for network
management systems due to several factors including the rising and
fluctuating energy costs, the increased awareness of the ecological
impact of operating networks and devices, and the regulation of
governments on energy consumption and production''~\cite{ietf.eman}.
%
The American Council for an Energy-Efficient Economy (ACEEE) states that
``The potential for new energy efficiency remains enormous, (...) looking
ahead we must take a systems-based approach to dramatically scale up energy
efficiency.
%Intelligent efficiency is a systems-based approach to efficiency that can help
%to meet this need.
(...) intelligent efficiency differs from component energy efficiency in that
it is adaptive, anticipatory, and networked.''~\cite{aceee.1}
%
The pioneer \emph{ENERGY STAR} program~\cite{energystar} states that
``networked devices and networking equipment, as integrated systems, have the
potential to contribute significant net energy savings''.
% \footnote{ENERGY STAR program \url{https://www.energystar.gov/index.cfm?c=prod_development.prod_development_epa_workshop}}
%
The US Environmental Protection Agency (EPA) and Information Technology
Industry Council (ITI) held a workshop to explore roles for the \emph{ENERGY
STAR} and others program in promoting savings through intelligent energy
efficiency~\cite{aceee.2}.
%, which differs from individualized
%component-based energy efficiency in that it is adaptive, anticipatory, and
%networked.

With regard to concrete initiatives, IEA's \emph{Electronic Devices and Networks}
focuses specifically on the issue of networked device standby%
\footnote{EDNA initiative: \url{https://edna.iea-4e.org/}}.
%It is one of the six fronts of G20's \emph{Energy Efficiency Action Plan}%
%\footnote{\url{https://www.iea-4e.org/projects/g20}}.
%Network standby refers to the minimum power a device requires to maintain its
%connection when not communicating or computing actively.
ACEEE's \emph{Intelligent Efficiency} promotes a systematic approach
to optimize the behavior of cooperating devices in order to achieve energy
savings as a whole.
Both approaches, device and system based, involve mostly software solutions,
since energy savings are dynamic policies that depend on the application
demands and device battery levels at a given moment in time.
%
There is also a number of low-power wireless standards for the IoT
infrastructure that suit different needs of range, throughput, and physical
distribution~\cite{iot.energy.2}.
For instance, \emph{Bluetooth Low-Energy (BLE)} is a replacement for classic
Bluetooth and is designed for lower data throughput in personal area
networks (PANs).
\emph{6LoWPAN} adapts the IPv6 internet standard to low-power and
limited-processing devices.
%
These technologies make radio transmissions more efficient,
support flexible network topologies, and reduce data traffic considerably.
They also enable sleep modes that cut energy consumption to minimum levels.
%
However, these technologies ultimately require software to control their
functionalities and to wisely switch between standby and active modes in order
to build an energy-efficient IoT.

\subsection{Objectives}
\label{sec.goals}

%The annual power consumption of network-connected devices worldwide is
%estimated to be greater than Germany's total electricity consumption.
%By 2020, there will be 50 billion ``things'' connected to the \emph{Internet of
%Things (IoT)}.
%
%Considering this alarming scenario, the International Energy Agency (IEA)
%launched an initiative to ensure that IoT devices consume a minimal amount of
%energy while in standby mode.

Effective use of low-power standby will play a fundamental role in energy
efficiency for the expected 50 billion IoT devices by 2020~\cite{iea.data}.
In order to succeed in this challenge, new solutions have to scale to the
forthcoming mass of IoT software that must use standby modes effectively.
%
This research project aims to address the software challenges, as determined by
IEA, towards an energy-efficient IoT~\cite{iea.data}:
%
\begin{itemize}
    \item To ensure that devices employ the lowest possible modes of standby consumption.
    \item To ensure that devices remain in longest possible periods of standby time.
\end{itemize}

Given the projected scale of the IoT and the role of low-power standby towards
energy efficiency, this research project has the following goals:

\begin{enumerate}
    \item Address energy efficiency through extensive use of standby.
    \item Target constrained embedded architectures that form the IoT.
    \item Provide standby mechanisms at the programming language level that scale to all applications.
    \item Support transparent/non-intrusive standby mechanisms that reduce barriers of adoption.
\end{enumerate}

% TODO: goal 1 and 3 are also the limitations: only standby, not holistic energy savings

This proposal lies at the bottom of the software development
layers---transparent programming language mechanisms---meaning that \emph{all}
applications would take advantage of low-power standby modes automatically,
without extra programming efforts.
%
We expect that existing energy-unaware applications will benefit from savings
in the order of 50\% based on IEA's estimates considering current standby
technologies~\cite{iea.data}, and previous third-party work on transparent energy
awareness~\cite{wsn.tos.2}.

\subsection{State of the Art}
\label{sec.related}

Many energy-aware languages, extensions, and operating systems have been
proposed recently.
%
Some proposals adjust QoS (quality of service) to reduce power consumption
(e.g., accuracy of computations, data sampling rates, scheduling times,
etc.)~\cite{os.ecosystem,os.odyssey.2,lang.green,lang.enerj,os.jouleguard,lang.lab,lang.flexjava,lang.greenweb}.
%
Other proposals offer mechanisms to switch behaviors depending on application
demands and battery levels (e.g., disable functionalities, switch UI,
etc.)~\cite{lang.eon,lang.energytypes,lang.gradual,lang.ent}.
%
None of these initiatives are concerned with taking advantage of standby modes
when idle,
but more with adapting or eliminating computations while in active modes (not
addressing goal 1 of Section~\ref{sec.goals}).
%
There are also specialized network protocols that make devices remain in
low-power standby modes for longer periods of time~\cite{wsn.energy}.
%
However, protocol-based initiatives only apply to the networked parts of
applications and have to be programmed carefully to take
advantage of standby modes (not addressing goals 3 and 4).
%(which is typically the most power hungry subsystem),
%
%This proposal is also unique in the aim to provide \emph{transparent
%mechanisms} such that there are no explicit programming efforts to make
%applications more energy efficient (goal 4 above).

\begin{comment}
\emph{Green}~\cite{lang.green} enables applications to approximate expensive
computations based on QoS constraints specified by the programmer.
%
\emph{GreenWeb}~\cite{lang.greenweb} provides energy-aware HTML extensions to
specify expected levels of QoS related to a webpage responsiveness.
The language runtime determines how to deliver the target QoS experience while
minimizing energy consumption.
%
\emph{Energy Types} and \emph{ENT} proposes a type system to specify phases
(i.e., workload demands) and modes (i.e., expected energy state).
The language runtime monitors the current mode and only executes admissible
phases.
%
\emph{EnerJ}~\cite{lang.enerj} also proposes a type system in which primitive
types (e.g., floats and integers) can be annotated as \emph{approximate}.
The language then uses non-exact but more energy efficient operations on these
types.
%
Unlocking energy
    - locks
        - busy wait vs sleep
        - CPU sleep mode

MEANTIME: achieving both minimal energy and timeliness with approximate computing
    - approximate computing
\end{comment}

Despite the advances in energy-aware languages, embedded systems are still
primarily programmed in the C language~\cite{es.c}.
%
The predominance of C is associated with its portability across architectures,
efficiency in terms of memory and CPU, extensive legacy codebase, and number of
programmers.
%
Therefore, the full IoT development stack typically relies on
C~\cite{wsn.tos,wsn.contiki,wsn.riot}, from high-level applications and
network protocols, down to operating systems, network drivers, and SoC
firmwares.
%
However, C is a simple hardware abstraction (sometimes referred to as the
\emph{portable assembly}) and has no awareness of the surrounding environment
it is embedded on.
%
For instance, there is no dedicated vocabulary in the language to express
concepts that naturally appear in IoT applications, such as time, communication
with the real world, concurrency of events, and energy awareness.
%
C is also memory unsafe, leading to all sorts of bugs such as memory leaks,
buffer overflows, and wild pointers~\cite{c.unsafe}.

TinyOS~\cite{wsn.tos} uses a C dialect~\cite{wsn.nesc} and partially addresses
our goals.
It targets constrained architectures and maximizes standby time (goals 1 and 2).
TinyOS is also transparent at the level of the operating system, meaning that
applications built on top of it will take advantage of standby power for free
(goals 3 and 4).
They claim to operate in sleep modes 44\% of the time~\cite{wsn.tos.2}, which
is close to IEA's estimates of 65\% energy savings.
However, an operating system codebase is complex and still written in C.
Our goal is to provide transparent support at the language level, which even an
OS implementation could take advantage (goal 3).
Another issue with TinyOS is related to the adoption barrier imposed by its
programming model (goal 4): applications have to be written in a
callback-driven style in C, which is widely recognized as hard and error
prone~\cite{sync_async.cooperative,wsn.protothreads,rp.goto,rp.deprecating,frp.rx}
%
%TODO: 1.1 register scan

\subsection{Preliminary Results}
\label{sec.ceu}

I have been working in the design of \CEU, a new reactive programming language
targeting resource-constrained embedded systems, for the past 8
years~\cite{ceu.sensys11,ceu.tr,ceu.sensys13,ceu.rem13,ceu.phd,ceu.rebls14,
ceu.mod15,ceu.rebls15,ceu.terra,ceu.media.webmedia16,ceu.tecs17}.
%
\CEU is grounded on the synchronous concurrency model, which has been
successfully adopted in the context of hard real-time systems such as avionics
and automobiles industry since the 80's~\cite{rp.twelve}.
%
The synchronous model trades power for reliability and has a simpler model
of time that suits most requirements of IoT applications.
%
On the one hand, this model cannot directly express time-consuming
computations, such as compression and cryptography algorithms, which are
typically either absent or delegated to auxiliary chips in the context of the
IoT.
%
On the other hand, all reactions to the external world are guaranteed to be
computed in bounded
time, ensuring that applications always reach an idle state amenable to standby
mode.

In a previous work~\cite{ceu.sensys13}, we adapted \CEU to execute on top of
TinyOS in the context of Wireless Sensor Networks, and developed a number of
applications, protocols, and device drivers.
We evaluated the expressiveness of \CEU in comparison to C and attested a
reduction in source code size (around 25\%) with a small increase in memory
usage (around 5-10\% for text and data).
We also evaluated CPU responsiveness under heavy network load and attested a
similar performance between \CEU and C, which implies comparable energy
efficiency while in active mode.
%
In a related work, we developed Terra, a virtual machine for \CEU targeting
constrained networked devices that supports low-power remote
reprogramming~\cite{ceu.terra}.
Virtual machine applications are an order of magnitude smaller than full
binaries, reducing data traffic considerably and, consequently, energy
consumption.

In a recent paper~\cite{ceu.tecs17}, we discuss in extent the design decisions
behind \CEU under the perspective of
synchronous languages, providing a deep comparison with the seminal language
Esterel~\cite{esterel.ieee91}.
The most fundamental difference to Esterel is \CEU's event-triggered notion of
time, in which the occurrence of an event (e.g., a button press) defines an
indivisible instant in an application logical timeline.
This characteristic makes the model of time of \CEU exclusively reactive to
incoming events.
As a consequence, time does not elapse during idle periods, making all
applications susceptible to standby. % modes.

\subsection{A Motivating Example}

{\linespread{1}
\begin{figure}[t]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[xrightmargin=1cm]
digitalWrite(13, 0);
int old = 0;
while (1) {
    int cur = digitalRead(2);
    if (cur != old) {
        digitalWrite(13, cur);
        old = cur;
    }
}

\end{lstlisting}
\centering\small{[a] Active polling in C}
\end{minipage}
%
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=left]
input  int PIN_02;
output int PIN_13;

emit PIN_13(0);
loop do
    var int v = await PIN_02;
    emit PIN_13(v);
end
.
\end{lstlisting}
\centering\small{[b] Reactive execution in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Controlling a LED from button state changes.
\label{lst.basic}
}
\end{figure}
}

Figure~\ref{lst.basic} shows a basic program that makes a button connected
to the microcontroller's pin 2 to control a LED connected to pin 13.
Whenever the state of the button changes (i.e., pressed or unpressed), the
state of the LED also changes (i.e., on or off).

The implementation in C in Figure~\ref{lst.basic}.a first turns off the LED
(ln. 1), then sets an initial state for the button (ln. 2), and then reads in a
loop (ln. 3-9) its current state (ln. 4) as fast as possible.
If the current state differs from the old state (ln. 5-8), the new state is
copied to the LED (ln. 6), and is also set as the old state (ln. 7).
%
This approach which actively samples the state of an external device (e.g., a
button) is known as \emph{polling} or \emph{software-driven I/O}.
%
On the one hand, this style does not involve dealing with the complexity of
callbacks (as discussed in Section~\ref{sec.related}) and is widely adopted in
Arduino, the most popular embedded platform (to be discussed in
Section~\ref{sec.method.software}).
On the other hand, polling wastes CPU cycles and prevents the device to enter in
standby.

The implementation in \CEU in Figure~\ref{lst.basic}.b
uses dedicated vocabulary to handle events from I/O devices without resorting
to callbacks.
The pins are first declared to determine their purpose in the program, i.e.,
\code{input} for the pin connected to the button and  \code{output} for the pin
connected to the LED (ln 1-2).
Then, the LED is turned off with an \code{emit} (ln. 4), which indicates an
output operation.
Then, the program enters in a loop to react to changes in the button (ln. 5-8).
The \code{await} primitive accounts for the reactive nature of \CEU.
The program literally waits for a change in the button (ln. 6) to be able to
proceed and set the new state of the LED (ln 7).
%
This way, when an application reaches an idle state, the language has precise
information about which events can awake that application.

By design, all lines of execution in \CEU always reach an \code{await} at the
end of a reaction to an event (otherwise, the application does not
compile~\cite{ceu.sensys13}).
%
This not only allows the application to enter standby mode, but effectively use
the deepest sleeping level considering all possible awaking events.
While in standby, only a hardware interrupt associated with the events can
awake the application, making it sleep for longest possible periods of time.
%
%This is still a theoretical assumption that needs to be proven in practice.

\section{METHODOLOGY}

\subsection{First Year}
\label{sec.first}

\subsubsection{IoT Hardware Infrastructure}

The IoT infrastructure is built on top of embedded systems such as sensor
nodes, network routers, and other low-power microcontroller-based devices and
appliances.

We will use off-the-shelf Arduinos~\cite{arduino} as the main hardware IoT
platform.
Most Arduinos are based on low-cost 8-bit microcontrollers from Atmel, such as
the popular \emph{ATmega328p}~\cite{arduino.atmega328p},
%This chip includes the \emph{picoPower} technology,
which supports six sleeping
modes which can reduce power consumption to very low levels.
Depending on some configurations (e.g., frequency and voltage), an Arduino can
draw from 45mA at full operation, down to 5uA in power down mode (the deepest
sleeping mode).
We will pursue that IoT applications operate only 50\% of the time in the
average, which is not unlikely~\cite{wsn.tos.2}.
Considering the negligible consumption in power down mode, this will meet our
goals of reducing energy consumption by 50\%.
% 45mA@5V = 0.2W: below 1W, but no radio and other devices

Besides the good support for low-power standby, Arduinos are inexpensive,
open source, and widely used.
%
We will take advantage of the many environments in which Arduino is popular:
%
\begin{description}
\item[Research]
    There is a lot of research using Arduino in the context of the IoT (e.g.,
    infrastructure~\cite{arduino.infra}, healthcare~\cite{arduino.health},
    home automation~\cite{arduino.home}, and energy
    awareness~\cite{arduino.energy}).
    In addition to the availability of support literature, the popularity of
    Arduino will make our own research more accessible and reproducible to
    other groups.
\item[Education]
    Arduino is used in many University courses worldwide
    (e.g. \cite{arduino.edu.1,arduino.edu.2,arduino.edu.3,arduino.edu.4}).
    We have been using Arduino with \CEU in an undergraduate course for the
    past 3 years, which will
    %Arduino lowers the learning barrier for students, and will
    allow us to evaluate
    our results with less experienced embedded programmers.
\item[Hobbyists]
    The Arduino hobbyist community cultivates an abundance of resources such as
    forums, wikis, and publicly available software.
    It is important for us to have available software that we can adapt to \CEU
    to evaluate energy efficiency gains.
\item[Non-Programmers]
    Arduino is also becoming popular among non-programmers and a lot of
    software will be produced by them in the near future.
    We are in the process of creating an interdisciplinary course targeting
    non-specialists (e.g., arts, design, architecture).
    This audience will be a good target to evaluate our transparent approach for
    energy efficiency.
\end{description}

%TODO: radio

\paragraph{\textbf{Expected Contributions}} %\mbox{}\\

Since we are adopting a third-party hardware platform with existing support for
low power modes, we do not expect significant contributions.

\paragraph{\textbf{Risks (very low: 1/5)}}

Arduino is a cheap and popular embedded platform with widespread use in the
academia.
%
There is a considerable number of reports that claim to achieve very low power
consumption with Arduino in deep standby modes.
%
%TODO: radio

\begin{comment}
http://www.home-automation-community.com/arduino-low-power-how-to-run-atmega328p-for-a-year-on-coin-cell-battery/

- HIGH
    - 5V
        - 45 mA current
- LOW
    - POWER-DOWN SLEEP
        - 5V
            - 23 uA
            - 5.8 uA (no power regulator)
        - 3V3
            - 54 uA (why higher than 5V?)
            - 4.5 uA (no power regulator)

ATmega328P Pro Mini Version	PWR Source	State	5V@16MHz    3V3@8MHz
Unmodified 	                RAW Pin	    ACT	    19.9 mA	    4.74 mA
Unmodified 	                RAW Pin	    PDS	    3.14 mA	    0.90 mA
No Power LED 	            RAW Pin	    ACT	    16.9 mA	    3.90 mA
No Power LED 	            RAW Pin	    PDS	    0.0232 mA*	0.0541 mA*
No Power LED, no Regulator	VCC Pin	    ACT	    12.7 mA	    3.58 mA
No Power LED, no Regulator	VCC Pin	    PDS	    0.0058 mA	0.0045 mA

That is 4 years on a 9 V battery with 1,200 mAh capacity or 2,000 times more efficient than the Arduino Uno

===

https://github.com/rocketscream/Low-Power

===

https://www.avrprogrammers.com/howto/atmega328-power

- HIGH
    - 5V@16MHz
        - 16mA
    - 3V3@1MHz
        - 92uA
- LOW
        - 100uA

Frequency	mA@5V	mA@3.3V
1MHz	    6.75	0.92
8MHz	    11.68	4.16
16MHz (LP)	16.32	7.40
16MHz (FS)	17.52	7.94
20MHz (LP)	19.71	8.86
20MHz (FS)	21.12	9.46

Peripheral	mA@3V3@8MHz
ADC	        0.23
USART	    0.08
SPI	        0.15
Timer0	    0.09
Timer1	    0.12
Timer2	    0.12
TWI	        0.18

They can each be turned off in the Power Reduction Register (PRR) by writing a
one to the bit position.

In this example code, we shut down the CPU and wake it again when the INT0
interrupt fires (the INT0 pin goes low). The average current draw is 380µA@5V,
which is pretty low. At 3.3V the current drops to 72uA, but if we wake up and
write to an SD card, for instance, the current draw will jump to 13mA for
100mS, then 50mA for 50mS, before it drops back down to the 72µA level again.
\end{comment}

\subsubsection{IoT Software Infrastructure}
\label{sec.method.software}

Arduino provides a thin software layer on top of C and C++, and the typical way
in which applications are built is through polling (as discussed in
Section~\ref{sec.ceu}).
In Arduino, even basic functionality, such as timers
    (\code{delayMicroseconds}%
        %\footnote{Source code for \code{delayMicroseconds}: \url{https://github.com/arduino/Arduino/blob/master/hardware/arduino/avr/cores/arduino/wiring.c#L120}}%
    ),
A/D converters
    (\code{analogRead}%
        %\footnote{Source code for \code{analogRead}: \url{https://github.com/arduino/Arduino/blob/master/hardware/arduino/avr/cores/arduino/wiring_analog.c#L38}}%
    ),
and SPI
    (\code{SPI.transfer}%
        %\footnote{Source code for \code{SPI.transfer}: \url{https://github.com/arduino/Arduino/blob/master/hardware/arduino/avr/libraries/SPI/src/SPI.h#L208}}%
    ),
uses busy-wait loops that waste energy in active mode.
%\footnote{USART: \url{https://www.arduino.cc/en/Serial/Available}\\SPI:}%

In order to provide automatic standby, applications need to be entirely
reactive to events, and we will rebuild all of the software infrastructure from
scratch in \CEU.
%
This process will consist primarily of rewriting device drivers, which are the
pieces of software that interface directly with the hardware.
%
More concretely, at the lowest level, energy-efficient software depends on
\emph{interrupt service routines (ISRs)}, which awake the CPU from standby when
an external peripheral event occurs.

Recently, we have added support for ISRs as a primitive concept of
\CEU, which will allow us to rebuild the IoT software infrastructure with
standby awareness from the ground up.
%
Figure~\ref{lst.isrs} is an sketch of our intended approach to achieve transparent
standby modes (the lines with \code{<...>} are pseudo code).
%
The application in Figure~\ref{lst.isrs}.a requests, every hour (ln. 7-11), an
A/D conversion (ln. 8) and awaits its completion (ln. 9) to perform some action
(collapsed in ln. 10).
%
Note that the application uses abstract input/output events to interface with
the A/D converter (ln. 3-4) and does not manage energy explicitly.
%
Figure~\ref{lst.isrs}.b would be the associated driver with the A/D converter,
which implements the input/output events declared as abstractions in the
application code.
The output request (ln. 3-7) configures the A/D peripheral (ln. 4) and enables
interrupts in the microcontroller (ln. 5).
It also needs to set the deepest mode supported by the A/D peripheral (ln. 6)
so that the language uses it as soon as the program reaches the idle state.
The input definition (ln. 9-13) implements the ISR for the A/D converter.
When the conversion is done, the peripheral awakes the CPU, and this code
executes immediately in interrupt context.
The code disables A/D interrupts (ln. 10), reads the A/D converter data
register with the converted value (ln. 11), and returns this data to the
application (ln. 12).
A \code{return} in interrupt context will enqueue the input event
\code{ADC\_DONE}, which will awake the application code associated with it
(ln. 9 in Figure~\ref{lst.isrs}.a).

With this approach, the application code remains similar to the one would write
for Arduino (apart from minor syntax mismatches).
However, instead of wasting cycles in busy-wait loops, the applications will
enter the deepest possible standby mode when idle.

{\linespread{1}
\begin{figure}[t]
\begin{minipage}[t]{0.56\linewidth}
\begin{lstlisting}[xrightmargin=1cm]
// application.ceu

output none ADC_REQUEST;
input  int  ADC_DONE;
#include "adc.ceu" // driver implementation

every 1h do
    emit ADC_REQUEST;
    var int v = await ADC_DONE;
    <performs-some-action>;
end

.
\end{lstlisting}
\centering\small{[a] Application: no explicit energy management}
\end{minipage}
%
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[numbers=left]
// adc.ceu

output none ADC_REQUEST do
    <configures-ADC>;
    <enables-ADC-interrupts>;
    <sets-the-deepest-standby-mode>;
end

input int ADC_DONE [ADC_vect_num] do
    <disables-ADC-interrupts>;
    var int v = <reads-adc-data>;
    return v;
end
\end{lstlisting}
\centering\small{[b] Driver: explicit energy management }
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ ISRs in \CEU
\label{lst.isrs}
}
\end{figure}
}

\paragraph{\textbf{Expected Contributions}}

\begin{enumerate}
\item \textbf{An environment-aware system language:}
    \CEU's dedicated vocabulary to deal with events raises the
    programming abstraction to a level closer to the domain of IoT, providing
    more safety and expressiveness to programmers.
    These results have already been partially published in the
    past~\cite{ceu.sensys13}, when we implemented a radio device driver on top
    of TinyOS.
    Nevertheless, the novel support for ISRs would complete the whole
    development cycle, from the application down to the hardware, without
    operating system support.
\item \textbf{An energy-aware system language:}
    The method described in this section makes all applications subject to
    standby modes transparently.
    Only device drivers, which are typically provided by vendors, will require
    explicit energy management.
    In our case, we will need to rewrite each driver only once, as part of the
    software infrastructure, and all applications built on top of them would
    benefit from energy efficient automatically.
%\item[TODO: education]
\end{enumerate}

\paragraph{\textbf{Risks (high: 4/5)}}

%\begin{enumerate}
%\item \textbf{Primitive language support for ISRs:}
As far as we know, abstract ISR support at the language level has not been
tried in the past.
The challenge is to choose the right level of abstraction that is, at the same
time, powerful to support the range of devices and portable to work in multiple
platforms.
Since \CEU integrates seamlessly with C~\cite{ceu.tecs17}, it is always
possible to circumvent abstractions, at least in early stages of development,
until the appropriate level is found.
%\item \textbf{Programming efforts:}

From the engineering perspective, rewriting drivers is a laborious endeavour.
As a rough reference, about half of the 15 million lines of the Linux codebase
consists of device drivers.
With the tight integration between \CEU and C, however, significant parts of
existing drivers can be reused.
%\end{enumerate}

We already have some evidence of the feasibility of programming device drivers
in \CEU based on a project of this year's \emph{Google SoC}%
\footnote{GSoC project: \url{http://www.lua.inf.puc-rio.br/gsoc/ideas2017.html#ceu}}.
In the period of 3 months, an undergraduate student studied and implemented
drivers for an A/D and SPI peripherals using the recent support for
ISRs in \CEU.

\subsubsection{IoT Applications}
\label{sec.method.apps}

In order to evaluate the gains in energy efficiency with the proposed
infrastructure, we will need to evaluate the consumption of realistic applications.
%
The Arduino community has an abundance of open-source projects which can be
rewritten in \CEU to take advantage of transparent standby modes (as illustrated
in Figure~\ref{lst.isrs}).
%
%Also, most peripherals for Arduino ship with drivers and applications that
%explore their functionalities (e.g., accelerometers, touch screens, GPS, etc.).
%
Then, we will be able to compare the original and rewritten versions in terms of energy
consumption to draw conclusions about the effectiveness of transparent standby
modes.

We will start with simple applications, such as the \emph{Blink} that ships
with the Arduino distribution and blinks an LED every second.
The code is reproduced in Figure~\ref{lst.blink}.a and a very similar version
in \CEU appears in Figure~\ref{lst.blink}.b.
%
The example is in active mode for a negligible time just to set the I/O ports
(ln. 3 and 5), remaining idle most of the time.
%
We expect this application to consume much less energy in \CEU, since the
\code{await} statements (ln. 4 and 6) will put the device into standby
automatically, while the version in C uses a busy-wait loop inside the
\code{delay} statements (ln. 4 and 6).

{\linespread{1}
\begin{figure}[t]
\begin{minipage}[t]{0.49\linewidth}
\begin{lstlisting}[xrightmargin=1cm]
pinMode(13, OUTPUT);
while (1) {
    digitalWrite(13, 1);
    delay(1000);
    digitalWrite(13, 0);
    delay(1000);
}
\end{lstlisting}
\centering\small{[a] Blinking a LED in C}
\end{minipage}
%
\begin{minipage}[t]{0.49\linewidth}
\begin{lstlisting}[numbers=left]
output int PIN_13;
loop do
    emit PIN_13(1);
    await 1s;
    emit PIN_13(0);
    await 1s;
end
\end{lstlisting}
\centering\small{[b] Blinking a LED in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Blinking a LED connected to pin 13 every second.
\label{lst.blink}
}
\end{figure}
}

The most realistic scenarios for the IoT use radio communication extensively.
%
In this context, we will evaluate from simple handmade ad-hoc protocols to more
complex energy-aware protocols (as discussed in Section~\ref{sec.related}) and
see to what extent our proposal would effectively contribute to energy savings.
%
The \emph{RadioHead} project%
\footnote{RadioHead project: \url{http://www.airspayce.com/mikem/arduino/RadioHead/}}
supports dozens of RF (radio frequency) chipsets and provides examples that
explore mesh topologies as well as low-power energy modes.
%
We plan to support at least two chipsets: \emph{nRF24L01}~\cite{radio.nrf24l01}
and \emph{RFM69HCW}~\cite{radio.rfm69hcw}, both of which have good support for
low power consumption.
%
%We expect that applications that use energy-aware protocols extensively (but
%not much other peripherals) will not benefit from our proposal, since they
%implement standby policies explicitly (see also
%Section~\ref{sec.method.systematic}).
%
%Nevertheless, we can still evaluate the implementation of these protocols and
%assert that the new implementations in \CEU do not have explicit standby
%policies but are still as energy efficient than the originals.

In the long term, we expect to make the case for
developers to rewrite their applications in \CEU to take advantage of standby
modes for free.
%
Towards this direction, we will evaluate IoT applications rewritten in \CEU based on
the following criteria:
%
\begin{description}
\item[Time to rewrite]
This criteria relates to the incentives to rewrite existing applications to
\CEU.
It is a tradeoff between the expected rewriting times and gains in energy
efficiency as compared to standard Arduino code.
\item[Coding ``aesthetics'']
To lower the adoption barrier, it is also important that the proposed
programming style is sufficiently familiar, and, at least, as easy to write applications.
\item[Energy consumption]
More importantly, rewritten applications should have significant gains in
energy efficiency to justify a complete application rewrite.
\end{description}

\paragraph{\textbf{Expected Contributions}}
\begin{enumerate}
%\item \textbf{Higher-level programming abstractions:}
    %Similarly to IoT software infrastructure (discussed in
    %Section~\ref{sec.method.software}), we expect that applications will take
    %even more advantage of the abstraction level provided by \CEU.
\item \textbf{A realistic programming alternative for the IoT in the Arduino:}
    \CEU is a 8-year project and has a mature open-source implementation that
    is publicly available%
\footnote{\CEU repository: \url{https://github.com/fsantanna/ceu}}.
    With the adaptation to the context of energy-efficient IoT, \CEU may cross
    the academic fences in the short term and become a practical alternative
    for the Arduino community.
\item \textbf{Energy-efficient IoT applications:}
    We expect substantial gains in energy efficiency for all applications that
    do not manage energy explicitly.
    Most Arduino APIs use busy-wait polling to wait for input, which are very
    energy inefficient, and could take advantage of transparent standby.
    %We also expect gains in comparison to some ad-hoc energy-aware
    %implementations---we believe that, for the most cases, a transparent
    %solution is more effective than explicit programming efforts.
\end{enumerate}

\paragraph{\textbf{Risks (medium: 3/5)}}

%\begin{enumerate}
%\item \textbf{Effective energy efficiency:}
We do not know precisely the practical gains in energy efficiency in the
standby modes of Arduino.
Most reports that compare theoretical (from official datasheets)
and measured consumption claim good performance but are non peer-reviewed
research.
For instance, a specialized company%
\footnote{RocketScream: \url{http://www.rocketscream.com}}
provides Arduino-based IoT devices and a cross-platform library dedicated to
standby modes that is widely used by the community.
%\end{enumerate}

\subsection{The Three Years to Follow}

\subsubsection{Systematic Energy Awareness}
\label{sec.method.systematic}

As discussed in Section~\ref{sec.introduction}, systematic energy awareness
considers the IoT network as a whole, with cooperating devices trying to
maximize energy efficiency globally.
%
In this context, \emph{adaptive computing}~\cite{adaptive} is the capability of
the IoT to adapt energy consumption dynamically, based on application demands
and levels of power supply during execution.
%
Ultimately, the goal is still that each device maximizes its idle
periods to be more susceptible to standby modes.
Hence, the research of the
first year is a prerequisite for this phase.

We will continue to investigate how language mechanisms can aid energy
efficiency, but now in the context of adaptive computing.
%
Figure~\ref{lst.pause} is a sketch of an application that adapts its behavior
under low-power mode by suspending parts of its functionality (the lines with
\code{<...>} are pseudo code).
%
The application relies on the \code{par} construct of \CEU (ln. 3-16) which
creates three concurrent lines of execution (ln. 4, 6-8, and 10-15).
The first line of execution (collapsed in ln. 4) implements the mandatory
functionality of the application, which must execute regardless of the power
mode.
The second line of execution (ln. 6-8) implements extra functionality
(collapsed in ln. 7) that can
be suspended in low-power mode without affecting the main functionality.
The \code{pause/if} construct of \CEU (ln. 6) suspends the nested costly
functionality whenever the event
\code{enter\_low\_power\_mode} is set to \emph{true} and resumes when set to
\emph{false}.
The third line of execution is responsible for implementing the adaptive
manager by interacting with the external world.
In this case, every time the button is pressed (ln. 11 and 13), the device switches
the power mode by triggering the event \code{enter\_low\_power\_mode}
(ln. 12 and 14).

{\linespread{1}
\begin{figure}[t]
\begin{lstlisting}[xleftmargin=1cm,numbers=left]
input none BUTTON_PRESSED;             // input for every button press
event bool enter_low_power_mode;       // event to toggle power mode
par do
    <mandatory-functionality>          // omitted code with basic functionality
with
    pause/if enter_low_power_mode do   // the nested block is suspended in low mode
        <energy-costly-functionality>  // omitted code with costly functionality
    end
with
    loop do                            // toggles power mode on every button press
        await BUTTON_PRESSED;
        emit enter_low_power_mode(true);
        await BUTTON_PRESSED;
        emit enter_low_power_mode(false);
    end
end
\end{lstlisting}
%\rule{8.4cm}{0.37pt}
\caption{ Adaptive application that changes behavior depending on the power mode.
\label{lst.pause}
}
\end{figure}
}

The example suggests that \code{BUTTON\_PRESSED} represents a simple button
connected to the device.
%
However, the advantage of using the abstract concept of an input event is that
the actual implementation may vary.
%
For instance, \code{BUTTON\_PRESSED} might be a ``virtual button'' associated to a
remote device (or set of devices).
%
We will take advantage of the abstract concepts of input and output events to be
able to specify IoT interactions at a higher level.

Note that adaptive computations must be programmed explicitly, since they
are application dependent and have particular requirements.
%
Hence, since it is not possible to offer transparent language mechanisms (such
as in our proposal for standby modes), we will investigate \emph{non-intrusive
language mechanisms}.
For instance, note in the example that the \code{<energy-costly-functionality>}
(ln. 7) did not suffer any internal modifications to be suspended, but was
simply enclosed by the \code{pause/if} construct.
%
This way, making an application adaptive to embrace energy efficiency will
require less programming efforts.
Furthermore, the implementation of the application functionalities will not be
intermixed with energy awareness, making the code more modular.

The evaluation process for systematic energy awareness will be similar to the
one adopted for device-based awareness (as discussed in
Section~\ref{sec.method.apps}):
take existing energy-aware IoT protocols and applications for Arduino, and
rewrite them using non-intrusive mechanisms.
% TODO: which apps?
Then, employ the same evaluation criteria
(i.e., time to rewrite, coding aesthetics, and energy consumption).
%
\begin{comment}
\item[Time to rewrite:]
This criteria relates to the incentives to rewrite existing applications to
\CEU.
It is a tradeoff between the expected gains in energy efficiency and rewriting
time.
\item[Coding ``aesthetics'':]
To lower the adoption barrier, it is also important that the proposed
programming style is sufficiently familiar, and, at least, as easy to write.
\item[Energy consumption:]
More importantly, rewritten applications should have significant gains in
energy efficiency to justify a complete rewrite.
\end{comment}

\paragraph{\textbf{Expected Contributions}}
\begin{enumerate}
\item \textbf{Higher-level programming abstractions:}
    Similarly to the IoT software infrastructure and applications (discussed in
    Sections~\ref{sec.method.software}~and~\ref{sec.method.apps}), we expect
    that systematic energy awareness will take advantage of the non-intrusive mechanisms
    that \CEU provides and will be implemented at a higher level of abstraction.
\item \textbf{Adaptive and energy-efficient IoT applications:}
    We expect to achieve the same energy efficiency as that of hand-tweaked
    applications but with less programming efforts.
\end{enumerate}

\paragraph{\textbf{Risks (medium: 3/5)}}

%\begin{enumerate}
%\item \textbf{Effective energy efficiency:}
On the one hand, hand-tweaked applications always have a theoretical higher
efficiency than applications based on richer abstractions.
On the other hand, having to deal with the complexity of hand-tweaked
applications makes them less amenable to adaptive computations.
In some situations, we may face this tradeoff and sacrifice adaptiveness to
achieve satisfactory energy efficiency.
%\item \textbf{Programming efforts:}

From the engineering perspective, rewriting complex applications takes a lot of
time.
However, considering the period of 4 years, time estimates can
be reconsidered periodically to employ more human resources into the project.
%\end{enumerate}

%\vspace{-1cm}
\subsubsection{Complex IoT Architectures and Applications}
\label{sec.method.complex}

The niche of constrained embedded systems, which includes Arduino, covers the
substantial (and increasing) portion of IoT applications.
They typically do not require significant computing resources but are sensitive
to physical dimensions and energy consumption.
%
However, the IoT also consists of more traditional networked devices, such as
routers, servers, and smartphones.
%
As of 2016, there were 3.9 billion smartphone subscriptions worldwide, and
this number is expected to reach 6.8 billions by 2022~\cite{ericsson.mobility}.

Smartphones rely on much more complex architectures than microcontroller-based
embedded systems (e.g., 32-bit CPUs with memory management unit).
They typically rely on an operating system, a complete TCP/IP stack, and can
execute multiple applications at the same time.
In addition to reactive applications, typical of the IoT, smartphones also
perform active computations, such as audio/image/video processing and
cryptographic functions.
Nevertheless, they are an important piece of the IoT, serving as a common human
interface to process, visualize, and actuate on the network.

Smartphones have similar restrictions in battery consumption and could also
take advantage of the techniques we propose for constrained embedded systems.
%
%Therefore, we will address energy efficiency through transparent language
%mechanisms to achieve extensive use of standby.
%
In order to transpose the barrier from constrained devices to smartphones for
the IoT, we will take a similar path as presented in Section~\ref{sec.first}:
%
\begin{description}
\item[Hardware Infrastructure]
We will use the BeagleBone Black~\cite{bbb.manual}, which shares similar goals with
Arduino, providing a cheap and open-source platform but which is suitable for rich
applications, such as graphical user interfaces, multimedia, and games.
It is based on the Texas Instruments (TI) Sitara ARM
architecture~\cite{bbb.sitara}, featuring a 1GHz CPU, with 512MB RAM and 2GB
flash.
Like happens with Arduino, the BeagleBoard community offers an abundance of
open projects and publicly available software.
\item[Software Infrastructure]
In order to ensure automatic standby for applications, all software
infrastructure, mostly device drivers, needs to be rebuilt from scratch using
ISRs in \CEU.
However, smartphones have much more complex peripherals, such as WiFi adapters
and graphics processors.
To reduce the barrier adoption,
TI provides \emph{StarterWare}%
\footnote{TI Sitara: \url{http://www.ti.com/tool/starterware-sitara}},
a free software development package with open
support for the basic Sitara peripheral drivers.
\item[Applications]
In addition to IoT applications, typical smartphone applications, such as
instant messaging and internet browsing, can also maximize energy efficiency
through standby.
We will rewrite from simple applications, such as a graphical clock, to more
complex networked applications such as a web browser:
\begin{itemize}
\item \textbf{xclock} displays the time in analog or digital and the source code is around 5k lines in C.
% TODO: 15-20 loc
\item \textbf{Dillo}%
\footnote{Dillo web browser: \url{https://www.dillo.org/}}
is a minimalistic web browser with the source code around 50k lines in C.
\end{itemize}
\end{description}

\paragraph{\textbf{Expected Contributions}}

\begin{enumerate}
\item \textbf{Energy-efficient smartphones:}
    The transition from simple mobile phones to smartphones resulted in a
    degrading effect on battery lives.
    Phones used to have autonomy of weeks and now require daily recharges.
    Most of the time, however, smartphones are sitting in our pockets wasting
    energy.
    We expect to increase the battery autonomy considerably while keeping the
    functionality of a modern smartphone.
\item \textbf{Towards an environment- and energy-aware operating system:}
    The whole software technology of Smartphones is built on top of bloated and
    antiquated personal computer operating systems like Windows and UNIX (e.g.,
    macOS and Android).
    These operating systems were not designed with mobile systems and energy
    awareness in mind.
    Our proposal is a step towards a new environment- and energy-aware
    operating system.
\end{enumerate}

\paragraph{\textbf{Risks (very high: 5/5)}}
%\begin{enumerate}
%\item \textbf{Programming efforts:}
Modern smartphone hardware architectures are much more complex than of
constrained embedded systems.
For this reason, making an application to be energy-aware from the ground up,
starting from a hardware interrupt, is an enduring challenge.
Nonetheless, the proposed language infrastructure is the same, and all other
required pieces such as drivers are publicly available.
%\end{enumerate}

\begin{comment}
- Rust is a memory-safe alternative to C~\cite{rust.os}
    - problems with embedded systems
        \cite{rust.problems}

- Ceu
    - interoperable with C

- HW platform Mini + NRF
    - radio + microcontroller + development board
\end{comment}

\section{TIMELINE}

Figure~\ref{fig.timeline} shows the project timeline for the full period of 4
years.
The two lines in black refer to the \emph{constrained} and \emph{complex} IoT
architectures phases (Sections~\ref{sec.first}/\ref{sec.method.systematic} and
Section~\ref{sec.method.complex}, respectively).
They only intersect during the second semester of the second year, when the
constrained phase is about to terminate and the complex phase kicks off.
At the end of the first year, we expect to address energy efficiency for the
constrained IoT architectures, with the hardware and software
infrastructure set, and the basic and networked applications completed
(Section~\ref{sec.first}).
The second year covers systematic energy awareness for the constrained IoT
(Section~\ref{sec.method.systematic}) and marks the beginning of the complex IoT
phase (Section~\ref{sec.method.complex}), which extends to the end of the
period of 4 years.

The small tasks in red are related to hardware acquisition and setup of
existing IoT architectures.
The tasks in yellow refer to the software infrastructure, which will be in
continuous iterations during the whole project cycle.
They require the most experienced researchers and consist of designing the
language mechanisms, employing automatic standby, and implementing the most
essential drivers.
The tasks in green refer to the applications built on top of the software
architecture.
They have delimited development cycles and require less experienced
researchers.

\begin{sidewaysfigure}
    \vspace{17cm}
    \includegraphics[width=\columnwidth]{serra-big}
    \caption{Project timeline
        \label{fig.timeline}
    }
\end{sidewaysfigure}

\section{RESOURCES}

For the period of 3 years, we will require a budget of R\$345k:
\begin{itemize}
    \item R\$300k for staff payment (87\%),
    \item R\$20k for equipment (6\%), and
    \item R\$25k for publication and travel costs (7\%).
\end{itemize}

\subsection{The Team}

Most of the resources will be employed in research staff consisting
of undergraduate, MSc, and PhD students.
%
The quality and long-term commitment of the students is paramount for the
success of the project.
%
For this reason, we want to offer a salary compatible with the industry, which
is considerably larger than typical research scholarships.

\begin{description}
\item[Proponent]
The proponent will work in all project fronts: managing the resources and
staff, writing papers and supervising students, and working on the technical
challenges of the project.
\item[Advisors]
The proponent is an associate professor at UERJ and a member of the LabLua%
\footnote{LabLua: \url{http://www.lua.inf.puc-rio.br/}.}
at PUC-Rio.
The advisors are experienced researchers from UERJ and PUC-Rio in the field of
programming languages, distributed systems, and wireless sensor networks.
They will support the academic duties, such as prospecting and supervising
students, writing papers, and helping on the general project guidelines.
\item[Research staff]
At the beginning of the first year, we will appoint two students to join the
project full time:
    one early-stage PhD student
and
    one MSc student or late-stage undergraduate student.
Every six months, we plan to add a new member to the staff until we reach a
limit of 6 students (consisting of 2 or 3 PhD students).
The PhD students will work with the software infrastructure and the other
students with the applications.
We also expect that the PhD students will work on academic papers and
take advantage of the project in their thesis.
Considering an average of 4 students during the whole project with and an
average salary of R\$2k, we expect to spend around R\$300k with payments.
\end{description}

\subsection{Infrastructure and Equipment}

The required equipment is divided between support for the research staff (printer and
laptops) and the project embedded systems (Arduino, Beagle Black, etc.):
\newline

%{%\linespread{0.5}
{\renewcommand{\arraystretch}{0.8}
\begin{tabular}{ | l | r | r | r | }
\hline			
EQUIPMENT       & PRICE & QTD   &   TOTAL   \\
\hline  
Laser printer   & 3000  &   1   &   3.000    \\
Laptop          & 4000  &   3   &  12.000    \\
Arduino kits    &  200  &   5   &   1.000    \\
Sensors \& radios & -   &   -   &   2.000    \\
Beagle Black    &  600  &   2   &   1.200    \\
Beagle Display  &  400  &   2   &     800    \\
\hline  
                &       &       &  R\$20k \\
\hline  
\end{tabular}
\newline
}

We expect that half of these resources will be acquired at the beginning of the
project.
The other half will be acquired gradually during the course of the 3 years.

\subsection{Publication and Travel Costs}

Our goal is to publish our results on international journals in the field of
embedded systems and sensor networks, such as \emph{ACM TECS} and \emph{ACM TOSN}.
%
However, the process to publish a paper in a journal may take longer than a year after
the submission.
For this reason, we also intend to publish partial results in conferences.

During the 3 years, we plan to attend two Brazilian conferences (e.g.,
\emph{SBRC}),
%and the \emph{Brazilian Symposium on Computing Systems Engineering (SBESC)}),
and one prestigious international conference (e.g., \emph{ACM SenSys}),
These venues are important to cultivate future cooperation with other research
groups.
For the Brazilian conferences, we will support 2 participants, and for the
international conference, only 1 participant.
%
Using the CNPq travel grants as a reference, attending a national conference
costs around R\$3k (including travel and accommodation for 3 days), while a 
international conference costs around US\$4k (including travel and
accommodation for 5 days).
Considering the two Brazilian conferences with two participants each, and one
international conference with one participant, the total travel costs for the
3-year period will be around R\$25k.

\begin{comment}
\section{CONCLUSION}

TODO

- conslidated research
- some steps already done
- we need human resources

realistic research
    - academic results
    - but also an open-source publicly available language that people can download and use
        - already available
        - all publics: research, hobbyists, students

TODO: long-term research
    - 8 years in the same project
\end{comment}

\newpage
\bibliographystyle{abbrv}
\bibliography{serra,my}
\end{document}

### 2018-01-05 Submissao/small.tex
(+) Documentos da submissao do projeto.

\documentclass[12pt,english]{amsart}
%\usepackage[margin=0.5in]{geometry}
\usepackage[a4paper, margin=2cm]{geometry}
\linespread{1.5}

%\usepackage{fontspec}
%\setmainfont{Times New Roman}
\usepackage{mathptmx}

\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{lscape}
\usepackage{rotating}
\usepackage{epstopdf}

\usepackage{verbatim}
\usepackage{url}
\usepackage{xspace}
%\usepackage{wrapfig}

\usepackage{listings}
\usepackage{color}
    \definecolor{light}{gray}{0.97}
    \definecolor{dark}{gray}{0.30}
\lstset{
%columns=fullflexible,
%basicstyle=\ttfamily,
escapeinside={||},
    %mathescape=true,
    language=C, % choose the language of the code
    basicstyle=\fontfamily{pcr}\selectfont\scriptsize\color{black},
    keywordstyle=\color{black}\bfseries, % style for keywords
    numbers=none, % where to put the line-numbers
    numberstyle=\tiny, % the size of the fonts that are used for the line-numbers
    backgroundcolor=\color{light},
    showspaces=false, % show spaces adding particular underscores
    showstringspaces=false, % underline spaces within strings
    showtabs=false, % show tabs within strings adding particular underscores
    %frame=single, % adds a frame around the code
    tabsize=2, % sets default tabsize to 2 spaces
    %rulesepcolor=\color{gray}
    captionpos=b, % sets the caption-position to bottom
    breaklines=false, % sets automatic line breaking
    %breakatwhitespace=false,
    numbersep=2em,
    % C was used in the blocksworld example to refer to block C and nowhere else
    emph={par,or,hor,do,end,loop,code,await,pause,emit,input,event,call,with,%
          var,and,then,else,return,pure,deterministic,nohold,finalize,%
          class, every, FOREVER, this, spawn, in, pool, watching, until, 
          interface, each, abort, when, signal, PROC, CHAN, SIGNAL, PAR, not,
          bool, tag, escape, traverse,implementation,output,true,false,
          native,@const,@pure,@safe,define,public,private,none},
    emphstyle={\bfseries},
    commentstyle=\color{dark}\scriptsize,
    %xleftmargin=20pt,
    %xrightmargin=20pt,
    framesep=20pt,
    %upquote=true,
    %aboveskip={1.5\baselineskip},
}

%\newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
\newcommand{\code}[1] {{\small{\texttt{#1}}}}

\usepackage{enumitem}
\setlist{nolistsep}

% Auto-Standby
\title{Energy Efficiency for IoT Software in the Large
        }%\\ \large{Research project proposal submitted to the Instituto Serrapilheira}}

%\author{Anonymous Author}

\begin{document}

\date{}
\maketitle

\vspace{-1cm}
\begin{comment}
\abstract{
TODO
}
\end{comment}

%\newpage
\section{Professional Narrative}

I work in the field of \emph{Programming Languages} with a focus on
\emph{Real-Time Reactive Systems}.
Reactive systems interact continuously with the external world through sensors
and actuators (e.g., buttons, LEDs, and motors).
When these interactions have to comply with strict deadlines, the system is
said to be \emph{real time}.
Most everyday applications, such as office suites, phone apps, and video games
are reactive but fairly tolerant to delays.
In contrast, embedded real-time systems, such as medical appliances,
avionic systems, and IoT devices, must conform to stricter deadlines.

%Since the beginning of my Masters at PUC-Rio, I have been working with
%programming languages support for reactive systems.
In the beginning of my Masters, I joined a group supporting
\emph{Ginga}, the software standard of the Brazilian Digital TV
System~\cite{ncl.abnt}, which also became an UN's ITU
standard~\cite{ncl.itu}.
%In a digital system, broadcasters can transmit and execute software in real
%time along with the primary audio-visual content.
% in order to enrich the
%audience experience (e.g., focused content, online shopping, etc.).
During this period, I worked on a non-intrusive integration of the two
standardized languages NCL and Lua through a reactive programming
interface.
%I have also participated in standardization meetings with broadcasters, TV
%manufacturers, government, and academia.
The outcome of this work was published in the \emph{ACM DocEng} in 2009.
I also wrote the chapter on developing with Lua and NCL in Ginga's main
reference book. % on developing Digital TV applications.

\begin{comment}
NCL is a declarative language to specify how multiple media objects are
composed together in cause-effect relationships to form a complete multimedia
presentation (e.g., after 1 minute of a video playback, show an image for 10
seconds).
Lua is a scripting language that extends the power of NCL with generic
programming language constructs such as loops, variables, and arithmetic
expressions.
During the same period, I wrote my Masters thesis on the design of
\emph{LuaGravity}, a programming language that extends Lua with support to
handle events from the environment.
Programs written in LuaGravity behave like spreadsheets, in which a user
event (e.g., typing a new number in a cell) propagates throughout the program
and changes all of its dependencies automatically.
\end{comment}

In 2009, I started the PhD program with the goal of designing a new
reactive language from scratch, now targeting resource-constrained embedded
systems.
On the one hand, these platforms are typically six orders of magnitude less
powerful than off-the-shelf personal computers in terms of memory size and CPU
speed.
On the other hand, embedded applications need higher reliability to operate
without human supervision for long periods of time.
Furthermore, a significant subset of embedded systems is deployed in locations
without power lines and need to run on batteries.
Hence, a programming language targeting constrained embedded systems needs to
be small, reliable, and energy efficient.
At the end of 2012, I earned a scholarship from SAAB Aerospace for a six-month
research period. % in the field of \emph{Secure Wireless Sensor Networks}.
I joined an University in Sweden and
we rewrote industrial-grade network drivers and protocols using my
research language and attested a considerable reduction in source code size
with similar resource usage in comparison to C. % of memory and CPU.
In 2013, we published a paper with these results in the \emph{ACM SenSys}.

After graduating, I contributed to the
project \emph{Cidades Inteligentes} (Smart Cities), which involves 20 
Universities and aims to build an infrastructure for the IoT.
%
In 2015, in cooperation with a PhD student, we published a paper the
\emph{ACM TOSN}, showing that we can reprogram embedded devices remotely using
less energy than the state of the art.
%
During the same period, I published a paper in the \emph{Int'l Conference on Modularity}
proposing a new concurrency abstraction that leads to more modular
programs.
%
I was also invited to present my research language in the annual
meeting of the \emph{Working Group on Language Design}, which is affiliated
with UNESCO's IFIP.%\footnote{IFIP: \url{http://www.ifip.org/}}.

Currently, as an associate professor at a Brazilian University,
%In 2016, I became an associate professor (\emph{professor adjunto}) at UERJ.
I continue to investigate how the design of languages can affect
the development of reactive applications.
For the last 4 years, I have have been either publishing in or acting in the
committee of the \emph{Int'l Workshop on Reactive Languages}.
In 2017, I published a consolidating paper on the design of
my research language in the \emph{ACM TECS}.

% multimedia, games, sensor networks, embedded systems

%- mobile phone game industry
%- empresa americana
%- Future of Programming
%- OpenSource
%- GSoC

%\newpage
\section{Science Outreach: The Role of IoT Software in Energy Efficiency}

According to the International Energy Agency (IEA)~\cite{iea.data}, there were
around 14 billion traditional network-connected devices in 2013 (e.g., mobile
phones and smart TVs).
This number is expected to increase to 50 billion by 2020 with
the proliferation of IoT devices (e.g., smart bulbs and fitness wearables).
%
IoT and traditional network-connected devices already outnumber people on the
planet by a factor of two, and the amount of data traffic is expected to grow
at an exponential rate in the next years.
%By 2050, the energy demands of networked devices should exceed the current
%consumption of Germany and Canada combined, corresponding to 6\% of current
%global electricity consumption.
%
However, most of the energy to power these devices is consumed while they are
in \emph{standby mode}.
%, i.e., when their software is neither transmitting or processing data.
%
The annual $CO_2$ emissions related to standby in Australia are equivalent to
those of 1 million cars.
%
%This undesirable inefficiency opens opportunities for improvements in energy
%savings closer to 65\% only considering currently available technologies.
%
The projected growth of IoT devices together with the surprising effects of
standby consumption, made network standby efficiency one of the six
pillars of G20's \emph{Energy Efficiency Action Plan}%
\footnote{G20's Energy Efficiency Action Plan: \url{https://www.iea-4e.org/projects/g20}}.

Other organizations have also reported the importance of energy savings for networked devices.
%
For the Internet Engineering Task Force (IETF), ``energy management is becoming an
additional requirement for networks due to several factors including the rising
energy costs, the increased awareness of the ecological
impact of operating networks and devices, and the regulation of
energy''~\cite{ietf.eman}.
%
For the American Council for an Energy-Efficient Economy (ACEEE),
``the potential for new energy efficiency remains enormous, (...) we must take
a systems-based approach to scale up energy efficiency.
%Intelligent efficiency is a systems-based approach to efficiency that can help
%to meet this need.
(...) intelligent efficiency is adaptive, anticipatory, and networked''~\cite{aceee.1}.
%
%The pioneer \emph{ENERGY STAR} program~\cite{energystar} states that
%``networked devices and networking equipment, as integrated systems, have the
%potential to contribute significant net energy savings''.
% \footnote{ENERGY STAR program \url{https://www.energystar.gov/index.cfm?c=prod_development.prod_development_epa_workshop}}
%
%The US Environmental Protection Agency (EPA) and Information Technology
%Industry Council (ITI) held a workshop to explore roles for the \emph{ENERGY
%STAR} and others program in promoting savings through intelligent energy
%efficiency~\cite{aceee.2}.
%, which differs from individualized
%component-based energy efficiency in that it is adaptive, anticipatory, and
%networked.

With regard to concrete initiatives, IEA's \emph{Electronic Devices and Networks}
focuses specifically on the issue of networked device standby%
\footnote{EDNA initiative: \url{https://edna.iea-4e.org/}}.
%It is one of the six fronts of G20's \emph{Energy Efficiency Action Plan}%
%\footnote{\url{https://www.iea-4e.org/projects/g20}}.
%Network standby refers to the minimum power a device requires to maintain its
%connection when not communicating or computing actively.
ACEEE's \emph{Intelligent Efficiency} promotes a systematic approach
to optimize the behavior of cooperating devices in order to achieve energy
savings as a whole.
Both approaches, device and system based, involve mostly software solutions,
since energy savings are dynamic policies that depend on the application
demands and device battery levels at a given moment in time.
%
There are also low-power standards for the IoT
infrastructure that suit different needs of range, throughput, and physical
distribution~\cite{iot.energy.2}.
For instance, \emph{Bluetooth Low-Energy (BLE)} is a replacement for classic
Bluetooth and is designed for lower data throughput in personal area
networks (PANs).
\emph{6LoWPAN} adapts the IPv6 internet standard to low-power and
limited-processing devices.
%
These technologies make radio transmissions more efficient,
support flexible network topologies, and reduce data traffic considerably.
They also enable sleep modes that cut energy consumption to minimum levels.
%
However, these technologies ultimately require software to control their
functionalities and to wisely switch between standby and active modes in order
to build an energy-efficient IoT.

%\newpage
\section{Scientific Summary}

Effective use of low-power standby will play a fundamental role in energy
efficiency for the expected 50 billion IoT devices by 2020~\cite{iea.data}.
%In order to succeed in this challenge, new solutions have to scale to the
%forthcoming mass of IoT software that must use standby modes effectively.
%
This research project aims to address the software challenges, as determined by
IEA, towards an energy-efficient IoT~\cite{iea.data}:
    to ensure that devices employ the lowest possible modes of standby, and
    that devices remain in longest possible periods of standby.
%
\begin{comment}
    \item To ensure that devices employ the lowest possible modes of standby consumption.
    \item To ensure that devices remain in longest possible periods of standby time.
\end{comment}

Given the projected scale of the IoT and the role of low-power standby towards
energy efficiency, this research project has the following goals:

\begin{enumerate}
    \item Address energy efficiency through extensive use of standby.
    \item Target constrained embedded architectures that form the IoT.
    \item Provide standby mechanisms at the programming language level that scale to all applications.
    \item Support transparent/non-intrusive standby mechanisms that reduce barriers of adoption.
\end{enumerate}

% TODO: goal 1 and 3 are also the limitations: only standby, not holistic energy savings

%\goodbreak\noindent
Many energy-aware languages, extensions, and operating systems have been
proposed recently.
%
Some proposals adjust QoS (quality of service) to reduce power consumption~\cite{os.ecosystem,lang.green,lang.enerj,lang.greenweb}.
%(e.g., accuracy of computations)
%os.odyssey.2,os.jouleguard,lang.lab,lang.flexjava,
%
Other proposals offer mechanisms to switch behaviors depending on application
demands and battery levels~\cite{lang.eon,lang.energytypes,lang.gradual,lang.ent}.
%(e.g., disable some functionalities)
%
None of these initiatives take advantage of standby modes
when idle,
but only adapt or eliminate computations while in active modes (not
addressing goal 1).
%
%There are also specialized network protocols that make devices remain in
%low-power standby modes for longer periods of time~\cite{wsn.energy}.
%
%However, protocol-based initiatives only apply to the networked parts of
%applications and have to be programmed carefully to take
%advantage of standby modes (not addressing goals 3 and 4).
%(which is typically the most power hungry subsystem),
%
%This proposal is also unique in the aim to provide \emph{transparent
%mechanisms} such that there are no explicit programming efforts to make
%applications more energy efficient (goal 4 above).

I have been working in the design of a new reactive programming language
targeting resource-constrained embedded systems for the past 8 years.
%
The language is grounded on the synchronous concurrency model,
%which has been
%successfully adopted in the context of hard real-time systems such as avionics
%and automobiles industry since the 80's~\cite{rp.twelve}.
%
%The synchronous model
which trades power for reliability and has a simpler model
of time that suits most requirements of IoT applications.
%
%On the one hand, this model cannot directly express time-consuming
%computations, such as compression and cryptography algorithms, which are
%typically either absent or delegated to auxiliary chips in the context of the
%IoT.
%
%On the other hand,
In this model, all reactions to the external world are guaranteed to be
computed in bounded
time, ensuring that applications always reach an idle state amenable to standby
mode.
%
%Furthermore, reactions to events are part of the vocabulary of the language.
%This way, when an application reaches an idle state, the language has precise
%information about which events can awake that application.
%This not only allows the application to enter standby mode, but effectively use
%the deepest sleeping level considering all possible awaking events.
%While in standby, only a hardware interrupt associated with the events can
%awake the application, making it sleep for longest possible periods of time.
%Despite the advances in energy-aware languages, embedded systems and are still
%primarily programmed in the C language~\cite{es.c}.
%
%This proposal lies at the bottom of the software development
%layers---transparent programming language mechanisms---meaning that \emph{all}
%applications would take advantage of low-power standby modes automatically,
%without extra programming efforts.
%
We expect that existing energy-unaware applications will benefit from savings
in the order of 50\% based on IEA's estimates considering current standby
technologies~\cite{iea.data}, and previous third-party work on transparent energy
awareness~\cite{wsn.tos.2}.

%\newpage
\section{Scientific Project}

\subsection{First Year}
\label{sec.first}

We will require a budget of R\$70k:
R\$60k to support three students, R\$5k for equipment, and
R\$5k for publication and travel costs.
The details are described in the long proposal.

\subsubsection{IoT Hardware Infrastructure}

%The IoT infrastructure is built on top of embedded systems such as sensor
%nodes, network routers, and other low-power microcontroller-based devices and
%appliances.
%
We will use off-the-shelf Arduinos~\cite{arduino} as the main hardware IoT
platform.
Most Arduinos are based on low-cost microcontrollers from
Atmel, such as the \emph{ATmega328p}~\cite{arduino.atmega328p}, supporting
six sleeping modes that can reduce power consumption to very low levels.

%Arduino is also the most popular embedded platform worldwide.
%
    In the academia, there is a lot of research using Arduino in the context of
    the IoT~\cite{arduino.infra,arduino.health,arduino.home,arduino.energy}.
% (e.g., infrastructure~\cite{arduino.infra},
    %healthcare~\cite{arduino.health}, home automation~\cite{arduino.home}, and
    %energy awareness~\cite{arduino.energy}).
    %In addition to the availability of support literature,
    The popularity of
    Arduino will make our own research more accessible and reproducible to
    other groups.
%
    In education, many University courses use Arduino~\cite{arduino.edu.1,arduino.edu.2,arduino.edu.3,arduino.edu.4}.
    We have also been using Arduino in an undergraduate course for the
    past 3 years,
    %Arduino lowers the learning barrier for students, and
    which will allow us to evaluate
    our results with less experienced embedded programmers.
%
    In the hobbyists community, there is an abundance of publicly available software
    which we can adapt to our
    language to evaluate energy efficiency gains.
%
    %Among non-programmers, Arduino is also becoming popular and a lot of
    %software will be produced by them in the near future.
    %We are in the process of creating an interdisciplinary course targeting
    %non-specialists (e.g., arts, design, architecture).
    %This audience will be a good target to evaluate our transparent approach for
    %energy efficiency.

\subsubsection{IoT Software Infrastructure}
\label{sec.method.software}

The typical way to interact with the external world in Arduino is through
polling, which samples an
external device actively to detect changes on its state. % to take an action.
Polling wastes CPU cycles and prevents the device to enter in standby.
In Arduino, even basic functionality, such as timers, A/D converters, and
SPI, uses active polling loops that waste energy in active mode.

In order to provide automatic standby, applications need to be entirely
reactive to events.
%
Recently, we have added support for \emph{interrupt service routines (ISRs)} as
a primitive concept of our language, which will allow us to rebuild the IoT
software infrastructure with standby awareness from the ground up.
%
This process will consist primarily of rewriting device drivers, which are the
pieces of software that interface directly with the hardware.
%
%More concretely, at the lowest level, energy-efficient software depends on
%\emph{interrupt service routines (ISRs)}, which awake the CPU from standby when
%an external peripheral event occurs.
%
This approach will not affect how applications are written at higher levels,
which will remain similar to Arduino applications.
However, instead of wasting cycles in busy-wait loops, the applications will
enter the deepest possible standby mode when idle.

\subsubsection{IoT Applications}
\label{sec.method.apps}

In order to evaluate the gains in energy efficiency with the proposed
infrastructure, we will need to evaluate the consumption of realistic applications.
%
The Arduino community has an abundance of open-source projects which can be
rewritten in our language to take advantage of transparent standby modes.
%
%Also, most peripherals for Arduino ship with drivers and applications that
%explore their functionalities (e.g., accelerometers, touch screens, GPS, etc.).
%
Then, we will be able to compare the original and rewritten versions in terms of energy
consumption to draw conclusions about the effectiveness of transparent standby
modes.
%
The most realistic scenarios for the IoT use radio communication extensively.
%
In this context, we will evaluate from simple handmade ad-hoc protocols to more
complex energy-aware protocols and see to what extent our proposal would
effectively contribute to energy savings.
%
%The \emph{RadioHead} project%
%\footnote{RadioHead project: \url{http://www.airspayce.com/mikem/arduino/RadioHead/}}
%supports dozens of RF (radio frequency) chipsets and provides examples that
%explore ad-hoc mesh as well as low-power energy modes.
%
%We plan to support at least two chipsets, such as \emph{nRF24L01}~\cite{radio.nrf24l01}
%and \emph{RFM69HCW}~\cite{radio.rfm69hcw}, both of which have good support for
%low power consumption.
%
In the long term, we expect to make the case for
developers to rewrite their applications in our language to take advantage of
standby modes for free.
%
%Towards this direction, we will evaluate rewritten IoT applications based on the
%following criteria:
Towards this direction, we will evaluate the time it takes to rewrite an application and
the real gains in energy efficiency.
%
\begin{comment}
\item[Time to rewrite:]
This criteria relates to the incentives to rewrite existing applications to
our language.
It is a tradeoff between the expected rewriting times and gains in energy
efficiency as compared to standard Arduino code.
\item[Coding ``aesthetics'':]
To lower the adoption barrier, it is also important that the proposed
programming style is sufficiently familiar, and, at least, as easy to write applications.
\item[Energy consumption:]
More importantly, rewritten applications should have significant gains in
energy efficiency to justify a complete application rewrite.
\end{comment}

\subsubsection{\textbf{Expected Contributions}}

%\begin{enumerate}

%\item \textbf{An environment-aware system language:}
    The dedicated vocabulary of our language for events raises the
    programming abstraction to a level closer to the domain of IoT, providing
    more safety and expressiveness to programmers.
    As far as we know, ISR support at the language level has not been
    tried in the past.
    %The novel support for ISRs would complete the whole development cycle, from
    %the application down to the hardware, without operating system support.

%\item \textbf{An energy-aware system language:}
    Our proposal aims to make all applications subject to standby modes
    transparently.
    Being part of the software infrastructure, only device drivers will require
    explicit energy management, and all applications built on top of them will
    benefit from energy efficient automatically.

%\item \textbf{A realistic programming alternative for the IoT in the Arduino:}
    Our language is a 8-year project and has a mature open-source
    implementation that is publicly available.
    With the adaptation to the context of energy-efficient IoT, the language
    may become a practical alternative for Arduino in the short term.

%\item \textbf{Effective energy-efficient IoT applications:}
    %We expect substantial gains in energy efficiency for all applications that
    %do not manage energy explicitly.
    %Most Arduino APIs use busy-wait polling to wait for input, which are very
    %energy inefficient, and could take advantage of transparent standby.
    %We also expect gains in comparison to some ad-hoc energy-aware
    %implementations---we believe that, for the most cases, a transparent
    %solution is more effective than explicit programming efforts.
%\end{enumerate}

%\newpage
\subsection{The Three Years to Follow}

\subsubsection{Systematic Energy Awareness}
\label{sec.method.systematic}

Systematic energy awareness
considers the IoT network as a whole, with cooperating devices trying to
maximize energy efficiency globally.
%
In this context, \emph{adaptive computing}~\cite{adaptive} is the capability of
the IoT to adapt energy consumption dynamically, based on application demands
and levels of power supply during execution.
%
Ultimately, the goal is still that each device maximizes its idle
periods to be more susceptible to standby modes.
Hence, the research of the
first year is a prerequisite for this phase.
%
We will continue to investigate how language mechanisms can aid energy
efficiency, but now in the context of adaptive computing.

The research method for systematic energy awareness will be similar to the
one adopted for device-based awareness (as discussed in
Section~\ref{sec.method.apps}):
take existing energy-aware IoT protocols and applications for Arduino, rewrite
them using non-intrusive mechanisms, and use the same evaluation criteria
(i.e., time to rewrite, coding aesthetics, and energy consumption).

\subsubsection{Complex IoT Architectures and Applications}
\label{sec.method.complex}

%The niche of constrained embedded systems, which includes Arduino, covers the
%substantial (and increasing) portion of IoT applications.
%They typically do not require significant computing resources but are sensitive
%to physical dimensions and energy consumption.
%
%However,
The IoT also consists of more traditional networked devices, such as
routers, servers, and smartphones.
%
As of 2016, there were 3.9 billion smartphone subscriptions worldwide, and
this number is expected to reach 6.8 billions by 2022~\cite{ericsson.mobility}.

Smartphones have similar restrictions in battery consumption and could also
take advantage of the techniques we propose for constrained embedded systems.
%
In order to transpose the barrier from constrained devices to smartphones for
the IoT, we will take a similar path as presented in Section~\ref{sec.first}:
%
\begin{description}
\item[Hardware Infrastructure]
We will use the BeagleBone Black~\cite{bbb.manual}, which shares similar goals with
Arduino, providing a cheap and open-source platform but which is suitable for rich
applications, such as graphical user interfaces, multimedia, and games.
\item[Software Infrastructure]
In order to ensure automatic standby for applications, all software
infrastructure, mostly device drivers, needs to be rebuilt from scratch using
ISRs in our language.
\item[Applications]
In addition to IoT applications, typical smartphone applications, such as
instant messaging and internet browsing, can also maximize energy efficiency
through standby.
We will rewrite from simple applications, such as a graphical clock, to more
complex networked applications, such as a web browser, and evaluate their energy consumption.
\end{description}

\subsubsection{\textbf{Expected Contributions}}

%\begin{enumerate}

%\item \textbf{Adaptive and energy-efficient IoT applications:}
    We expect that systematic energy awareness will take advantage of the
    non-intrusive mechanisms of our language to be implemented at a higher
    level of abstraction, resulting in energy-efficient applications that
    require less programming efforts.

%\item \textbf{Energy-efficient smartphones:}
    The transition from simple mobile phones to smartphones resulted in a
    degrading effect on battery lives.
    %Phones used to have autonomy of weeks and now require daily recharges.
    Most of the time, however, smartphones are sitting in our pockets wasting
    energy.
    We expect to increase the battery autonomy considerably while keeping the
    functionality of a modern smartphone.
%\end{enumerate}

\begin{comment}
%\newpage
\section{CONCLUSION}

- story
    - environment awareness
    - boot up a language
    - test in the environment
    - energy awareness
    - ISRs
    - test in the environment
    - OS
    - C is 45 years

TODO

- conslidated research
- some steps already done
- we need human resources

realistic research
    - academic results
    - but also an open-source publicly available language that people can download and use
        - already available
        - all publics: research, hobbyists, students

TODO: long-term research
    - 8 years in the same project
\end{comment}

\newpage
\bibliographystyle{abbrv}
\bibliography{serra}
\end{document}

### 2018-01-29 LCTES/body.tex
(+) init

--- 2018-01-05 Submissao/big.tex
@@
-\documentclass[12pt,english]{amsart}
-%\usepackage[margin=0.5in]{geometry}
-\usepackage[a4paper, margin=2cm]{geometry}
-\linespread{1.5}
-
-%\usepackage{fontspec}
-%\setmainfont{Times New Roman}
-\usepackage{mathptmx}
-
-\usepackage{graphicx}
-\usepackage{wrapfig}
-\usepackage{lscape}
-\usepackage{rotating}
-\usepackage{epstopdf}
-
-\usepackage{verbatim}
-\usepackage{url}
-\usepackage{xspace}
-
-\usepackage{listings}
-\usepackage{color}
-    \definecolor{light}{gray}{0.97}
-    \definecolor{dark}{gray}{0.30}
-\lstset{
-%columns=fullflexible,
-%basicstyle=\ttfamily,
-escapeinside={||},
-    %mathescape=true,
-    language=C, % choose the language of the code
-    basicstyle=\fontfamily{pcr}\selectfont\scriptsize\color{black},
-    keywordstyle=\color{black}\bfseries, % style for keywords
-    numbers=none, % where to put the line-numbers
-    numberstyle=\tiny, % the size of the fonts that are used for the line-numbers
-    backgroundcolor=\color{light},
-    showspaces=false, % show spaces adding particular underscores
-    showstringspaces=false, % underline spaces within strings
-    showtabs=false, % show tabs within strings adding particular underscores
-    %frame=single, % adds a frame around the code
-    tabsize=2, % sets default tabsize to 2 spaces
-    %rulesepcolor=\color{gray}
-    captionpos=b, % sets the caption-position to bottom
-    breaklines=false, % sets automatic line breaking
-    %breakatwhitespace=false,
-    numbersep=2em,
-    % C was used in the blocksworld example to refer to block C and nowhere else
-    emph={par,or,hor,do,end,loop,code,await,pause,emit,input,event,call,with,%
-          var,and,then,else,return,pure,deterministic,nohold,finalize,%
-          class, every, FOREVER, this, spawn, in, pool, watching, until, 
-          interface, each, abort, when, signal, PROC, CHAN, SIGNAL, PAR, not,
-          bool, tag, escape, traverse,implementation,output,true,false,
-          native,@const,@pure,@safe,define,public,private,none},
-    emphstyle={\bfseries},
-    commentstyle=\color{dark}\scriptsize,
-    %xleftmargin=20pt,
-    %xrightmargin=20pt,
-    framesep=20pt,
-    %upquote=true,
-    %aboveskip={1.5\baselineskip},
-}
-
-\newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
-\newcommand{\code}[1] {{\small{\texttt{#1}}}}
-
-\usepackage{enumitem}
-\setlist{nolistsep}
-
-% Auto-Standby
-\title{Energy Efficiency for IoT Software in the Large
-        } %\\ \large{Research project proposal submitted to the Instituto Serrapilheira}}
-
-%\author{Anonymous Author}
-
-\begin{document}
-
-\date{}
-\maketitle
-
-\vspace{-0.5cm}
-\begin{comment}
-\abstract{
-The annual power consumption of network-connected devices worldwide is
-estimated to be greater than Germany's total electricity consumption.
-By 2020, there will be 50 billion ``things'' connected to the \emph{Internet of
-Things (IoT)}.
-%
-Currently, most of the energy to power these devices is wasted in \emph{standby
-mode}, when they are not performing their main functionality.
-%
-Considering this alarming scenario, the International Energy Agency (IEA)
-launched an initiative to ensure that IoT devices consume a minimal amount of
-energy while in standby mode.
-%
-IEA estimates that it is possible to save 65\% of the energy used by
-network-enabled devices with existing technologies.
-
-This research project aims to address the software challenges, as determined by
-IEA, towards an energy-efficient IoT:
-    to ensure that devices power down to lowest possible consumption modes,
-    and that they remain there for longest possible periods of time.
-We propose to investigate transparent mechanisms in a programming language that
-ensure deep and long periods of standby, meaning that \emph{all} programs ever
-written in this language would benefit from low-power modes automatically,
-without extra programming efforts.
-}
-\end{comment}
-
-%\newpage
-
-\section{INTRODUCTION}
+\section{Introduction}
 \label{sec.introduction}
@@
 According to the International Energy Agency (IEA)~\cite{iea.data}, there were
-around 14 billion traditional network-connected devices in 2013 (e.g., mobile
-phones and smart TVs).
+around 14 billion traditional network-connected devices in 2013.
+% (e.g., mobile phones and smart TVs).
 This number is expected to increase to 50 billion by 2020 with
-the proliferation of IoT devices (e.g., smart bulbs and fitness wearables).
+the proliferation of IoT devices. % (e.g., smart bulbs and fitness wearables).
 %
-IoT and traditional network-connected devices already outnumber people on the
-planet by a factor of two, and the amount of data traffic is expected to grow
-at an exponential rate in the next years.
+%IoT and traditional network-connected devices already outnumber people on the
+%planet by a factor of two, and the amount of data traffic is expected to grow
+%at an exponential rate in the next years.
 %By 2050, the energy demands of networked devices should exceed the current
@@
 %
-The annual $CO_2$ emissions related to standby in Australia are equivalent to
-those of 1 million cars.
+For instance, standby power accounts for 10--15\% of residential electricity
+consumption, and $CO_2$ emissions related to standby are equivalent to those of
+1 million cars~\cite{iea.data,standby.australia}.
 %
@@
 
-Many other organizations have also reported the importance of energy savings
-specifically in the context of the IoT and networked devices.
-%
-In 2010, the Internet Engineering Task Force (IETF) created a working group on energy
-management (EMAN) with the following statement:
-``Energy management is becoming an additional requirement for network
-management systems due to several factors including the rising and
-fluctuating energy costs, the increased awareness of the ecological
-impact of operating networks and devices, and the regulation of
-governments on energy consumption and production''~\cite{ietf.eman}.
-%
-The American Council for an Energy-Efficient Economy (ACEEE) states that
-``The potential for new energy efficiency remains enormous, (...) looking
-ahead we must take a systems-based approach to dramatically scale up energy
-efficiency.
-%Intelligent efficiency is a systems-based approach to efficiency that can help
-%to meet this need.
-(...) intelligent efficiency differs from component energy efficiency in that
-it is adaptive, anticipatory, and networked.''~\cite{aceee.1}
-%
-The pioneer \emph{ENERGY STAR} program~\cite{energystar} states that
-``networked devices and networking equipment, as integrated systems, have the
-potential to contribute significant net energy savings''.
-% \footnote{ENERGY STAR program \url{https://www.energystar.gov/index.cfm?c=prod_development.prod_development_epa_workshop}}
-%
-The US Environmental Protection Agency (EPA) and Information Technology
-Industry Council (ITI) held a workshop to explore roles for the \emph{ENERGY
-STAR} and others program in promoting savings through intelligent energy
-efficiency~\cite{aceee.2}.
-%, which differs from individualized
-%component-based energy efficiency in that it is adaptive, anticipatory, and
-%networked.
+In this work
 
-With regard to concrete initiatives, IEA's \emph{Electronic Devices and Networks}
-focuses specifically on the issue of networked device standby%
-\footnote{EDNA initiative: \url{https://edna.iea-4e.org/}}.
-%It is one of the six fronts of G20's \emph{Energy Efficiency Action Plan}%
-%\footnote{\url{https://www.iea-4e.org/projects/g20}}.
-%Network standby refers to the minimum power a device requires to maintain its
-%connection when not communicating or computing actively.
-ACEEE's \emph{Intelligent Efficiency} promotes a systematic approach
-to optimize the behavior of cooperating devices in order to achieve energy
-savings as a whole.
-Both approaches, device and system based, involve mostly software solutions,
-since energy savings are dynamic policies that depend on the application
-demands and device battery levels at a given moment in time.
-%
-There is also a number of low-power wireless standards for the IoT
-infrastructure that suit different needs of range, throughput, and physical
-distribution~\cite{iot.energy.2}.
-For instance, \emph{Bluetooth Low-Energy (BLE)} is a replacement for classic
-Bluetooth and is designed for lower data throughput in personal area
-networks (PANs).
-\emph{6LoWPAN} adapts the IPv6 internet standard to low-power and
-limited-processing devices.
-%
-These technologies make radio transmissions more efficient,
-support flexible network topologies, and reduce data traffic considerably.
-They also enable sleep modes that cut energy consumption to minimum levels.
-%
-However, these technologies ultimately require software to control their
-functionalities and to wisely switch between standby and active modes in order
-to build an energy-efficient IoT.
+Synchronous language
 
-\subsection{Objectives}
-\label{sec.goals}
-
-%The annual power consumption of network-connected devices worldwide is
-%estimated to be greater than Germany's total electricity consumption.
-%By 2020, there will be 50 billion ``things'' connected to the \emph{Internet of
-%Things (IoT)}.
-%
-%Considering this alarming scenario, the International Energy Agency (IEA)
-%launched an initiative to ensure that IoT devices consume a minimal amount of
-%energy while in standby mode.
-
-Effective use of low-power standby will play a fundamental role in energy
-efficiency for the expected 50 billion IoT devices by 2020~\cite{iea.data}.
-In order to succeed in this challenge, new solutions have to scale to the
-forthcoming mass of IoT software that must use standby modes effectively.
-%
-This research project aims to address the software challenges, as determined by
-IEA, towards an energy-efficient IoT~\cite{iea.data}:
 %
@@
 
-\subsection{State of the Art}
-\label{sec.related}
+\begin{acks}
+  The authors would like to thank Dr. Yuhua Li for providing the
+  MATLAB code of the \textit{BEPS} method.
 
-Many energy-aware languages, extensions, and operating systems have been
-proposed recently.
-%
-Some proposals adjust QoS (quality of service) to reduce power consumption
-(e.g., accuracy of computations, data sampling rates, scheduling times,
-etc.)~\cite{os.ecosystem,os.odyssey.2,lang.green,lang.enerj,os.jouleguard,lang.lab,lang.flexjava,lang.greenweb}.
-%
-Other proposals offer mechanisms to switch behaviors depending on application
-demands and battery levels (e.g., disable functionalities, switch UI,
-etc.)~\cite{lang.eon,lang.energytypes,lang.gradual,lang.ent}.
-%
-None of these initiatives are concerned with taking advantage of standby modes
-when idle,
-but more with adapting or eliminating computations while in active modes (not
-addressing goal 1 of Section~\ref{sec.goals}).
-%
-There are also specialized network protocols that make devices remain in
-low-power standby modes for longer periods of time~\cite{wsn.energy}.
-%
-However, protocol-based initiatives only apply to the networked parts of
-applications and have to be programmed carefully to take
-advantage of standby modes (not addressing goals 3 and 4).
-%(which is typically the most power hungry subsystem),
-%
-%This proposal is also unique in the aim to provide \emph{transparent
-%mechanisms} such that there are no explicit programming efforts to make
-%applications more energy efficient (goal 4 above).
+  The authors would also like to thank the anonymous referees for
+  their valuable comments and helpful suggestions. The work is
+  supported by the \grantsponsor{GS501100001809}{National Natural
+    Science Foundation of
+    China}{http://dx.doi.org/10.13039/501100001809} under Grant
+  No.:~\grantnum{GS501100001809}{61273304}
+  and~\grantnum[http://www.nnsf.cn/youngscientists]{GS501100001809}{Young
+    Scientists' Support Program}.
 
-\begin{comment}
-\emph{Green}~\cite{lang.green} enables applications to approximate expensive
-computations based on QoS constraints specified by the programmer.
-%
-\emph{GreenWeb}~\cite{lang.greenweb} provides energy-aware HTML extensions to
-specify expected levels of QoS related to a webpage responsiveness.
-The language runtime determines how to deliver the target QoS experience while
-minimizing energy consumption.
-%
-\emph{Energy Types} and \emph{ENT} proposes a type system to specify phases
-(i.e., workload demands) and modes (i.e., expected energy state).
-The language runtime monitors the current mode and only executes admissible
-phases.
-%
-\emph{EnerJ}~\cite{lang.enerj} also proposes a type system in which primitive
-types (e.g., floats and integers) can be annotated as \emph{approximate}.
-The language then uses non-exact but more energy efficient operations on these
-types.
-%
-Unlocking energy
-    - locks
-        - busy wait vs sleep
-        - CPU sleep mode
-
-MEANTIME: achieving both minimal energy and timeliness with approximate computing
-    - approximate computing
-\end{comment}
-
-Despite the advances in energy-aware languages, embedded systems are still
-primarily programmed in the C language~\cite{es.c}.
-%
-The predominance of C is associated with its portability across architectures,
-efficiency in terms of memory and CPU, extensive legacy codebase, and number of
-programmers.
-%
-Therefore, the full IoT development stack typically relies on
-C~\cite{wsn.tos,wsn.contiki,wsn.riot}, from high-level applications and
-network protocols, down to operating systems, network drivers, and SoC
-firmwares.
-%
-However, C is a simple hardware abstraction (sometimes referred to as the
-\emph{portable assembly}) and has no awareness of the surrounding environment
-it is embedded on.
-%
-For instance, there is no dedicated vocabulary in the language to express
-concepts that naturally appear in IoT applications, such as time, communication
-with the real world, concurrency of events, and energy awareness.
-%
-C is also memory unsafe, leading to all sorts of bugs such as memory leaks,
-buffer overflows, and wild pointers~\cite{c.unsafe}.
-
-TinyOS~\cite{wsn.tos} uses a C dialect~\cite{wsn.nesc} and partially addresses
-our goals.
-It targets constrained architectures and maximizes standby time (goals 1 and 2).
-TinyOS is also transparent at the level of the operating system, meaning that
-applications built on top of it will take advantage of standby power for free
-(goals 3 and 4).
-They claim to operate in sleep modes 44\% of the time~\cite{wsn.tos.2}, which
-is close to IEA's estimates of 65\% energy savings.
-However, an operating system codebase is complex and still written in C.
-Our goal is to provide transparent support at the language level, which even an
-OS implementation could take advantage (goal 3).
-Another issue with TinyOS is related to the adoption barrier imposed by its
-programming model (goal 4): applications have to be written in a
-callback-driven style in C, which is widely recognized as hard and error
-prone~\cite{sync_async.cooperative,wsn.protothreads,rp.goto,rp.deprecating,frp.rx}
-%
-%TODO: 1.1 register scan
-
-\subsection{Preliminary Results}
-\label{sec.ceu}
-
-I have been working in the design of \CEU, a new reactive programming language
-targeting resource-constrained embedded systems, for the past 8
-years~\cite{ceu.sensys11,ceu.tr,ceu.sensys13,ceu.rem13,ceu.phd,ceu.rebls14,
-ceu.mod15,ceu.rebls15,ceu.terra,ceu.media.webmedia16,ceu.tecs17}.
-%
-\CEU is grounded on the synchronous concurrency model, which has been
-successfully adopted in the context of hard real-time systems such as avionics
-and automobiles industry since the 80's~\cite{rp.twelve}.
-%
-The synchronous model trades power for reliability and has a simpler model
-of time that suits most requirements of IoT applications.
-%
-On the one hand, this model cannot directly express time-consuming
-computations, such as compression and cryptography algorithms, which are
-typically either absent or delegated to auxiliary chips in the context of the
-IoT.
-%
-On the other hand, all reactions to the external world are guaranteed to be
-computed in bounded
-time, ensuring that applications always reach an idle state amenable to standby
-mode.
-
-In a previous work~\cite{ceu.sensys13}, we adapted \CEU to execute on top of
-TinyOS in the context of Wireless Sensor Networks, and developed a number of
-applications, protocols, and device drivers.
-We evaluated the expressiveness of \CEU in comparison to C and attested a
-reduction in source code size (around 25\%) with a small increase in memory
-usage (around 5-10\% for text and data).
-We also evaluated CPU responsiveness under heavy network load and attested a
-similar performance between \CEU and C, which implies comparable energy
-efficiency while in active mode.
-%
-In a related work, we developed Terra, a virtual machine for \CEU targeting
-constrained networked devices that supports low-power remote
-reprogramming~\cite{ceu.terra}.
-Virtual machine applications are an order of magnitude smaller than full
-binaries, reducing data traffic considerably and, consequently, energy
-consumption.
-
-In a recent paper~\cite{ceu.tecs17}, we discuss in extent the design decisions
-behind \CEU under the perspective of
-synchronous languages, providing a deep comparison with the seminal language
-Esterel~\cite{esterel.ieee91}.
-The most fundamental difference to Esterel is \CEU's event-triggered notion of
-time, in which the occurrence of an event (e.g., a button press) defines an
-indivisible instant in an application logical timeline.
-This characteristic makes the model of time of \CEU exclusively reactive to
-incoming events.
-As a consequence, time does not elapse during idle periods, making all
-applications susceptible to standby. % modes.
-
-\subsection{A Motivating Example}
-
-{\linespread{1}
-\begin{figure}[t]
-\begin{minipage}[t]{0.50\linewidth}
-\begin{lstlisting}[xrightmargin=1cm]
-digitalWrite(13, 0);
-int old = 0;
-while (1) {
-    int cur = digitalRead(2);
-    if (cur != old) {
-        digitalWrite(13, cur);
-        old = cur;
-    }
-}
-
-\end{lstlisting}
-\centering\small{[a] Active polling in C}
-\end{minipage}
-%
-\begin{minipage}[t]{0.45\linewidth}
-\begin{lstlisting}[numbers=left]
-input  int PIN_02;
-output int PIN_13;
-
-emit PIN_13(0);
-loop do
-    var int v = await PIN_02;
-    emit PIN_13(v);
-end
-.
-\end{lstlisting}
-\centering\small{[b] Reactive execution in \CEU}
-\end{minipage}
-%\rule{8.4cm}{0.37pt}
-\caption{ Controlling a LED from button state changes.
-\label{lst.basic}
-}
-\end{figure}
-}
-
-Figure~\ref{lst.basic} shows a basic program that makes a button connected
-to the microcontroller's pin 2 to control a LED connected to pin 13.
-Whenever the state of the button changes (i.e., pressed or unpressed), the
-state of the LED also changes (i.e., on or off).
-
-The implementation in C in Figure~\ref{lst.basic}.a first turns off the LED
-(ln. 1), then sets an initial state for the button (ln. 2), and then reads in a
-loop (ln. 3-9) its current state (ln. 4) as fast as possible.
-If the current state differs from the old state (ln. 5-8), the new state is
-copied to the LED (ln. 6), and is also set as the old state (ln. 7).
-%
-This approach which actively samples the state of an external device (e.g., a
-button) is known as \emph{polling} or \emph{software-driven I/O}.
-%
-On the one hand, this style does not involve dealing with the complexity of
-callbacks (as discussed in Section~\ref{sec.related}) and is widely adopted in
-Arduino, the most popular embedded platform (to be discussed in
-Section~\ref{sec.method.software}).
-On the other hand, polling wastes CPU cycles and prevents the device to enter in
-standby.
-
-The implementation in \CEU in Figure~\ref{lst.basic}.b
-uses dedicated vocabulary to handle events from I/O devices without resorting
-to callbacks.
-The pins are first declared to determine their purpose in the program, i.e.,
-\code{input} for the pin connected to the button and  \code{output} for the pin
-connected to the LED (ln 1-2).
-Then, the LED is turned off with an \code{emit} (ln. 4), which indicates an
-output operation.
-Then, the program enters in a loop to react to changes in the button (ln. 5-8).
-The \code{await} primitive accounts for the reactive nature of \CEU.
-The program literally waits for a change in the button (ln. 6) to be able to
-proceed and set the new state of the LED (ln 7).
-%
-This way, when an application reaches an idle state, the language has precise
-information about which events can awake that application.
-
-By design, all lines of execution in \CEU always reach an \code{await} at the
-end of a reaction to an event (otherwise, the application does not
-compile~\cite{ceu.sensys13}).
-%
-This not only allows the application to enter standby mode, but effectively use
-the deepest sleeping level considering all possible awaking events.
-While in standby, only a hardware interrupt associated with the events can
-awake the application, making it sleep for longest possible periods of time.
-%
-%This is still a theoretical assumption that needs to be proven in practice.
-
-\section{METHODOLOGY}
-
-\subsection{First Year}
-\label{sec.first}
-
-\subsubsection{IoT Hardware Infrastructure}
-
-The IoT infrastructure is built on top of embedded systems such as sensor
-nodes, network routers, and other low-power microcontroller-based devices and
-appliances.
-
-We will use off-the-shelf Arduinos~\cite{arduino} as the main hardware IoT
-platform.
-Most Arduinos are based on low-cost 8-bit microcontrollers from Atmel, such as
-the popular \emph{ATmega328p}~\cite{arduino.atmega328p},
-%This chip includes the \emph{picoPower} technology,
-which supports six sleeping
-modes which can reduce power consumption to very low levels.
-Depending on some configurations (e.g., frequency and voltage), an Arduino can
-draw from 45mA at full operation, down to 5uA in power down mode (the deepest
-sleeping mode).
-We will pursue that IoT applications operate only 50\% of the time in the
-average, which is not unlikely~\cite{wsn.tos.2}.
-Considering the negligible consumption in power down mode, this will meet our
-goals of reducing energy consumption by 50\%.
-% 45mA@5V = 0.2W: below 1W, but no radio and other devices
-
-Besides the good support for low-power standby, Arduinos are inexpensive,
-open source, and widely used.
-%
-We will take advantage of the many environments in which Arduino is popular:
-%
-\begin{description}
-\item[Research]
-    There is a lot of research using Arduino in the context of the IoT (e.g.,
-    infrastructure~\cite{arduino.infra}, healthcare~\cite{arduino.health},
-    home automation~\cite{arduino.home}, and energy
-    awareness~\cite{arduino.energy}).
-    In addition to the availability of support literature, the popularity of
-    Arduino will make our own research more accessible and reproducible to
-    other groups.
-\item[Education]
-    Arduino is used in many University courses worldwide
-    (e.g. \cite{arduino.edu.1,arduino.edu.2,arduino.edu.3,arduino.edu.4}).
-    We have been using Arduino with \CEU in an undergraduate course for the
-    past 3 years, which will
-    %Arduino lowers the learning barrier for students, and will
-    allow us to evaluate
-    our results with less experienced embedded programmers.
-\item[Hobbyists]
-    The Arduino hobbyist community cultivates an abundance of resources such as
-    forums, wikis, and publicly available software.
-    It is important for us to have available software that we can adapt to \CEU
-    to evaluate energy efficiency gains.
-\item[Non-Programmers]
-    Arduino is also becoming popular among non-programmers and a lot of
-    software will be produced by them in the near future.
-    We are in the process of creating an interdisciplinary course targeting
-    non-specialists (e.g., arts, design, architecture).
-    This audience will be a good target to evaluate our transparent approach for
-    energy efficiency.
-\end{description}
-
-%TODO: radio
-
-\paragraph{\textbf{Expected Contributions}} %\mbox{}\\
-
-Since we are adopting a third-party hardware platform with existing support for
-low power modes, we do not expect significant contributions.
-
-\paragraph{\textbf{Risks (very low: 1/5)}}
-
-Arduino is a cheap and popular embedded platform with widespread use in the
-academia.
-%
-There is a considerable number of reports that claim to achieve very low power
-consumption with Arduino in deep standby modes.
-%
-%TODO: radio
-
-\begin{comment}
-http://www.home-automation-community.com/arduino-low-power-how-to-run-atmega328p-for-a-year-on-coin-cell-battery/
-
-- HIGH
-    - 5V
-        - 45 mA current
-- LOW
-    - POWER-DOWN SLEEP
-        - 5V
-            - 23 uA
-            - 5.8 uA (no power regulator)
-        - 3V3
-            - 54 uA (why higher than 5V?)
-            - 4.5 uA (no power regulator)
-
-ATmega328P Pro Mini Version	PWR Source	State	5V@16MHz    3V3@8MHz
-Unmodified 	                RAW Pin	    ACT	    19.9 mA	    4.74 mA
-Unmodified 	                RAW Pin	    PDS	    3.14 mA	    0.90 mA
-No Power LED 	            RAW Pin	    ACT	    16.9 mA	    3.90 mA
-No Power LED 	            RAW Pin	    PDS	    0.0232 mA*	0.0541 mA*
-No Power LED, no Regulator	VCC Pin	    ACT	    12.7 mA	    3.58 mA
-No Power LED, no Regulator	VCC Pin	    PDS	    0.0058 mA	0.0045 mA
-
-That is 4 years on a 9 V battery with 1,200 mAh capacity or 2,000 times more efficient than the Arduino Uno
-
-===
-
-https://github.com/rocketscream/Low-Power
-
-===
-
-https://www.avrprogrammers.com/howto/atmega328-power
-
-- HIGH
-    - 5V@16MHz
-        - 16mA
-    - 3V3@1MHz
-        - 92uA
-- LOW
-        - 100uA
-
-Frequency	mA@5V	mA@3.3V
-1MHz	    6.75	0.92
-8MHz	    11.68	4.16
-16MHz (LP)	16.32	7.40
-16MHz (FS)	17.52	7.94
-20MHz (LP)	19.71	8.86
-20MHz (FS)	21.12	9.46
-
-Peripheral	mA@3V3@8MHz
-ADC	        0.23
-USART	    0.08
-SPI	        0.15
-Timer0	    0.09
-Timer1	    0.12
-Timer2	    0.12
-TWI	        0.18
-
-They can each be turned off in the Power Reduction Register (PRR) by writing a
-one to the bit position.
-
-In this example code, we shut down the CPU and wake it again when the INT0
-interrupt fires (the INT0 pin goes low). The average current draw is 380µA@5V,
-which is pretty low. At 3.3V the current drops to 72uA, but if we wake up and
-write to an SD card, for instance, the current draw will jump to 13mA for
-100mS, then 50mA for 50mS, before it drops back down to the 72µA level again.
-\end{comment}
-
-\subsubsection{IoT Software Infrastructure}
-\label{sec.method.software}
-
-Arduino provides a thin software layer on top of C and C++, and the typical way
-in which applications are built is through polling (as discussed in
-Section~\ref{sec.ceu}).
-In Arduino, even basic functionality, such as timers
-    (\code{delayMicroseconds}%
-        %\footnote{Source code for \code{delayMicroseconds}: \url{https://github.com/arduino/Arduino/blob/master/hardware/arduino/avr/cores/arduino/wiring.c#L120}}%
-    ),
-A/D converters
-    (\code{analogRead}%
-        %\footnote{Source code for \code{analogRead}: \url{https://github.com/arduino/Arduino/blob/master/hardware/arduino/avr/cores/arduino/wiring_analog.c#L38}}%
-    ),
-and SPI
-    (\code{SPI.transfer}%
-        %\footnote{Source code for \code{SPI.transfer}: \url{https://github.com/arduino/Arduino/blob/master/hardware/arduino/avr/libraries/SPI/src/SPI.h#L208}}%
-    ),
-uses busy-wait loops that waste energy in active mode.
-%\footnote{USART: \url{https://www.arduino.cc/en/Serial/Available}\\SPI:}%
-
-In order to provide automatic standby, applications need to be entirely
-reactive to events, and we will rebuild all of the software infrastructure from
-scratch in \CEU.
-%
-This process will consist primarily of rewriting device drivers, which are the
-pieces of software that interface directly with the hardware.
-%
-More concretely, at the lowest level, energy-efficient software depends on
-\emph{interrupt service routines (ISRs)}, which awake the CPU from standby when
-an external peripheral event occurs.
-
-Recently, we have added support for ISRs as a primitive concept of
-\CEU, which will allow us to rebuild the IoT software infrastructure with
-standby awareness from the ground up.
-%
-Figure~\ref{lst.isrs} is an sketch of our intended approach to achieve transparent
-standby modes (the lines with \code{<...>} are pseudo code).
-%
-The application in Figure~\ref{lst.isrs}.a requests, every hour (ln. 7-11), an
-A/D conversion (ln. 8) and awaits its completion (ln. 9) to perform some action
-(collapsed in ln. 10).
-%
-Note that the application uses abstract input/output events to interface with
-the A/D converter (ln. 3-4) and does not manage energy explicitly.
-%
-Figure~\ref{lst.isrs}.b would be the associated driver with the A/D converter,
-which implements the input/output events declared as abstractions in the
-application code.
-The output request (ln. 3-7) configures the A/D peripheral (ln. 4) and enables
-interrupts in the microcontroller (ln. 5).
-It also needs to set the deepest mode supported by the A/D peripheral (ln. 6)
-so that the language uses it as soon as the program reaches the idle state.
-The input definition (ln. 9-13) implements the ISR for the A/D converter.
-When the conversion is done, the peripheral awakes the CPU, and this code
-executes immediately in interrupt context.
-The code disables A/D interrupts (ln. 10), reads the A/D converter data
-register with the converted value (ln. 11), and returns this data to the
-application (ln. 12).
-A \code{return} in interrupt context will enqueue the input event
-\code{ADC\_DONE}, which will awake the application code associated with it
-(ln. 9 in Figure~\ref{lst.isrs}.a).
-
-With this approach, the application code remains similar to the one would write
-for Arduino (apart from minor syntax mismatches).
-However, instead of wasting cycles in busy-wait loops, the applications will
-enter the deepest possible standby mode when idle.
-
-{\linespread{1}
-\begin{figure}[t]
-\begin{minipage}[t]{0.56\linewidth}
-\begin{lstlisting}[xrightmargin=1cm]
-// application.ceu
-
-output none ADC_REQUEST;
-input  int  ADC_DONE;
-#include "adc.ceu" // driver implementation
-
-every 1h do
-    emit ADC_REQUEST;
-    var int v = await ADC_DONE;
-    <performs-some-action>;
-end
-
-.
-\end{lstlisting}
-\centering\small{[a] Application: no explicit energy management}
-\end{minipage}
-%
-\begin{minipage}[t]{0.43\linewidth}
-\begin{lstlisting}[numbers=left]
-// adc.ceu
-
-output none ADC_REQUEST do
-    <configures-ADC>;
-    <enables-ADC-interrupts>;
-    <sets-the-deepest-standby-mode>;
-end
-
-input int ADC_DONE [ADC_vect_num] do
-    <disables-ADC-interrupts>;
-    var int v = <reads-adc-data>;
-    return v;
-end
-\end{lstlisting}
-\centering\small{[b] Driver: explicit energy management }
-\end{minipage}
-%\rule{8.4cm}{0.37pt}
-\caption{ ISRs in \CEU
-\label{lst.isrs}
-}
-\end{figure}
-}
-
-\paragraph{\textbf{Expected Contributions}}
-
-\begin{enumerate}
-\item \textbf{An environment-aware system language:}
-    \CEU's dedicated vocabulary to deal with events raises the
-    programming abstraction to a level closer to the domain of IoT, providing
-    more safety and expressiveness to programmers.
-    These results have already been partially published in the
-    past~\cite{ceu.sensys13}, when we implemented a radio device driver on top
-    of TinyOS.
-    Nevertheless, the novel support for ISRs would complete the whole
-    development cycle, from the application down to the hardware, without
-    operating system support.
-\item \textbf{An energy-aware system language:}
-    The method described in this section makes all applications subject to
-    standby modes transparently.
-    Only device drivers, which are typically provided by vendors, will require
-    explicit energy management.
-    In our case, we will need to rewrite each driver only once, as part of the
-    software infrastructure, and all applications built on top of them would
-    benefit from energy efficient automatically.
-%\item[TODO: education]
-\end{enumerate}
-
-\paragraph{\textbf{Risks (high: 4/5)}}
-
-%\begin{enumerate}
-%\item \textbf{Primitive language support for ISRs:}
-As far as we know, abstract ISR support at the language level has not been
-tried in the past.
-The challenge is to choose the right level of abstraction that is, at the same
-time, powerful to support the range of devices and portable to work in multiple
-platforms.
-Since \CEU integrates seamlessly with C~\cite{ceu.tecs17}, it is always
-possible to circumvent abstractions, at least in early stages of development,
-until the appropriate level is found.
-%\item \textbf{Programming efforts:}
-
-From the engineering perspective, rewriting drivers is a laborious endeavour.
-As a rough reference, about half of the 15 million lines of the Linux codebase
-consists of device drivers.
-With the tight integration between \CEU and C, however, significant parts of
-existing drivers can be reused.
-%\end{enumerate}
-
-We already have some evidence of the feasibility of programming device drivers
-in \CEU based on a project of this year's \emph{Google SoC}%
-\footnote{GSoC project: \url{http://www.lua.inf.puc-rio.br/gsoc/ideas2017.html#ceu}}.
-In the period of 3 months, an undergraduate student studied and implemented
-drivers for an A/D and SPI peripherals using the recent support for
-ISRs in \CEU.
-
-\subsubsection{IoT Applications}
-\label{sec.method.apps}
-
-In order to evaluate the gains in energy efficiency with the proposed
-infrastructure, we will need to evaluate the consumption of realistic applications.
-%
-The Arduino community has an abundance of open-source projects which can be
-rewritten in \CEU to take advantage of transparent standby modes (as illustrated
-in Figure~\ref{lst.isrs}).
-%
-%Also, most peripherals for Arduino ship with drivers and applications that
-%explore their functionalities (e.g., accelerometers, touch screens, GPS, etc.).
-%
-Then, we will be able to compare the original and rewritten versions in terms of energy
-consumption to draw conclusions about the effectiveness of transparent standby
-modes.
-
-We will start with simple applications, such as the \emph{Blink} that ships
-with the Arduino distribution and blinks an LED every second.
-The code is reproduced in Figure~\ref{lst.blink}.a and a very similar version
-in \CEU appears in Figure~\ref{lst.blink}.b.
-%
-The example is in active mode for a negligible time just to set the I/O ports
-(ln. 3 and 5), remaining idle most of the time.
-%
-We expect this application to consume much less energy in \CEU, since the
-\code{await} statements (ln. 4 and 6) will put the device into standby
-automatically, while the version in C uses a busy-wait loop inside the
-\code{delay} statements (ln. 4 and 6).
-
-{\linespread{1}
-\begin{figure}[t]
-\begin{minipage}[t]{0.49\linewidth}
-\begin{lstlisting}[xrightmargin=1cm]
-pinMode(13, OUTPUT);
-while (1) {
-    digitalWrite(13, 1);
-    delay(1000);
-    digitalWrite(13, 0);
-    delay(1000);
-}
-\end{lstlisting}
-\centering\small{[a] Blinking a LED in C}
-\end{minipage}
-%
-\begin{minipage}[t]{0.49\linewidth}
-\begin{lstlisting}[numbers=left]
-output int PIN_13;
-loop do
-    emit PIN_13(1);
-    await 1s;
-    emit PIN_13(0);
-    await 1s;
-end
-\end{lstlisting}
-\centering\small{[b] Blinking a LED in \CEU}
-\end{minipage}
-%\rule{8.4cm}{0.37pt}
-\caption{ Blinking a LED connected to pin 13 every second.
-\label{lst.blink}
-}
-\end{figure}
-}
-
-The most realistic scenarios for the IoT use radio communication extensively.
-%
-In this context, we will evaluate from simple handmade ad-hoc protocols to more
-complex energy-aware protocols (as discussed in Section~\ref{sec.related}) and
-see to what extent our proposal would effectively contribute to energy savings.
-%
-The \emph{RadioHead} project%
-\footnote{RadioHead project: \url{http://www.airspayce.com/mikem/arduino/RadioHead/}}
-supports dozens of RF (radio frequency) chipsets and provides examples that
-explore mesh topologies as well as low-power energy modes.
-%
-We plan to support at least two chipsets: \emph{nRF24L01}~\cite{radio.nrf24l01}
-and \emph{RFM69HCW}~\cite{radio.rfm69hcw}, both of which have good support for
-low power consumption.
-%
-%We expect that applications that use energy-aware protocols extensively (but
-%not much other peripherals) will not benefit from our proposal, since they
-%implement standby policies explicitly (see also
-%Section~\ref{sec.method.systematic}).
-%
-%Nevertheless, we can still evaluate the implementation of these protocols and
-%assert that the new implementations in \CEU do not have explicit standby
-%policies but are still as energy efficient than the originals.
-
-In the long term, we expect to make the case for
-developers to rewrite their applications in \CEU to take advantage of standby
-modes for free.
-%
-Towards this direction, we will evaluate IoT applications rewritten in \CEU based on
-the following criteria:
-%
-\begin{description}
-\item[Time to rewrite]
-This criteria relates to the incentives to rewrite existing applications to
-\CEU.
-It is a tradeoff between the expected rewriting times and gains in energy
-efficiency as compared to standard Arduino code.
-\item[Coding ``aesthetics'']
-To lower the adoption barrier, it is also important that the proposed
-programming style is sufficiently familiar, and, at least, as easy to write applications.
-\item[Energy consumption]
-More importantly, rewritten applications should have significant gains in
-energy efficiency to justify a complete application rewrite.
-\end{description}
-
-\paragraph{\textbf{Expected Contributions}}
-\begin{enumerate}
-%\item \textbf{Higher-level programming abstractions:}
-    %Similarly to IoT software infrastructure (discussed in
-    %Section~\ref{sec.method.software}), we expect that applications will take
-    %even more advantage of the abstraction level provided by \CEU.
-\item \textbf{A realistic programming alternative for the IoT in the Arduino:}
-    \CEU is a 8-year project and has a mature open-source implementation that
-    is publicly available%
-\footnote{\CEU repository: \url{https://github.com/fsantanna/ceu}}.
-    With the adaptation to the context of energy-efficient IoT, \CEU may cross
-    the academic fences in the short term and become a practical alternative
-    for the Arduino community.
-\item \textbf{Energy-efficient IoT applications:}
-    We expect substantial gains in energy efficiency for all applications that
-    do not manage energy explicitly.
-    Most Arduino APIs use busy-wait polling to wait for input, which are very
-    energy inefficient, and could take advantage of transparent standby.
-    %We also expect gains in comparison to some ad-hoc energy-aware
-    %implementations---we believe that, for the most cases, a transparent
-    %solution is more effective than explicit programming efforts.
-\end{enumerate}
-
-\paragraph{\textbf{Risks (medium: 3/5)}}
-
-%\begin{enumerate}
-%\item \textbf{Effective energy efficiency:}
-We do not know precisely the practical gains in energy efficiency in the
-standby modes of Arduino.
-Most reports that compare theoretical (from official datasheets)
-and measured consumption claim good performance but are non peer-reviewed
-research.
-For instance, a specialized company%
-\footnote{RocketScream: \url{http://www.rocketscream.com}}
-provides Arduino-based IoT devices and a cross-platform library dedicated to
-standby modes that is widely used by the community.
-%\end{enumerate}
-
-\subsection{The Three Years to Follow}
-
-\subsubsection{Systematic Energy Awareness}
-\label{sec.method.systematic}
-
-As discussed in Section~\ref{sec.introduction}, systematic energy awareness
-considers the IoT network as a whole, with cooperating devices trying to
-maximize energy efficiency globally.
-%
-In this context, \emph{adaptive computing}~\cite{adaptive} is the capability of
-the IoT to adapt energy consumption dynamically, based on application demands
-and levels of power supply during execution.
-%
-Ultimately, the goal is still that each device maximizes its idle
-periods to be more susceptible to standby modes.
-Hence, the research of the
-first year is a prerequisite for this phase.
-
-We will continue to investigate how language mechanisms can aid energy
-efficiency, but now in the context of adaptive computing.
-%
-Figure~\ref{lst.pause} is a sketch of an application that adapts its behavior
-under low-power mode by suspending parts of its functionality (the lines with
-\code{<...>} are pseudo code).
-%
-The application relies on the \code{par} construct of \CEU (ln. 3-16) which
-creates three concurrent lines of execution (ln. 4, 6-8, and 10-15).
-The first line of execution (collapsed in ln. 4) implements the mandatory
-functionality of the application, which must execute regardless of the power
-mode.
-The second line of execution (ln. 6-8) implements extra functionality
-(collapsed in ln. 7) that can
-be suspended in low-power mode without affecting the main functionality.
-The \code{pause/if} construct of \CEU (ln. 6) suspends the nested costly
-functionality whenever the event
-\code{enter\_low\_power\_mode} is set to \emph{true} and resumes when set to
-\emph{false}.
-The third line of execution is responsible for implementing the adaptive
-manager by interacting with the external world.
-In this case, every time the button is pressed (ln. 11 and 13), the device switches
-the power mode by triggering the event \code{enter\_low\_power\_mode}
-(ln. 12 and 14).
-
-{\linespread{1}
-\begin{figure}[t]
-\begin{lstlisting}[xleftmargin=1cm,numbers=left]
-input none BUTTON_PRESSED;             // input for every button press
-event bool enter_low_power_mode;       // event to toggle power mode
-par do
-    <mandatory-functionality>          // omitted code with basic functionality
-with
-    pause/if enter_low_power_mode do   // the nested block is suspended in low mode
-        <energy-costly-functionality>  // omitted code with costly functionality
-    end
-with
-    loop do                            // toggles power mode on every button press
-        await BUTTON_PRESSED;
-        emit enter_low_power_mode(true);
-        await BUTTON_PRESSED;
-        emit enter_low_power_mode(false);
-    end
-end
-\end{lstlisting}
-%\rule{8.4cm}{0.37pt}
-\caption{ Adaptive application that changes behavior depending on the power mode.
-\label{lst.pause}
-}
-\end{figure}
-}
-
-The example suggests that \code{BUTTON\_PRESSED} represents a simple button
-connected to the device.
-%
-However, the advantage of using the abstract concept of an input event is that
-the actual implementation may vary.
-%
-For instance, \code{BUTTON\_PRESSED} might be a ``virtual button'' associated to a
-remote device (or set of devices).
-%
-We will take advantage of the abstract concepts of input and output events to be
-able to specify IoT interactions at a higher level.
-
-Note that adaptive computations must be programmed explicitly, since they
-are application dependent and have particular requirements.
-%
-Hence, since it is not possible to offer transparent language mechanisms (such
-as in our proposal for standby modes), we will investigate \emph{non-intrusive
-language mechanisms}.
-For instance, note in the example that the \code{<energy-costly-functionality>}
-(ln. 7) did not suffer any internal modifications to be suspended, but was
-simply enclosed by the \code{pause/if} construct.
-%
-This way, making an application adaptive to embrace energy efficiency will
-require less programming efforts.
-Furthermore, the implementation of the application functionalities will not be
-intermixed with energy awareness, making the code more modular.
-
-The evaluation process for systematic energy awareness will be similar to the
-one adopted for device-based awareness (as discussed in
-Section~\ref{sec.method.apps}):
-take existing energy-aware IoT protocols and applications for Arduino, and
-rewrite them using non-intrusive mechanisms.
-% TODO: which apps?
-Then, employ the same evaluation criteria
-(i.e., time to rewrite, coding aesthetics, and energy consumption).
-%
-\begin{comment}
-\item[Time to rewrite:]
-This criteria relates to the incentives to rewrite existing applications to
-\CEU.
-It is a tradeoff between the expected gains in energy efficiency and rewriting
-time.
-\item[Coding ``aesthetics'':]
-To lower the adoption barrier, it is also important that the proposed
-programming style is sufficiently familiar, and, at least, as easy to write.
-\item[Energy consumption:]
-More importantly, rewritten applications should have significant gains in
-energy efficiency to justify a complete rewrite.
-\end{comment}
-
-\paragraph{\textbf{Expected Contributions}}
-\begin{enumerate}
-\item \textbf{Higher-level programming abstractions:}
-    Similarly to the IoT software infrastructure and applications (discussed in
-    Sections~\ref{sec.method.software}~and~\ref{sec.method.apps}), we expect
-    that systematic energy awareness will take advantage of the non-intrusive mechanisms
-    that \CEU provides and will be implemented at a higher level of abstraction.
-\item \textbf{Adaptive and energy-efficient IoT applications:}
-    We expect to achieve the same energy efficiency as that of hand-tweaked
-    applications but with less programming efforts.
-\end{enumerate}
-
-\paragraph{\textbf{Risks (medium: 3/5)}}
-
-%\begin{enumerate}
-%\item \textbf{Effective energy efficiency:}
-On the one hand, hand-tweaked applications always have a theoretical higher
-efficiency than applications based on richer abstractions.
-On the other hand, having to deal with the complexity of hand-tweaked
-applications makes them less amenable to adaptive computations.
-In some situations, we may face this tradeoff and sacrifice adaptiveness to
-achieve satisfactory energy efficiency.
-%\item \textbf{Programming efforts:}
-
-From the engineering perspective, rewriting complex applications takes a lot of
-time.
-However, considering the period of 4 years, time estimates can
-be reconsidered periodically to employ more human resources into the project.
-%\end{enumerate}
-
-%\vspace{-1cm}
-\subsubsection{Complex IoT Architectures and Applications}
-\label{sec.method.complex}
-
-The niche of constrained embedded systems, which includes Arduino, covers the
-substantial (and increasing) portion of IoT applications.
-They typically do not require significant computing resources but are sensitive
-to physical dimensions and energy consumption.
-%
-However, the IoT also consists of more traditional networked devices, such as
-routers, servers, and smartphones.
-%
-As of 2016, there were 3.9 billion smartphone subscriptions worldwide, and
-this number is expected to reach 6.8 billions by 2022~\cite{ericsson.mobility}.
-
-Smartphones rely on much more complex architectures than microcontroller-based
-embedded systems (e.g., 32-bit CPUs with memory management unit).
-They typically rely on an operating system, a complete TCP/IP stack, and can
-execute multiple applications at the same time.
-In addition to reactive applications, typical of the IoT, smartphones also
-perform active computations, such as audio/image/video processing and
-cryptographic functions.
-Nevertheless, they are an important piece of the IoT, serving as a common human
-interface to process, visualize, and actuate on the network.
-
-Smartphones have similar restrictions in battery consumption and could also
-take advantage of the techniques we propose for constrained embedded systems.
-%
-%Therefore, we will address energy efficiency through transparent language
-%mechanisms to achieve extensive use of standby.
-%
-In order to transpose the barrier from constrained devices to smartphones for
-the IoT, we will take a similar path as presented in Section~\ref{sec.first}:
-%
-\begin{description}
-\item[Hardware Infrastructure]
-We will use the BeagleBone Black~\cite{bbb.manual}, which shares similar goals with
-Arduino, providing a cheap and open-source platform but which is suitable for rich
-applications, such as graphical user interfaces, multimedia, and games.
-It is based on the Texas Instruments (TI) Sitara ARM
-architecture~\cite{bbb.sitara}, featuring a 1GHz CPU, with 512MB RAM and 2GB
-flash.
-Like happens with Arduino, the BeagleBoard community offers an abundance of
-open projects and publicly available software.
-\item[Software Infrastructure]
-In order to ensure automatic standby for applications, all software
-infrastructure, mostly device drivers, needs to be rebuilt from scratch using
-ISRs in \CEU.
-However, smartphones have much more complex peripherals, such as WiFi adapters
-and graphics processors.
-To reduce the barrier adoption,
-TI provides \emph{StarterWare}%
-\footnote{TI Sitara: \url{http://www.ti.com/tool/starterware-sitara}},
-a free software development package with open
-support for the basic Sitara peripheral drivers.
-\item[Applications]
-In addition to IoT applications, typical smartphone applications, such as
-instant messaging and internet browsing, can also maximize energy efficiency
-through standby.
-We will rewrite from simple applications, such as a graphical clock, to more
-complex networked applications such as a web browser:
-\begin{itemize}
-\item \textbf{xclock} displays the time in analog or digital and the source code is around 5k lines in C.
-% TODO: 15-20 loc
-\item \textbf{Dillo}%
-\footnote{Dillo web browser: \url{https://www.dillo.org/}}
-is a minimalistic web browser with the source code around 50k lines in C.
-\end{itemize}
-\end{description}
-
-\paragraph{\textbf{Expected Contributions}}
-
-\begin{enumerate}
-\item \textbf{Energy-efficient smartphones:}
-    The transition from simple mobile phones to smartphones resulted in a
-    degrading effect on battery lives.
-    Phones used to have autonomy of weeks and now require daily recharges.
-    Most of the time, however, smartphones are sitting in our pockets wasting
-    energy.
-    We expect to increase the battery autonomy considerably while keeping the
-    functionality of a modern smartphone.
-\item \textbf{Towards an environment- and energy-aware operating system:}
-    The whole software technology of Smartphones is built on top of bloated and
-    antiquated personal computer operating systems like Windows and UNIX (e.g.,
-    macOS and Android).
-    These operating systems were not designed with mobile systems and energy
-    awareness in mind.
-    Our proposal is a step towards a new environment- and energy-aware
-    operating system.
-\end{enumerate}
-
-\paragraph{\textbf{Risks (very high: 5/5)}}
-%\begin{enumerate}
-%\item \textbf{Programming efforts:}
-Modern smartphone hardware architectures are much more complex than of
-constrained embedded systems.
-For this reason, making an application to be energy-aware from the ground up,
-starting from a hardware interrupt, is an enduring challenge.
-Nonetheless, the proposed language infrastructure is the same, and all other
-required pieces such as drivers are publicly available.
-%\end{enumerate}
-
-\begin{comment}
-- Rust is a memory-safe alternative to C~\cite{rust.os}
-    - problems with embedded systems
-        \cite{rust.problems}
-
-- Ceu
-    - interoperable with C
-
-- HW platform Mini + NRF
-    - radio + microcontroller + development board
-\end{comment}
-
-\section{TIMELINE}
-
-Figure~\ref{fig.timeline} shows the project timeline for the full period of 4
-years.
-The two lines in black refer to the \emph{constrained} and \emph{complex} IoT
-architectures phases (Sections~\ref{sec.first}/\ref{sec.method.systematic} and
-Section~\ref{sec.method.complex}, respectively).
-They only intersect during the second semester of the second year, when the
-constrained phase is about to terminate and the complex phase kicks off.
-At the end of the first year, we expect to address energy efficiency for the
-constrained IoT architectures, with the hardware and software
-infrastructure set, and the basic and networked applications completed
-(Section~\ref{sec.first}).
-The second year covers systematic energy awareness for the constrained IoT
-(Section~\ref{sec.method.systematic}) and marks the beginning of the complex IoT
-phase (Section~\ref{sec.method.complex}), which extends to the end of the
-period of 4 years.
-
-The small tasks in red are related to hardware acquisition and setup of
-existing IoT architectures.
-The tasks in yellow refer to the software infrastructure, which will be in
-continuous iterations during the whole project cycle.
-They require the most experienced researchers and consist of designing the
-language mechanisms, employing automatic standby, and implementing the most
-essential drivers.
-The tasks in green refer to the applications built on top of the software
-architecture.
-They have delimited development cycles and require less experienced
-researchers.
-
-\begin{sidewaysfigure}
-    \vspace{17cm}
-    \includegraphics[width=\columnwidth]{serra-big}
-    \caption{Project timeline
-        \label{fig.timeline}
-    }
-\end{sidewaysfigure}
-
-\section{RESOURCES}
-
-For the period of 3 years, we will require a budget of R\$345k:
-\begin{itemize}
-    \item R\$300k for staff payment (87\%),
-    \item R\$20k for equipment (6\%), and
-    \item R\$25k for publication and travel costs (7\%).
-\end{itemize}
-
-\subsection{The Team}
-
-Most of the resources will be employed in research staff consisting
-of undergraduate, MSc, and PhD students.
-%
-The quality and long-term commitment of the students is paramount for the
-success of the project.
-%
-For this reason, we want to offer a salary compatible with the industry, which
-is considerably larger than typical research scholarships.
-
-\begin{description}
-\item[Proponent]
-The proponent will work in all project fronts: managing the resources and
-staff, writing papers and supervising students, and working on the technical
-challenges of the project.
-\item[Advisors]
-The proponent is an associate professor at UERJ and a member of the LabLua%
-\footnote{LabLua: \url{http://www.lua.inf.puc-rio.br/}.}
-at PUC-Rio.
-The advisors are experienced researchers from UERJ and PUC-Rio in the field of
-programming languages, distributed systems, and wireless sensor networks.
-They will support the academic duties, such as prospecting and supervising
-students, writing papers, and helping on the general project guidelines.
-\item[Research staff]
-At the beginning of the first year, we will appoint two students to join the
-project full time:
-    one early-stage PhD student
-and
-    one MSc student or late-stage undergraduate student.
-Every six months, we plan to add a new member to the staff until we reach a
-limit of 6 students (consisting of 2 or 3 PhD students).
-The PhD students will work with the software infrastructure and the other
-students with the applications.
-We also expect that the PhD students will work on academic papers and
-take advantage of the project in their thesis.
-Considering an average of 4 students during the whole project with and an
-average salary of R\$2k, we expect to spend around R\$300k with payments.
-\end{description}
-
-\subsection{Infrastructure and Equipment}
-
-The required equipment is divided between support for the research staff (printer and
-laptops) and the project embedded systems (Arduino, Beagle Black, etc.):
-\newline
-
-%{%\linespread{0.5}
-{\renewcommand{\arraystretch}{0.8}
-\begin{tabular}{ | l | r | r | r | }
-\hline			
-EQUIPMENT       & PRICE & QTD   &   TOTAL   \\
-\hline  
-Laser printer   & 3000  &   1   &   3.000    \\
-Laptop          & 4000  &   3   &  12.000    \\
-Arduino kits    &  200  &   5   &   1.000    \\
-Sensors \& radios & -   &   -   &   2.000    \\
-Beagle Black    &  600  &   2   &   1.200    \\
-Beagle Display  &  400  &   2   &     800    \\
-\hline  
-                &       &       &  R\$20k \\
-\hline  
-\end{tabular}
-\newline
-}
-
-We expect that half of these resources will be acquired at the beginning of the
-project.
-The other half will be acquired gradually during the course of the 3 years.
-
-\subsection{Publication and Travel Costs}
-
-Our goal is to publish our results on international journals in the field of
-embedded systems and sensor networks, such as \emph{ACM TECS} and \emph{ACM TOSN}.
-%
-However, the process to publish a paper in a journal may take longer than a year after
-the submission.
-For this reason, we also intend to publish partial results in conferences.
-
-During the 3 years, we plan to attend two Brazilian conferences (e.g.,
-\emph{SBRC}),
-%and the \emph{Brazilian Symposium on Computing Systems Engineering (SBESC)}),
-and one prestigious international conference (e.g., \emph{ACM SenSys}),
-These venues are important to cultivate future cooperation with other research
-groups.
-For the Brazilian conferences, we will support 2 participants, and for the
-international conference, only 1 participant.
-%
-Using the CNPq travel grants as a reference, attending a national conference
-costs around R\$3k (including travel and accommodation for 3 days), while a 
-international conference costs around US\$4k (including travel and
-accommodation for 5 days).
-Considering the two Brazilian conferences with two participants each, and one
-international conference with one participant, the total travel costs for the
-3-year period will be around R\$25k.
-
-\begin{comment}
-\section{CONCLUSION}
-
-TODO
-
-- conslidated research
-- some steps already done
-- we need human resources
-
-realistic research
-    - academic results
-    - but also an open-source publicly available language that people can download and use
-        - already available
-        - all publics: research, hobbyists, students
-
-TODO: long-term research
-    - 8 years in the same project
-\end{comment}
-
-\newpage
-\bibliographystyle{abbrv}
-\bibliography{serra,my}
-\end{document}
+\end{acks}

### 2018-01-29 LCTES/lctes.tex
(+) init

\documentclass[sigplan, anonymous, review]{acmart}

\usepackage{booktabs} % For formal tables

% Copyright
%\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% DOI
\acmDOI{10.475/123_4}

% ISBN
\acmISBN{123-4567-24-567/08/06}

%Conference
\acmConference[WOODSTOCK'97]{ACM Woodstock conference}{July 1997}{El
  Paso, Texas USA}
\acmYear{1997}
\copyrightyear{2016}

\acmPrice{15.00}

%\acmBadgeL[http://ctuning.org/ae/ppopp2016.html]{ae-logo}
%\acmBadgeR[http://ctuning.org/ae/ppopp2016.html]{ae-logo}

\begin{document}
\title{SIG Proceedings Paper in LaTeX Format}
\titlenote{Produces the permission block, and
  copyright information}
\subtitle{Extended Abstract}
\subtitlenote{The full version of the author's guide is available as
  \texttt{acmart.pdf} document}

\author{Ben Trovato}
\authornote{Dr.~Trovato insisted his name be first.}
\orcid{1234-5678-9012}
\affiliation{%
  \institution{Institute for Clarity in Documentation}
  \streetaddress{P.O. Box 1212}
  \city{Dublin}
  \state{Ohio}
  \postcode{43017-6221}
}
\email{trovato@corporation.com}

\author{G.K.M. Tobin}
\authornote{The secretary disavows any knowledge of this author's actions.}
\affiliation{%
  \institution{Institute for Clarity in Documentation}
  \streetaddress{P.O. Box 1212}
  \city{Dublin}
  \state{Ohio}
  \postcode{43017-6221}
}
\email{webmaster@marysville-ohio.com}

\author{Lars Th{\o}rv{\"a}ld}
\authornote{This author is the
  one who did all the really hard work.}
\affiliation{%
  \institution{The Th{\o}rv{\"a}ld Group}
  \streetaddress{1 Th{\o}rv{\"a}ld Circle}
  \city{Hekla}
  \country{Iceland}}
\email{larst@affiliation.org}

\author{Valerie B\'eranger}
\affiliation{%
  \institution{Inria Paris-Rocquencourt}
  \city{Rocquencourt}
  \country{France}
}
\author{Aparna Patel}
\affiliation{%
 \institution{Rajiv Gandhi University}
 \streetaddress{Rono-Hills}
 \city{Doimukh}
 \state{Arunachal Pradesh}
 \country{India}}
\author{Huifen Chan}
\affiliation{%
  \institution{Tsinghua University}
  \streetaddress{30 Shuangqing Rd}
  \city{Haidian Qu}
  \state{Beijing Shi}
  \country{China}}

\author{Charles Palmer}
\affiliation{%
  \institution{Palmer Research Laboratories}
  \streetaddress{8600 Datapoint Drive}
  \city{San Antonio}
  \state{Texas}
  \postcode{78229}}
\email{cpalmer@prl.com}

\author{John Smith}
\affiliation{\institution{The Th{\o}rv{\"a}ld Group}}
\email{jsmith@affiliation.org}

\author{Julius P.~Kumquat}
\affiliation{\institution{The Kumquat Consortium}}
\email{jpkumquat@consortium.net}

% The default list of authors is too long for headers.
\renewcommand{\shortauthors}{B. Trovato et al.}

\begin{abstract}
This paper provides a sample of a \LaTeX\ document which conforms,
somewhat loosely, to the formatting guidelines for
ACM SIG Proceedings.
\end{abstract}

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below.
%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010575.10010755</concept_id>
  <concept_desc>Computer systems organization~Redundancy</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010553.10010554</concept_id>
  <concept_desc>Computer systems organization~Robotics</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>10003033.10003083.10003095</concept_id>
  <concept_desc>Networks~Network reliability</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Computer systems organization~Embedded systems}
\ccsdesc[300]{Computer systems organization~Redundancy}
\ccsdesc{Computer systems organization~Robotics}
\ccsdesc[100]{Networks~Network reliability}

\keywords{ACM proceedings, \LaTeX, text tagging}

\begin{teaserfigure}
  %\includegraphics[width=\textwidth]{sampleteaser}
  \caption{This is a teaser}
  \label{fig:teaser}
\end{teaserfigure}

\maketitle

\input{body}

\bibliographystyle{ACM-Reference-Format}
\bibliography{serra,my,standby}

\end{document}

### 2018-01-30 LCTES/body.tex
(>) LCTES

@@
 
-According to the International Energy Agency (IEA)~\cite{iea.data}, there were
-around 14 billion traditional network-connected devices in 2013.
-% (e.g., mobile phones and smart TVs).
-This number is expected to increase to 50 billion by 2020 with
-the proliferation of IoT devices. % (e.g., smart bulbs and fitness wearables).
+According to the International Energy Agency (IEA), the number of
+network-connected devices is expected to reach to 50 billion by 2020 with
+the expansion of the Internet of Things (IoT).~\cite{iea.data}
 %
-%IoT and traditional network-connected devices already outnumber people on the
-%planet by a factor of two, and the amount of data traffic is expected to grow
-%at an exponential rate in the next years.
-%By 2050, the energy demands of networked devices should exceed the current
-%consumption of Germany and Canada combined, corresponding to 6\% of current
-%global electricity consumption.
+However, most of the energy to power these devices will be consumed in
+\emph{standby mode}, i.e., when they are neither transmitting or processing
+data.
 %
-However, most of the energy to power these devices is consumed while they are
-in \emph{standby mode}, i.e., when their software is neither transmitting or
-processing data.
+For instance, standby power currently accounts for 10--15\% of residential
+electricity consumption, and $CO_2$ emissions related to standby are equivalent
+to those of 1 million cars~\cite{iea.data,standby.australia}.
 %
-For instance, standby power accounts for 10--15\% of residential electricity
-consumption, and $CO_2$ emissions related to standby are equivalent to those of
-1 million cars~\cite{iea.data,standby.australia}.
-%
-%This undesirable inefficiency opens opportunities for improvements in energy
-%savings closer to 65\% only considering currently available technologies.
-%
-The projected growth of IoT devices together with the surprising effects of
+The projected growth of IoT devices, together with the surprising effects of
 standby consumption, made network standby efficiency one of the six
@@
 
-In this work
-
-Synchronous language
-
-%
-\begin{itemize}
-    \item To ensure that devices employ the lowest possible modes of standby consumption.
-    \item To ensure that devices remain in longest possible periods of standby time.
-\end{itemize}
-
 Given the projected scale of the IoT and the role of low-power standby towards
-energy efficiency, this research project has the following goals:
+energy efficiency, this paper has the following goals:
 
@@
 \item Address energy efficiency through extensive use of standby.
-    \item Target constrained embedded architectures that form the IoT.
-    \item Provide standby mechanisms at the programming language level that scale to all applications.
-    \item Support transparent/non-intrusive standby mechanisms that reduce barriers of adoption.
+\item Target low-power, resource-constrained embedded architectures that form
+      the IoT.
+\item Provide standby mechanisms at the programming language level that scale
+      to all applications.
+\item Support transparent/non-intrusive standby mechanisms that reduce barriers
+      of adoption.
 \end{enumerate}
 
-% TODO: goal 1 and 3 are also the limitations: only standby, not holistic energy savings
+Our approach lies at the bottom of the software development
+layers---transparent programming language mechanisms---meaning that \emph{all}
+applications take advantage of low-power standby modes automatically, without
+extra programming efforts.
+%
+We extend the synchronous programming language
+\CEU~\cite{ceu.sensys13,ceu.tecs17} with interrupt service routines (ISRs) and
+a simple power management runtime (PMR).
+%
+Each supported microcontroller requires bindings in C for the ISRs and PMR, and
+each peripheral requires a driver in \CEU.
+This is a one-time process and is typically packaged and distributed in a SDK.
+%
+All new applications built on top of these drivers will take advantage of
+standby automatically.
+%
+As a proof of concept, we support the 8-bit \emph{AVR/ATmega} and 32-bit
+\emph{ARM/Cortex-M0} microcontrollers, and a variety of peripherals (e.g., ADC,
+SPI, USART, nRF24L01 transceiver).
+The project is open-source.%
+\footnote{https://github.com/fsantanna/ceu-arduino/}
 
-This proposal lies at the bottom of the software development
-layers---transparent programming language mechanisms---meaning that \emph{all}
-applications would take advantage of low-power standby modes automatically,
-without extra programming efforts.
+ Thank you, I will improve the page.
+
+In summary:
+
+Reactive: code executes in reactions to events
+
+Synchronous: reactions run to completion, i.e., there's no implicit preemption or real parallelism (this avoids explicit synchronization: locks, queues, etc)
+
+Structured: programs use structured control mechanisms, such as "await" (to suspend a line of execution), and "par" (to combine multiple awaiting lines of execution)
+
+Structured programming avoids deep nesting of callbacks letting you write programs in direct/sequential style.
+In addition, when a line of execution is aborted, all allocated resources are safely released.
+
+In comparison to FRP/dataflow, it is more imperative supporting sequences/loops/conditionals/parallels.
+The notion of (multiple) program counter is explicit.
+Also, everything is lexically scoped, there's no GC involved.
+
+In comparison to promises/futures, it provides lexical parallel constructs, allowing the branches to share local variables and, more importantly, supporting safe abortion of code (with the "par/or").
+
+- application on IoT
+- gains??
+
+- structured programming
+- lexical scope
+- automatic power management: standby as well as enabled/disabled, powered, switched
+
+needs a requires a energy-aware driver
+
+process for ISR bindings and
+A infra-structure programmer writes device drivers for
+Each peripheral requires a driver
+An architecture
+Infra
+Once a driver is written, all applications
+
+- synchronous languages / \CEU
+
+- except drivers, which are also written in Ceu
+
+I have been working in the design of \CEU, a new reactive programming language
+targeting resource-constrained embedded systems, for the past 8
+years~\cite{ceu.sensys11,ceu.tr,ceu.sensys13,ceu.rem13,ceu.phd,ceu.rebls14,
+ceu.mod15,ceu.rebls15,ceu.terra,ceu.media.webmedia16,ceu.tecs17}.
+%
+\CEU is grounded on the synchronous concurrency model, which has been
+successfully adopted in the context of hard real-time systems such as avionics
+and automobiles industry since the 80's~\cite{rp.twelve}.
+%
+The synchronous model trades power for reliability and has a simpler model
+of time that suits most requirements of IoT applications.
+%
+On the one hand, this model cannot directly express time-consuming
+computations, such as compression and cryptography algorithms, which are
+typically either absent or delegated to auxiliary chips in the context of the
+IoT.
+%
+On the other hand, all reactions to the external world are guaranteed to be
+computed in bounded
+time, ensuring that applications always reach an idle state amenable to standby
+mode.
+
+- results
+    - atmega 328p/2560
+    - arm cortex-m0
+    - how much efficiency?
 %
@@
 awareness~\cite{wsn.tos.2}.
+
+- limitations
+
+- paper structure
+
+In Section~\ref{sec.ceu}, we introduce the synchronous model together with the
+programming language \CEU.
+In Section~\ref{sec.ext}, we present the language extensions that support
+transparent standby.
+In Section~\ref{sec.eval}, we evaluate .
+
+\section{The Synchronous Language CEU}
+\label{sec.ceu}
+
+\section{Transparent Standby Mechanisms}
+\label{sec.ceu}
 

### 2018-01-30 LCTES/body.tex
(>) LCTES

@@
 
- Thank you, I will improve the page.
+- automatic power management: standby as well as enabled/disabled, powered, switched
+
+- application on IoT
+
+- results
+    - atmega 328p/2560
+    - arm cortex-m0
+    - how much efficiency?
+
+- limitations
+
+- paper structure
+
+In Section~\ref{sec.ceu}, we introduce the structured synchronous model and the
+programming language \CEU.
+In Section~\ref{sec.ext}, we present the language extensions that support
+transparent standby.
+In Section~\ref{sec.eval}, we evaluate .
+
+\section{The Structured Synchronous Programming Language \CEU}
+
+\begin{comment}
+- structured programming
+- lexical scope
 
@@
 In comparison to promises/futures, it provides lexical parallel constructs, allowing the branches to share local variables and, more importantly, supporting safe abortion of code (with the "par/or").
+\end{comment}
 
-- application on IoT
-- gains??
-
-- structured programming
-- lexical scope
-- automatic power management: standby as well as enabled/disabled, powered, switched
-
-needs a requires a energy-aware driver
-
-process for ISR bindings and
-A infra-structure programmer writes device drivers for
-Each peripheral requires a driver
-An architecture
-Infra
-Once a driver is written, all applications
-
-- synchronous languages / \CEU
-
-- except drivers, which are also written in Ceu
-
-I have been working in the design of \CEU, a new reactive programming language
-targeting resource-constrained embedded systems, for the past 8
-years~\cite{ceu.sensys11,ceu.tr,ceu.sensys13,ceu.rem13,ceu.phd,ceu.rebls14,
-ceu.mod15,ceu.rebls15,ceu.terra,ceu.media.webmedia16,ceu.tecs17}.
+\CEU is a Esterel-based\cite{ceu.tecs17} reactive programming language
+targeting resource-constrained embedded systems~\cite{ceu.sensys13}.
 %
-\CEU is grounded on the synchronous concurrency model, which has been
+It is grounded on the synchronous concurrency model, which has been
 successfully adopted in the context of hard real-time systems such as avionics
@@
 On the other hand, all reactions to the external world are guaranteed to be
-computed in bounded
-time, ensuring that applications always reach an idle state amenable to standby
-mode.
+computed in bounded time, ensuring that applications always reach an idle state
+amenable to standby mode.
 
-- results
-    - atmega 328p/2560
-    - arm cortex-m0
-    - how much efficiency?
+\subsection{A Motivating Example}
+\label{sec.ceu}
+
+{\linespread{1}
+\begin{figure}[t]
+\begin{minipage}[t]{0.50\linewidth}
+\begin{lstlisting}[xrightmargin=1cm]
+digitalWrite(13, 0);
+int old = 0;
+while (1) {
+    int cur = digitalRead(2);
+    if (cur != old) {
+        digitalWrite(13, cur);
+        old = cur;
+    }
+}
+
+\end{lstlisting}
+\centering\small{[a] Active polling in C}
+\end{minipage}
 %
-We expect that existing energy-unaware applications will benefit from savings
-in the order of 50\% based on IEA's estimates considering current standby
-technologies~\cite{iea.data}, and previous third-party work on transparent energy
-awareness~\cite{wsn.tos.2}.
+\begin{minipage}[t]{0.45\linewidth}
+\begin{lstlisting}[numbers=left]
+input  int PIN_02;
+output int PIN_13;
 
-- limitations
+emit PIN_13(0);
+loop do
+    var int v = await PIN_02;
+    emit PIN_13(v);
+end
+.
+\end{lstlisting}
+\centering\small{[b] Reactive execution in \CEU}
+\end{minipage}
+%\rule{8.4cm}{0.37pt}
+\caption{ Controlling a LED from button state changes.
+\label{lst.basic}
+}
+\end{figure}
+}
 
-- paper structure
+Figure~\ref{lst.basic} shows a basic program that makes a button connected
+to the microcontroller's pin 2 to control a LED connected to pin 13.
+Whenever the state of the button changes (i.e., pressed or unpressed), the
+state of the LED also changes (i.e., on or off).
 
-In Section~\ref{sec.ceu}, we introduce the synchronous model together with the
-programming language \CEU.
-In Section~\ref{sec.ext}, we present the language extensions that support
-transparent standby.
-In Section~\ref{sec.eval}, we evaluate .
+The implementation in C in Figure~\ref{lst.basic}.a first turns off the LED
+(ln. 1), then sets an initial state for the button (ln. 2), and then reads in a
+loop (ln. 3-9) its current state (ln. 4) as fast as possible.
+If the current state differs from the old state (ln. 5-8), the new state is
+copied to the LED (ln. 6), and is also set as the old state (ln. 7).
+%
+This approach which actively samples the state of an external device (e.g., a
+button) is known as \emph{polling} or \emph{software-driven I/O}.
+%
+On the one hand, this style does not involve dealing with the complexity of
+callbacks (as discussed in Section~\ref{sec.related}) and is widely adopted in
+Arduino, the most popular embedded platform (to be discussed in
+Section~\ref{sec.method.software}).
+On the other hand, polling wastes CPU cycles and prevents the device to enter in
+standby.
 
-\section{The Synchronous Language CEU}
-\label{sec.ceu}
+The implementation in \CEU in Figure~\ref{lst.basic}.b
+uses dedicated vocabulary to handle events from I/O devices without resorting
+to callbacks.
+The pins are first declared to determine their purpose in the program, i.e.,
+\code{input} for the pin connected to the button and  \code{output} for the pin
+connected to the LED (ln 1-2).
+Then, the LED is turned off with an \code{emit} (ln. 4), which indicates an
+output operation.
+Then, the program enters in a loop to react to changes in the button (ln. 5-8).
+The \code{await} primitive accounts for the reactive nature of \CEU.
+The program literally waits for a change in the button (ln. 6) to be able to
+proceed and set the new state of the LED (ln 7).
+%
+This way, when an application reaches an idle state, the language has precise
+information about which events can awake that application.
+
+By design, all lines of execution in \CEU always reach an \code{await} at the
+end of a reaction to an event (otherwise, the application does not
+compile~\cite{ceu.sensys13}).
+%
+This not only allows the application to enter standby mode, but effectively use
+the deepest sleeping level considering all possible awaking events.
+While in standby, only a hardware interrupt associated with the events can
+awake the application, making it sleep for longest possible periods of time.
+%
+%This is still a theoretical assumption that needs to be proven in practice.
 

### 2018-01-30 LCTES/lctes.tex
(>) LCTES

@@
 \usepackage{booktabs} % For formal tables
+\usepackage{xspace}
+
+\usepackage{listings}
+\usepackage{color}
+    \definecolor{light}{gray}{0.97}
+    \definecolor{dark}{gray}{0.30}
+\lstset{
+%columns=fullflexible,
+%basicstyle=\ttfamily,
+escapeinside={||},
+    %mathescape=true,
+    language=C, % choose the language of the code
+    basicstyle=\fontfamily{pcr}\selectfont\scriptsize\color{black},
+    keywordstyle=\color{black}\bfseries, % style for keywords
+    numbers=none, % where to put the line-numbers
+    numberstyle=\tiny, % the size of the fonts that are used for the line-numbers
+    backgroundcolor=\color{light},
+    showspaces=false, % show spaces adding particular underscores
+    showstringspaces=false, % underline spaces within strings
+    showtabs=false, % show tabs within strings adding particular underscores
+    %frame=single, % adds a frame around the code
+    tabsize=2, % sets default tabsize to 2 spaces
+    %rulesepcolor=\color{gray}
+    captionpos=b, % sets the caption-position to bottom
+    breaklines=false, % sets automatic line breaking
+    %breakatwhitespace=false,
+    numbersep=2em,
+    % C was used in the blocksworld example to refer to block C and nowhere else
+    emph={par,or,hor,do,end,loop,code,await,pause,emit,input,event,call,with,%
+          var,and,then,else,return,pure,deterministic,nohold,finalize,%
+          class, every, FOREVER, this, spawn, in, pool, watching, until, 
+          interface, each, abort, when, signal, PROC, CHAN, SIGNAL, PAR, not,
+          bool, tag, escape, traverse,implementation,output,true,false,
+          native,@const,@pure,@safe,define,public,private,none},
+    emphstyle={\bfseries},
+    commentstyle=\color{dark}\scriptsize,
+    %xleftmargin=20pt,
+    %xrightmargin=20pt,
+    framesep=20pt,
+    %upquote=true,
+    %aboveskip={1.5\baselineskip},
+}
+
+\newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
+\newcommand{\code}[1] {{\small{\texttt{#1}}}}
 
@@
 \begin{document}
-\title{SIG Proceedings Paper in LaTeX Format}
-\titlenote{Produces the permission block, and
-  copyright information}
-\subtitle{Extended Abstract}
-\subtitlenote{The full version of the author's guide is available as
-  \texttt{acmart.pdf} document}
+%\title{A Language-Based Approach for Standby in Embedded Systems}
+\title{Transparent Standby for Low-Power, Resource-Constrained Embedded Systems}
+\subtitle{A Programming Language-Based Approach}
+
+%\titlenote{Produces the permission block, and copyright information}
+%\subtitlenote{The full version of the author's guide is available as \texttt{acmart.pdf} document}
 
@@
 
-\begin{teaserfigure}
+%\begin{teaserfigure}
   %\includegraphics[width=\textwidth]{sampleteaser}
-  \caption{This is a teaser}
-  \label{fig:teaser}
-\end{teaserfigure}
+  %\caption{This is a teaser}
+  %\label{fig:teaser}
+%\end{teaserfigure}
 

### 2018-02-16 LCTES/body.tex
Major review

--- 2018-01-30 LCTES/body.tex
@@
-\section{Introduction}
 \label{sec.introduction}
@@
 According to the International Energy Agency (IEA), the number of
-network-connected devices is expected to reach to 50 billion by 2020 with
-the expansion of the Internet of Things (IoT).~\cite{iea.data}
+network-connected devices is expected to reach 50 billion by 2020 with the
+expansion of the Internet of Things (IoT)~\cite{iea.data}.
 %
@@
 standby consumption, made network standby efficiency one of the six
-pillars of G20's \emph{Energy Efficiency Action Plan}%
+pillars of the G20's \emph{Energy Efficiency Action Plan}%
 \footnote{G20's Energy Efficiency Action Plan: \url{https://www.iea-4e.org/projects/g20}}.
@@
 Our approach lies at the bottom of the software development
-layers---transparent programming language mechanisms---meaning that \emph{all}
-applications take advantage of low-power standby modes automatically, without
-extra programming efforts.
+layers---programming language mechanisms---meaning that \emph{all} applications
+take advantage of low-power standby modes automatically, without extra
+programming efforts.
 %
 We extend the synchronous programming language
-\CEU~\cite{ceu.sensys13,ceu.tecs17} with interrupt service routines (ISRs) and
-a simple power management runtime (PMR).
+\CEU~\cite{ceu.sensys13,ceu.tecs17} with support for interrupt service routines
+(ISRs) and with a simple power management runtime (PMR).
 %
@@
 each peripheral requires a driver in \CEU.
-This is a one-time process and is typically packaged and distributed in a SDK.
+These are a one-time procedures and are typically packaged and distributed in a
+software development kit (SDK).
 %
-All new applications built on top of these drivers will take advantage of
+Then, all new applications built on top of these drivers take advantage of
 standby automatically.
 %
-As a proof of concept, we support the 8-bit \emph{AVR/ATmega} and 32-bit
-\emph{ARM/Cortex-M0} microcontrollers, and a variety of peripherals (e.g., ADC,
-SPI, USART, nRF24L01 transceiver).
-The project is open-source.%
-\footnote{https://github.com/fsantanna/ceu-arduino/}
+As a proof of concept, we provide an open source SDK %
+%\footnote{https://github.com/fsantanna/ceu-arduino/}
+with support for 8-bit
+\emph{AVR/ATmega} and 32-bit \emph{ARM/Cortex-M0} microcontrollers, and a
+variety of peripherals, such as GPIO, ADC, USART, SPI, and the nRF24L01
+transceiver.
 
- Thank you, I will improve the page.
+We developed a number of applications using these peripherals concurrently and
+could verify that the applications remain in deepest standby modes for longest
+periods of time.
+%
+We also compare the structure of programs in \CEU and
+Arduino~\cite{arduino.book}, whose primary goal is to reduce the barrier of
+adoption for non-technical users (e.g., designers and artists).
+We show that we can keep the intended sequential reasoning of Arduino even when
+applications require non-trivial concurrent behaviors.
+
+% - standby as well as enabled/disabled, powered, switched
+% - application on IoT
+% - results
+%   - atmega 328p/2560
+%   - arm cortex-m0
+%   - how much efficiency?
+% - limitations
+
+%In Section~\ref{sec.ceu}, we introduce the structured synchronous model and the
+%programming language \CEU.
+%In Section~\ref{sec.apps}, we evaluate .
+%In Section~\ref{sec.ext}, we present the language extensions that support
+%transparent standby.
+
+\section{The Structured Synchronous Programming Language \CEU}
+
+\begin{comment}
+- structured programming
+- lexical scope
 
@@
 On the other hand, all reactions to the external world are guaranteed to be
-computed in bounded
-time, ensuring that applications always reach an idle state amenable to standby
-mode.
+computed in bounded time, ensuring that applications always reach an idle state
+amenable to standby mode.
+%
+Overall, \CEU aims to offer a concurrent, safe, and expressive alternative to C
+with the characteristics that follow:
+%
+\begin{description}
+\item [Reactive:] code only executes in reactions to events.
+\item [Structured:] programs use structured control mechanisms, such as
+    \code{await} (to suspend a line of execution), and \code{par} (to combine
+    multiple lines of execution).
+\item [Synchronous:] reactions run atomically and to completion on each line of
+    execution, i.e., there's no implicit preemption or real parallelism.
+\end{description}
+%
+Structured reactive programming lets developers write code in direct style,
+recovering from the inversion of control imposed by event-driven
+execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
 
-- results
-    - atmega 328p/2560
-    - arm cortex-m0
-    - how much efficiency?
+\subsection*{A Motivating Example}
+\label{sec.ceu.example}
+
+{\linespread{1}
+\begin{figure}[t]
+\begin{minipage}[t]{0.49\linewidth}
+\begin{lstlisting}[xrightmargin=0.5cm]
+while (1) {
+  delay(1000);
+  int v =
+    analogRead(A0);
+  radioWrite(v);
+}
+\end{lstlisting}
+\centering\small{[a] Version in Arduino}
+\end{minipage}
 %
-We expect that existing energy-unaware applications will benefit from savings
-in the order of 50\% based on IEA's estimates considering current standby
-technologies~\cite{iea.data}, and previous third-party work on transparent energy
-awareness~\cite{wsn.tos.2}.
+\begin{minipage}[t]{0.49\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=-0.2cm]
+loop do
+  await 1s;
+  var int v =
+    await AnalogRead(A0);
+  await RadioWrite(v);
+end
+\end{lstlisting}
+\centering\small{[b] Version in \CEU}
+\end{minipage}
+%\rule{8.4cm}{0.37pt}
+\caption{ Sequence of I/O operations running in a loop.
+\label{lst.direct}
+}
+\end{figure}
+}
 
-- limitations
+Figure~\ref{lst.direct}.a shows a simple, easy-to-read program chunk in Arduino
+that executes forever in a loop a sequence of operations as follows:
+    waits for 1 second (ln. 2),
+    performs an analog to digital conversion (ln. 3--4), and
+    broadcasts the value read (ln. 5).
+%
+Figure~\ref{lst.direct}.b shows the same chunk in \CEU, with a noteworthy
+difference that operations that interact with the environment and take time use
+the \code{await} keyword. % and can be easily identified in the program.
+%
+The traditional structured paradigm encouraged in Arduino (with blocks, loops,
+and sequences) allows for simple and readable code, avoiding the complexity of
+dealing with ISRs.
+%
+However, the use of blocking operations, such as \code{delay(1000)}, prevents
+that other operations execute concurrently.
 
-- paper structure
+Suppose we now want that, at any time, receiving a message via radio should
+immediately abort the loop in Figure~\ref{lst.direct}.a.
+%
+Since the message might arrive concurrently with any of the blocking
+operations, we need to change the structure of the program.
+%
+Figure~\ref{lst.inversion}.a changes the blocking operation \code{delay} to the
+polling operation \code{millis}, which immediately returns the number of
+milliseconds since the reset.
+Now, we start by registering the current time (ln. 1--2) and, on each loop
+iteration, we recheck the time to see if one second has elapsed (ln. 7--9).
+Since these operations are non-blocking, we can intercalate the execution with
+checks for message arrivals (ln. 4--6).
+If the time is up, we start counting it again (ln. 10) before proceeding to the
+original operations in sequence (ln. 11--13).
+%
+The original structured style has been drastically violated to accommodate
+concurrency.
+In the example, we only adapted the \code{delay} operation, but the other
+blocking operations (\code{analogRead} and \code{radioWrite}) would also need
+to be changed to achieve maximum concurrency.
+%
+Alternatively, we could resort to ISRs or implement an event-driven
+scheduler to handle the operations~\cite{tos.nesc}, but ultimately, the
+program readability would still be compromised.
 
-In Section~\ref{sec.ceu}, we introduce the synchronous model together with the
-programming language \CEU.
-In Section~\ref{sec.ext}, we present the language extensions that support
-transparent standby.
-In Section~\ref{sec.eval}, we evaluate .
+The program in Figure~\ref{lst.inversion}.b in \CEU extends the one in
+Figure~\ref{lst.direct}.b to accommodate concurrency.
+%
+The original code remains unmodified (ln. 4--9) and concurrency is achieved
+through the \code{par/or} construct, which creates two lines of execution and
+terminates when either of them terminates, aborting the other automatically.
+%
+This approach preserves the sequential, easy-to-read style while accommodating
+concurrency seamlessly.
 
-\section{The Synchronous Language CEU}
-\label{sec.ceu}
+\begin{figure}[t]
+\begin{minipage}[t]{0.49\linewidth}
+\begin{lstlisting}[xrightmargin=0.5cm]
+uint32_t prv =
+  millis();
+while (1) {
+  if (radioAvail()) {
+    break;
+  }
+  uint32_t cur =
+    millis();
+  if (cur>prv+1000) {
+    prv = cur;
+    int v =
+      analogRead(A0);
+    radioWrite(v);
+  }
+}
+\end{lstlisting}
+\centering\small{[a] Version in Arduino}
+\end{minipage}
+%
+\begin{minipage}[t]{0.49\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=-0.2cm]
+par/or do
+  await RadioAvail();
+with
+  loop do
+    await 1s;
+    var int v =
+      await AnalogRead(A0);
+    await RadioWrite(v);
+  end
+end
+
+.
+\end{lstlisting}
+\centering\small{[b] Version in \CEU}
+\end{minipage}
+%\rule{8.4cm}{0.37pt}
+\caption{ Achieving concurrency between I/O operations.
+\label{lst.inversion}
+}
+\end{figure}
+
+\subsection*{Standby Considerations}
+\label{sec.ceu.standby}
+
+Another advantage of the example in Figure~\ref{lst.inversion}.b is that the
+structure of the program indicates which peripherals are active at a given
+time.
+%
+For instance, when the program is awaiting concurrently in lines 2 and 7,
+only the radio and ADC can awake the program.
+Hence, the language runtime can choose the most energy-efficient sleep mode
+that allows these peripherals to awake the microcontroller from appropriate
+interrupts.
+%
+Since the semantics of \CEU forces the program to always reach \code{await}
+statements in all active lines of execution, it is always possible to put the
+microcontroller into the optimal sleep mode.
+
+This way, when an application reaches an idle state, the language has precise
+information about which events can awake that application.
+
+By design, all lines of execution in \CEU always reach an \code{await} at the
+end of a reaction to an event (otherwise, the application does not
+compile~\cite{ceu.sensys13}).
+%
+This not only allows the application to enter standby mode, but effectively use
+the deepest sleeping level considering all possible awaking events.
+While in standby, only a hardware interrupt associated with the events can
+awake the application, making it sleep for longest possible periods of time.
 
@@
 \label{sec.ceu}
+
+\begin{figure}[t]
+\begin{lstlisting}[numbers=left]
+output (int pin) ANALOG_REQUEST do
+    <...>   // port manipulation to start the conversion
+end
+
+spawn async/isr [CEU_ADC_IRQ_N, 0] do
+    var int value = <...>; // port with the value read
+    emit ANALOG_DONE(value);
+end
+
+code AnalogRead (var int pin) -> int do
+    {ceu_pm_set(CEU_PM_ANALOG, 1);}
+    do finalize with
+        {ceu_pm_set(CEU_PM_ANALOG, 0);}
+    end
+
+    emit ANALOG_REQUEST(pin);
+    var int value = await ANALOG_DONE;
+
+    escape value;
+end
+\end{lstlisting}
+%\rule{8.4cm}{0.37pt}
+\caption{ \CEU driver for the analog-to-digital converter.
+\label{lst.adc}
+}
+\end{figure}
 

### 2018-02-16 LCTES/lctes.tex
Major review

--- 2018-01-29 LCTES/lctes.tex
@@
 \usepackage{booktabs} % For formal tables
+\usepackage{xspace}
+
+\usepackage{listings}
+\lstdefinelanguage{ceu}{%
+  language=C,
+  morekeywords={%
+    and, uint32_t, code,
+    await, bool, call, class, data, define, deterministic, do, each, else,
+    emit, end, escape, event, every, finalize, hor, implementation, in,
+    input, interface, loop, min, native, new, nohold, not, or, output, par,
+    pool, pure, return, signal, spawn, tag, then, this, traverse, until,
+    var, watching, when, with},
+}
+\lstset{
+  basicstyle=\ttfamily,
+  captionpos=b,
+  columns=flexible,
+  commentstyle=\rmfamily\itshape,
+  escapeinside={||},
+  frame=tb,
+  keepspaces=true,
+  keywordstyle=\bfseries,
+  language=ceu,
+  mathescape=true,
+  numbersep=4pt,
+  numberstyle=\scriptsize,
+  upquote=true,
+}
+
+\newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
+\newcommand{\code}[1] {{\small{\texttt{#1}}}}
 
@@
 \begin{abstract}
-This paper provides a sample of a \LaTeX\ document which conforms,
-somewhat loosely, to the formatting guidelines for
-ACM SIG Proceedings.
+Standby efficiency for connected devices is one of the priorities of the G20's
+\emph{Energy Efficiency Action Plan}.
+%
+We propose transparent programming language mechanisms to enforce that
+applications remain in deepest standby modes for longest periods of time.
+%
+We extend the synchronous programming language \CEU with support for interrupt
+service routines and with a simple power management runtime.
+%
+We developed device drivers based on these primitives on top of which
+applications can be built to take advantage of standby automatically.
+%
+We also show that programs in \CEU can keep a sequential structure to lower the
+barrier of adoption, even when applications require non-trivial concurrent
+behaviors.
 \end{abstract}
@@
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{serra,my,standby}
+\bibliography{my,standby,other}
 

### 2018-02-16 LCTES/body.tex
(>) small adjusts

@@
 \emph{AVR/ATmega} and 32-bit \emph{ARM/Cortex-M0} microcontrollers, and a
-variety of peripherals, such as GPIO, ADC, USART, SPI, and the nRF24L01
+variety of peripherals, such as GPIO, A/D converter, USART, SPI, and the nRF24L01
 transceiver.
@@
     waits for 1 second (ln. 2),
-    performs an analog to digital conversion (ln. 3--4), and
+    performs an A/D conversion (ln. 3--4), and
     broadcasts the value read (ln. 5).
@@
 
-Another advantage of the example in Figure~\ref{lst.inversion}.b is that the
-structure of the program indicates which peripherals are active at a given
-time.
+The structure of the program in Figure~\ref{lst.inversion}.b also indicates
+which peripherals are active at a given time.
 %
 For instance, when the program is awaiting concurrently in lines 2 and 7,
-only the radio and ADC can awake the program.
+only the radio transceiver and A/D converter can awake the program.
 Hence, the language runtime can choose the most energy-efficient sleep mode
-that allows these peripherals to awake the microcontroller from appropriate
+that allows these peripherals to awake the microcontroller from associated
 interrupts.
 %
-Since the semantics of \CEU forces the program to always reach \code{await}
+Since the semantics of \CEU enforces the program to always reach \code{await}
 statements in all active lines of execution, it is always possible to put the
-microcontroller into the optimal sleep mode.
-
-This way, when an application reaches an idle state, the language has precise
-information about which events can awake that application.
-
-By design, all lines of execution in \CEU always reach an \code{await} at the
-end of a reaction to an event (otherwise, the application does not
-compile~\cite{ceu.sensys13}).
-%
-This not only allows the application to enter standby mode, but effectively use
-the deepest sleeping level considering all possible awaking events.
-While in standby, only a hardware interrupt associated with the events can
-awake the application, making it sleep for longest possible periods of time.
+microcontroller into the optimal sleep mode after each reaction to the
+environment.
 
@@
 %\rule{8.4cm}{0.37pt}
-\caption{ \CEU driver for the analog-to-digital converter.
+\caption{ \CEU driver for the A/D converter.
 \label{lst.adc}

### 2018-02-16 LCTES/body.tex
(+) PM module

@@
 
-spawn async/isr [CEU_ADC_IRQ_N, 0] do
+async/isr ADC_vect_num do
     var int value = <...>; // port with the value read
@@
 code AnalogRead (var int pin) -> int do
-    {ceu_pm_set(CEU_PM_ANALOG, 1);}
+    {PM_SET(PM_ANALOG, 1);}
     do finalize with
-        {ceu_pm_set(CEU_PM_ANALOG, 0);}
+        {PM_SET(PM_ANALOG, 0);}
     end
@@
 \label{lst.adc}
+}
+\end{figure}
+
+\begin{figure}[t]
+\begin{lstlisting}[numbers=left]
+#define PM_GET(peripheral)       bitRead(pm, peripheral)
+#define PM_SET(peripheral,state) bitWrite(pm, peripheral, state);
+
+static u32 pm = 0;
+
+void pm_sleep (void) {
+  if (PM_GET(PM_TIMER1) || PM_GET(PM_USART) || PM_GET(PM_SPI)) {
+      LowPower.idle(PM_GET(PM_ADC),...)
+    } else if (PM_GET(PM_ADC)) {
+      LowPower.adcNoiseReduction(...);
+    } else {
+      LowPower.powerDown(...);
+    }
+  }
+}
+\end{lstlisting}
+%\rule{8.4cm}{0.37pt}
+\caption{ Power management module for the ATmega328p microcontroller.
+\label{lst.pm}
 }

### 2018-02-16 LCTES/lctes.tex
(+) PM module

@@
   morekeywords={%
-    and, uint32_t, code,
+    and, uint32_t, code, async, isr,
     await, bool, call, class, data, define, deterministic, do, each, else,

### 2018-02-17 LCTES/body.tex
(>) formatting

@@
   int v =
-    analogRead(A0);
+    analogRead();
   radioWrite(v);
@@
   var int v =
-    await AnalogRead(A0);
+    await AnalogRead();
   await RadioWrite(v);
@@
     int v =
-      analogRead(A0);
+      analogRead();
     radioWrite(v);
@@
     var int v =
-      await AnalogRead(A0);
+      await AnalogRead();
     await RadioWrite(v);
@@
 \section{Transparent Standby Mechanisms}
-\label{sec.ceu}
+\label{sec.standby}
+
+In order to empower the example in Figure~\ref{lst.inversion}.b with automatic
+standby, we had to make some modifications and extensions to \CEU:
+%
+\begin{itemize}
+\item Modify the runtime of \CEU to be interrupt driven and to put the
+      microcontroller in standby after each reaction to the environment.
+\item Provide operations for the drivers to indicate which interrupts can awake
+      the program.
+\item Support ISRs in \CEU to generate input events to the program.
+\end{itemize}
+
+Figure~\ref{lst.adc} shows the driver for .
 
@@
 \begin{lstlisting}[numbers=left]
-output (int pin) ANALOG_REQUEST do
+// Exposed driver functionality
+
+output void ANALOG_REQUEST;    // low-level request
+input  int  ANALOG_DONE;       // low-level response
+code AnalogRead (void) -> int; // high-level abstraction
+
+// Driver implementation
+
+output void ANALOG_REQUEST do
   <...>   // port manipulation to start the conversion
@@
 async/isr ADC_vect_num do
-    var int value = <...>; // port with the value read
+  var int value = <...>; // register with the value read
   emit ANALOG_DONE(value);
@@
 
-code AnalogRead (var int pin) -> int do
+code AnalogRead (void) -> int do
   {PM_SET(PM_ANALOG, 1);}
@@
 \begin{lstlisting}[numbers=left]
-#define PM_GET(peripheral)       bitRead(pm, peripheral)
-#define PM_SET(peripheral,state) bitWrite(pm, peripheral, state);
+#define PM_GET(peripheral) \
+        bitRead(pm, peripheral)
+#define PM_SET(peripheral,state) \
+        bitWrite(pm, peripheral, state)
 

### 2018-02-17 LCTES/body.tex
(>) small adjusts

@@
 
-Figure~\ref{lst.adc} shows the driver for .
+Figure~\ref{lst.adc} shows the driver for the A/D converter in \CEU.
+For simplicity, we assume in the paper that the converter has a single channel
+to avoid having to deal with multiplexing.
+%
+The driver exposes raw I/O events (ln. 3--4) that only deal with low-level
+port manipulation in the microcontroller.
+The output event (ln. 9--11)
 
@@
 output void ANALOG_REQUEST do
-  <...>   // port manipulation to start the conversion
+  {
+    ADMUX = 0x40 | (A0 & 0x07);
+    bitSet(ADCSRA, ADIE); // enables interrupt
+    bitSet(ADCSRA, ADSC); // starts the conversion
+  }
 end
@@
 async/isr ADC_vect_num do
-  var int value = <...>; // register with the value read
+  { bitClear(ADCSRA, ADIE); } // disables interrupt
+  var int value = {ADC}; // register with the value read
   emit ANALOG_DONE(value);
@@
 \begin{lstlisting}[numbers=left]
-#define PM_GET(peripheral) \
-        bitRead(pm, peripheral)
-#define PM_SET(peripheral,state) \
-        bitWrite(pm, peripheral, state)
+#define PM_GET(dev)   bitRead(pm,dev)
+#define PM_SET(dev,v) bitWrite(pm,dev,v)
 
@@
 void pm_sleep (void) {
-  if (PM_GET(PM_TIMER1) || PM_GET(PM_USART) || PM_GET(PM_SPI)) {
+  if (PM_GET(PM_TIMER1) ||
+      PM_GET(PM_USART)  ||
+      PM_GET(PM_SPI))
+  {
       LowPower.idle(PM_GET(PM_ADC),...)

### 2018-02-17 LCTES/lctes.tex
(>) small adjusts

@@
   escapeinside={||},
+  escapeinside={|},
   frame=tb,

### 2018-02-17 LCTES/body.tex
(>) standby infrastructure

@@
 Alternatively, we could resort to ISRs or implement an event-driven
-scheduler to handle the operations~\cite{tos.nesc}, but ultimately, the
+scheduler to handle the operations~\cite{wsn.nesc}, but ultimately, the
 program readability would still be compromised.
@@
 
-\section{Transparent Standby Mechanisms}
+\section{Standby Infrastructure}
 \label{sec.standby}
@@
 Figure~\ref{lst.adc} shows the driver for the A/D converter in \CEU.
+This code is specific to the \emph{ATmega328p} microcontroller and must be
+adapted to work in other platforms.
 For simplicity, we assume in the paper that the converter has a single channel
 to avoid having to deal with multiplexing.
-%
+
 The driver exposes raw I/O events (ln. 3--4) that only deal with low-level
 port manipulation in the microcontroller.
-The output event (ln. 9--11)
+Output events are triggered with the \code{emit} keyword (ln. 29), while input
+events are captured with the \code{await} keyword (ln. 30).
+%
+The output event \code{ADC\_REQUEST} (ln. 9--15) enables ADC interrupts and
+starts an analog-to-digital conversion in the single channel \code{A0}.
+Any code in between \code{\{} and \code{\}} is treated as an inline $C$ chunk,
+allowing for easy integration with $C$ for low-level operations.
+
+The \code{async/isr} construct of \CEU defines an ISR which executes
+asynchronously with the program when the specified interrupt occurs.
+Only ISRs can emit input events to the program.
+In the example, we define an ISR to handle \code{ADC} interrupts which fire
+whenever a conversion is complete (ln. 17--21).
+%
+Although the ISR block executes asynchronously on interrupts, the input
+emission (ln. 20) only takes effect on a subsequent reaction, when the
+synchronous part of the program becomes idle.
+%
+This way, race conditions are only possible within \code{async/isr} blocks,
+which are typically hidden inside device drivers.
+\CEU also provides an \code{atomic} primitive to protect critical sections of
+code.
+
+The low-level events are the pieces that vary among platforms.
+A driver can also offer a higher-level portable abstraction to be exposed to
+client code.
+%
+In the example, the \code{AnalogRead} abstraction (ln. 23--33) takes care of
+starting and awaiting the conversion (ln. 29--30), as well as dealing with the
+power management runtime (PMR).
+%
+The \code{PM\_SET(PM\_ADC,1)} (ln. 24) tells the system that, when entering in
+sleep mode, the \code{ADC} must be kept running.
+%
+The \code{PM\_SET(PM\_ADC,0)} inside the \code{finalize} clause (ln. 25--27)
+releases the \code{ADC} subsystem.
+
+The \code{finalize} construct of \CEU executes the nested code whenever the
+enclosing block terminates or is aborted externally.
+%
+The example of Figure~\ref{lst.inversion}.b invokes the \code{AnalogRead}
+abstraction (ln. 7) concurrently with \code{RadioAvail} (ln. 2).
+%
+It is possible that the \code{AnalogRead} terminates normally or that a radio
+message arrives during the A/D conversion, which causes the \code{AnalogRead}
+to abort abruptly.
+In either case, the \code{finalize} clause executes and puts the PMR in a
+consistent state.
+
+Client code for the driver
+
+ chunk,
 
@@
 
-output void ANALOG_REQUEST;    // low-level request
-input  int  ANALOG_DONE;       // low-level response
+output void ADC_REQUEST; // low-level request
+input  int  ADC_DONE;    // low-level response
 code AnalogRead (void) -> int; // high-level abstraction
@@
 
-output void ANALOG_REQUEST do
+output void ADC_REQUEST do
   {
@@
 
-async/isr ADC_vect_num do
+async/isr {ADC_vect_num} do
   { bitClear(ADCSRA, ADIE); } // disables interrupt
-  var int value = {ADC}; // register with the value read
-  emit ANALOG_DONE(value);
+  var int value = {ADC}; // reads register with the value
+  emit ADC_DONE(value);
 end
@@
 code AnalogRead (void) -> int do
-  {PM_SET(PM_ANALOG, 1);}
+  {PM_SET(PM_ADC, 1);}
   do finalize with
-      {PM_SET(PM_ANALOG, 0);}
+      {PM_SET(PM_ADC, 0);}
   end
 
-  emit ANALOG_REQUEST(pin);
-  var int value = await ANALOG_DONE;
+  emit ADC_REQUEST(pin);
+  var int value = await ADC_DONE;
 
@@
 %\rule{8.4cm}{0.37pt}
-\caption{ Power management module for the ATmega328p microcontroller.
+\caption{ Power management module for the \emph{ATmega328p} microcontroller.
 \label{lst.pm}

### 2018-02-17 LCTES/lctes.tex
(>) add WIP tag
(>) ACM conference

@@
 %Conference
-\acmConference[WOODSTOCK'97]{ACM Woodstock conference}{July 1997}{El
-  Paso, Texas USA}
-\acmYear{1997}
-\copyrightyear{2016}
+%\acmConference[WOODSTOCK'97]{ACM Woodstock conference}{July 1997}{El Paso, Texas USA}
+\acmConference[LCTES'18]{ACM SIGPLAN/SIGBED}{June 2018}{Philadelphia, Pennsylvania, USA}
+%\acmYear{1997}
+%\copyrightyear{2016}
 
-\acmPrice{15.00}
+%\acmPrice{15.00}
 
@@
 %\title{A Language-Based Approach for Standby in Embedded Systems}
-\title{Transparent Standby for Low-Power, Resource-Constrained Embedded Systems}
+\title{WIP: Transparent Standby for Low-Power, Resource-Constrained Embedded Systems}
 \subtitle{A Programming Language-Based Approach}

### 2018-02-17 LCTES/body.tex
(+) Conclusion

@@
 (ISRs) and with a simple power management runtime (PMR).
-%
+
 Each supported microcontroller requires bindings in C for the ISRs and PMR, and
 each peripheral requires a driver in \CEU.
-These are a one-time procedures and are typically packaged and distributed in a
+These are write-once code that are typically packaged and distributed in a
 software development kit (SDK).
@@
 transceiver.
+%
+We developed a number of simple applications using these peripherals
+concurrently and could verify that the applications remain in deepest standby
+modes for longest periods of time.
 
-We developed a number of applications using these peripherals concurrently and
-could verify that the applications remain in deepest standby modes for longest
-periods of time.
-%
-We also compare the structure of programs in \CEU and
+In Section~\ref{sec.ceu}, we compare the structure of programs in \CEU and
 Arduino~\cite{arduino.book}, whose primary goal is to reduce the barrier of
@@
 applications require non-trivial concurrent behaviors.
+%
+In Section~\ref{sec.standby}, we discuss the software infrastructure that
+allows for unmmodified programs in \CEU to take advantage of standby
+automatically.
+%
+In Section~\ref{sec.conclusion}, we discuss future work and conclude the paper.
 
@@
 \section{The Structured Synchronous Programming Language \CEU}
+\label{sec.ceu}
 
@@
 
-Client code for the driver
-
- chunk,
-
 \begin{figure}[t]
@@
 %\rule{8.4cm}{0.37pt}
-\caption{ \CEU driver for the A/D converter.
+\caption{ \CEU driver for the \emph{ATmega328p} A/D converter.
 \label{lst.adc}
@@
 \end{figure}
+
+The PMR also expects a platform-specific power management module to be able to
+put the microcontroller into the most efficient sleep mode possible.
+%
+Figure~\ref{lst.pm} implements the \code{pm\_sleep} function for the
+\emph{ATmega328p} microcontroller which the PMR calls when the program is idle.
+%
+Each device has an associated index (ln. 6--10) in the \code{pm} bit vector
+(ln. 4).
+%
+The driver manipulates its device's index to indicate its state
+(Figure~\ref{lst.adc}, ln. 24,26).
+%
+The \code{pm\_sleep} queries the vector to choose the appropriate sleep mode.
+In the example, if the timer is active (ln. 13), the microcontroller can only
+use the least efficient mode%
+\footnote{
+    We use an external library for the sleep modes:
+    \url{http://www.rocketscream.com/blog/2011/07/04/lightweight-low-power-arduino-library/}
+}
+(ln. 14).
+%
+In the best case, e.g., if only external interrupts are required, the
+microcontroller can use the most efficient mode (ln. 18).
+
+With all the standby infrastructure set, the program in
+Figure~\ref{lst.inversion}.b will automatically take advantage of the longest
+periods of time and deepest sleep modes possible.
 
@@
 
+enum {
+  CEU_PM_ADC = 0,
+  CEU_PM_TIMER1,
+  <...>,
+};
+
 void pm_sleep (void) {
-  if (PM_GET(PM_TIMER1) ||
-      PM_GET(PM_USART)  ||
-      PM_GET(PM_SPI))
-  {
-      LowPower.idle(PM_GET(PM_ADC),...)
+  if (PM_GET(PM_TIMER1) || <...>) {
+      LowPower.idle(PM_GET(PM_ADC),<...>)
     } else if (PM_GET(PM_ADC)) {
-      LowPower.adcNoiseReduction(...);
+      LowPower.adcNoiseReduction(<...>);
     } else {
-      LowPower.powerDown(...);
+      LowPower.powerDown(<...>);
     }
@@
 \end{figure}
+
+\section{Conclusion and Future Work}
+\label{sec.conclusion}
+
+In this work, we address standby efficiency for embedded devices at the level
+of programming languages.
+%
+We propose a software infrastructure for the programming language \CEU that
+encompass a power management runtime and support for interrupt service routines
+in the language.
+%
+Our approach relies on the synchronous semantics of the language which enforces
+that reactions to the environment always reach an idle state amenable to
+standby.
+%
+We then take energy-unaware programs in \CEU and show that, using the proposed
+infrastructure, they can take advantage of the longest periods of time and
+deepest sleep modes possible.
+
+In future work, in order to evaluate the gains in energy efficiency with the
+proposed infrastructure, we will evaluate the consumption of realistic
+applications.
+%
+The Arduino community has an abundance of open-source projects which can be
+rewritten in \CEU take advantage of transparent standby modes.
+%
+In this scenario, we can evaluate the time to rewrite, the resulting program
+structure, and the actual energy consumption efficiency.
 

### 2018-02-17 LCTES/body.tex
(>) complete review

@@
 standby consumption, made network standby efficiency one of the six
-pillars of the G20's \emph{Energy Efficiency Action Plan}%
+pillars of the \emph{G20's Energy Efficiency Action Plan}%
 \footnote{G20's Energy Efficiency Action Plan: \url{https://www.iea-4e.org/projects/g20}}.
@@
 \emph{AVR/ATmega} and 32-bit \emph{ARM/Cortex-M0} microcontrollers, and a
-variety of peripherals, such as GPIO, A/D converter, USART, SPI, and the nRF24L01
-transceiver.
+variety of peripherals, such as for GPIO, A/D converter, USART, SPI, and the
+nRF24L01 transceiver.
 %
 We developed a number of simple applications using these peripherals
-concurrently and could verify that the applications remain in deepest standby
-modes for longest periods of time.
+concurrently and could verify that the applications remain in the deepest
+standby modes for the longest periods of time.
 
@@
 Arduino~\cite{arduino.book}, whose primary goal is to reduce the barrier of
-adoption for non-technical users (e.g., designers and artists).
+adoption for a non-technical audience (e.g., designers and artists).
 We show that we can keep the intended sequential reasoning of Arduino even when
-applications require non-trivial concurrent behaviors.
+applications require non-trivial concurrent behavior.
 %
@@
 
-\CEU is a Esterel-based\cite{ceu.tecs17} reactive programming language
+\CEU is a Esterel-based~\cite{ceu.tecs17} reactive programming language
 targeting resource-constrained embedded systems~\cite{ceu.sensys13}.
@@
 %
-On the other hand, all reactions to the external world are guaranteed to be
-computed in bounded time, ensuring that applications always reach an idle state
-amenable to standby mode.
+On the other hand, all reactions to the external environment are guaranteed to
+be computed in bounded time, ensuring that applications always reach an idle
+state amenable to standby mode.
 %
@@
 
-Figure~\ref{lst.direct}.a shows a simple, easy-to-read program chunk in Arduino
-that executes forever in a loop a sequence of operations as follows:
+Figure~\ref{lst.direct}.a shows a straightforward, easy-to-read program chunk
+in Arduino that executes forever in a loop a sequence of operations as follows:
     waits for 1 second (ln. 2),
     performs an A/D conversion (ln. 3--4), and
-    broadcasts the value read (ln. 5).
+    broadcasts the read value (ln. 5).
 %
-Figure~\ref{lst.direct}.b shows the same chunk in \CEU, with a noteworthy
+Figure~\ref{lst.direct}.b shows the same chunk in \CEU, with the noteworthy
 difference that operations that interact with the environment and take time use
@@
 %
-However, the use of blocking operations, such as \code{delay(1000)}, prevents
-that other operations execute concurrently.
+However, the use of blocking operations, such as \code{delay(1000)} (ln. 2),
+prevents that other operations execute concurrently.
 
-Suppose we now want that, at any time, receiving a message via radio should
-immediately abort the loop in Figure~\ref{lst.direct}.a.
+Suppose that we now want to immediately abort the loop in
+Figure~\ref{lst.direct}.a at any time, as soon as a radio message arrives.
 %
@@
 %
-Figure~\ref{lst.inversion}.a changes the blocking operation \code{delay} to the
-polling operation \code{millis}, which immediately returns the number of
-milliseconds since the reset.
+Figure~\ref{lst.inversion}.a changes the blocking \code{delay} to the polling
+\code{millis}, which immediately returns the number of milliseconds since the
+reset.
 Now, we start by registering the current time (ln. 1--2) and, on each loop
 iteration, we recheck the time to see if one second has elapsed (ln. 7--9).
-Since these operations are non-blocking, we can intercalate the execution with
-checks for message arrivals (ln. 4--6).
+Since these operations are non-blocking, we can intercalate their execution
+with checks for message arrivals (ln. 4--6).
 If the time is up, we start counting it again (ln. 10) before proceeding to the
@@
 %
-This approach preserves the sequential, easy-to-read style while accommodating
+This approach preserves the sequential, easy-to-read style while introducing
 concurrency seamlessly.
@@
 In order to empower the example in Figure~\ref{lst.inversion}.b with automatic
-standby, we had to make some modifications and extensions to \CEU:
+standby, we developed some extensions to \CEU:
 %
 \begin{itemize}
-\item Modify the runtime of \CEU to be interrupt driven and to put the
+\item Make the runtime of \CEU to be interrupt driven and to put the
       microcontroller in standby after each reaction to the environment.
@@
 The output event \code{ADC\_REQUEST} (ln. 9--15) enables ADC interrupts and
-starts an analog-to-digital conversion in the single channel \code{A0}.
+starts an analog-to-digital conversion asynchronously in the preipheral for
+the single channel \code{A0}.
 Any code in between \code{\{} and \code{\}} is treated as an inline $C$ chunk,
@@
 %
-Although the ISR block executes asynchronously on interrupts, the input
-emission (ln. 20) only takes effect on a subsequent reaction, when the
-synchronous part of the program becomes idle.
+Although the ISR body executes asynchronously on interrupts, the input emission
+(ln. 20) only takes effect on a subsequent reaction, when the synchronous part
+of the program is idle.
 %
-This way, race conditions are only possible within \code{async/isr} blocks,
-which are typically hidden inside device drivers.
+This way, race conditions are only possible with \code{async/isr} blocks, which
+are typically hidden inside device drivers.
 \CEU also provides an \code{atomic} primitive to protect critical sections of
@@
 The low-level events are the pieces that vary among platforms.
-A driver can also offer a higher-level portable abstraction to be exposed to
-client code.
+A driver can also expose a higher-level portable abstraction to client code.
 %
@@
 The \code{PM\_SET(PM\_ADC,0)} inside the \code{finalize} clause (ln. 25--27)
-releases the \code{ADC} subsystem.
+releases the \code{ADC} subsystem from the PMR.
 
-The \code{finalize} construct of \CEU executes the nested code whenever the
+The \code{finalize} construct of \CEU executes the nested code whenever its
 enclosing block terminates or is aborted externally.
@@
 
-  emit ADC_REQUEST(pin);
+  emit ADC_REQUEST;
   var int value = await ADC_DONE;
@@
 
-With all the standby infrastructure set, the program in
-Figure~\ref{lst.inversion}.b will automatically take advantage of the longest
-periods of time and deepest sleep modes possible.
+With all the standby infrastructure set, the unmodified program of
+Figure~\ref{lst.inversion}.b will automatically take advantage of the deepest
+sleep modes for the longest periods of time possible.
 
@@
 The Arduino community has an abundance of open-source projects which can be
-rewritten in \CEU take advantage of transparent standby modes.
+rewritten in \CEU to take advantage of transparent standby.
 %

### 2018-02-17 LCTES/lctes.tex
(>) complete review

@@
 \begin{abstract}
-Standby efficiency for connected devices is one of the priorities of the G20's
-\emph{Energy Efficiency Action Plan}.
+Standby efficiency for connected devices is one of the priorities of the
+\emph{G20's Energy Efficiency Action Plan}.
 %
 We propose transparent programming language mechanisms to enforce that
-applications remain in deepest standby modes for longest periods of time.
+applications remain in the deepest standby modes for the longest periods of
+time.
 %
@@
 %
-We developed device drivers based on these primitives on top of which
+We provide device drivers based on these primitives on top of which
 applications can be built to take advantage of standby automatically.
@@
 barrier of adoption, even when applications require non-trivial concurrent
-behaviors.
+behavior.
 \end{abstract}

### 2018-02-20 LCTES/lctes.tex
Ajustes no header.

@@
 %\acmConference[WOODSTOCK'97]{ACM Woodstock conference}{July 1997}{El Paso, Texas USA}
-\acmConference[LCTES'18]{ACM SIGPLAN/SIGBED}{June 2018}{Philadelphia, Pennsylvania, USA}
+\acmConference[LCTES'18]{ACM SIGPLAN/SIGBED}{June 2018}{Philadelphia, USA}
 %\acmYear{1997}
@@
     <concept>
-  <concept_id>10010520.10010553.10010562</concept_id>
-  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
+        <concept_id>10010520.10010553.10010562.10010564</concept_id>
+        <concept_desc>Computer systems organization~Embedded software</concept_desc>
     <concept_significance>500</concept_significance>
@@
     <concept>
-  <concept_id>10010520.10010575.10010755</concept_id>
-  <concept_desc>Computer systems organization~Redundancy</concept_desc>
+        <concept_id>10011007.10011006.10011041.10011048</concept_id>
+        <concept_desc>Software and its engineering~Runtime environments</concept_desc>
         <concept_significance>300</concept_significance>
- </concept>
- <concept>
-  <concept_id>10010520.10010553.10010554</concept_id>
-  <concept_desc>Computer systems organization~Robotics</concept_desc>
-  <concept_significance>100</concept_significance>
- </concept>
- <concept>
-  <concept_id>10003033.10003083.10003095</concept_id>
-  <concept_desc>Networks~Network reliability</concept_desc>
-  <concept_significance>100</concept_significance>
     </concept>
@@
 
-\ccsdesc[500]{Computer systems organization~Embedded systems}
-\ccsdesc[300]{Computer systems organization~Redundancy}
-\ccsdesc{Computer systems organization~Robotics}
-\ccsdesc[100]{Networks~Network reliability}
+\ccsdesc[500]{Computer systems organization~Embedded software}
+\ccsdesc[300]{Software and its engineering~Runtime environments}
 
-\keywords{ACM proceedings, \LaTeX, text tagging}
+\keywords{Arduino, Concurrency, Embedded Systems, Esterel, IoT, Standby}
 

### 2018-02-20 LCTES/body.tex
Correções pontuais da Noemi (e algumas minhas).

@@
 %
-For instance, standby power currently accounts for 10--15\% of residential
-electricity consumption, and $CO_2$ emissions related to standby are equivalent
-to those of 1 million cars~\cite{iea.data,standby.australia}.
+For instance, standby power currently accounts for approximately 10--15\% of
+residential electricity consumption, and $CO_2$ emissions related to standby
+are equivalent to those of 1 million cars~\cite{iea.data,standby.australia}.
 %
@@
 \begin{description}
-\item [Reactive:] code only executes in reactions to events.
+\item [Reactive:] code only executes in reactions to events and is idle most of
+    the time.
 \item [Structured:] programs use structured control mechanisms, such as
@@
 \begin{itemize}
-\item Make the runtime of \CEU to be interrupt driven and to put the
-      microcontroller in standby after each reaction to the environment.
-\item Provide operations for the drivers to indicate which interrupts can awake
-      the program.
-\item Support ISRs in \CEU to generate input events to the program.
+\item We made the runtime of \CEU interrupt driven and put the microcontroller
+      in standby after each reaction to the environment.
+\item We provided operations for the drivers to indicate which interrupts can
+      awake the program.
+\item We included support for ISRs in \CEU to generate input events to the
+      program.
 \end{itemize}
@@
 The output event \code{ADC\_REQUEST} (ln. 9--15) enables ADC interrupts and
-starts an analog-to-digital conversion asynchronously in the preipheral for
+starts an analog-to-digital conversion asynchronously in the peripheral for
 the single channel \code{A0}.
-Any code in between \code{\{} and \code{\}} is treated as an inline $C$ chunk,
-allowing for easy integration with $C$ for low-level operations.
+In \CEU, any code in between \code{\{} and \code{\}} is treated as an inline
+$C$ chunk, allowing for easy integration with $C$ for low-level operations.
 
@@
 %
-It is possible that the \code{AnalogRead} terminates normally or that a radio
-message arrives during the A/D conversion, which causes the \code{AnalogRead}
-to abort abruptly.
+The \code{AnalogRead} may terminate normally or a radio message may arrive
+during the A/D conversion, causing the \code{AnalogRead} to abort abruptly.
 In either case, the \code{finalize} clause executes and puts the PMR in a
@@
 
-static u32 pm = 0;
+static u32 pm = 0; // up to 32 peripherals
 
@@
 We propose a software infrastructure for the programming language \CEU that
-encompass a power management runtime and support for interrupt service routines
-in the language.
+encompasses a power management runtime and support for interrupt service
+routines in the language.
 %
@@
 %
-We then take energy-unaware programs in \CEU and show that, using the proposed
-infrastructure, they can take advantage of the longest periods of time and
-deepest sleep modes possible.
+This way, application written in \CEU can take advantage of the longest periods
+of time and deepest sleep modes possible without extra programming efforts.
 

### 2018-02-20 LCTES/body.tex
Correções pontuais do Sztajnberg

@@
 Given the projected scale of the IoT and the role of low-power standby towards
-energy efficiency, this paper has the following goals:
+energy efficiency, our research has the following goals:
 
@@
 layers---programming language mechanisms---meaning that \emph{all} applications
-take advantage of low-power standby modes automatically, without extra
+should take advantage of low-power standby modes automatically, without extra
 programming efforts.
@@
 
-Figure~\ref{lst.direct}.a shows a straightforward, easy-to-read program chunk
+Figure~\ref{lst.direct}.a shows a straightforward, easy-to-read code snippet
 in Arduino that executes forever in a loop a sequence of operations as follows:
@@
 %
-Figure~\ref{lst.direct}.b shows the same chunk in \CEU, with the noteworthy
+Figure~\ref{lst.direct}.b shows the same code in \CEU, with the noteworthy
 difference that operations that interact with the environment and take time use
@@
 %
-The original structured style has been drastically violated to accommodate
-concurrency.
+The original structured style in Figure~\ref{lst.direct}.a has been drastically
+violated to accommodate concurrency in Figure~\ref{lst.inversion}.a.
 In the example, we only adapted the \code{delay} operation, but the other
@@
 %
-The original code remains unmodified (ln. 4--9) and concurrency is achieved
+In contrast with the Arduino version, the original code in \CEU remains
+unmodified (Figure~\ref{lst.inversion}.b, ln. 4--9) and concurrency is achieved
 through the \code{par/or} construct, which creates two lines of execution and
@@
 %
-Figure~\ref{lst.pm} implements the \code{pm\_sleep} function for the
-\emph{ATmega328p} microcontroller which the PMR calls when the program is idle.
+The code in Figure~\ref{lst.pm} implements the \code{pm\_sleep} function for
+the \emph{ATmega328p} microcontroller which the PMR calls when the program is
+idle.
 %

### 2018-02-20 LCTES/lctes.tex
Correções pontuais do Sztajnberg

@@
 %
-We also show that programs in \CEU can keep a sequential structure to lower the
-barrier of adoption, even when applications require non-trivial concurrent
-behavior.
+We also show that programs in \CEU can keep a sequential structure, even when
+applications require non-trivial concurrent behavior, to lower the barrier of
+adoption.
 \end{abstract}

### 2018-02-20 LCTES/body.tex
Sugestões Noemi e Sztajnberg:   - colocar o modelo síncrono como um pré requisito   - incluir os desafios de software relacionados ao standby efetivo

@@
 \footnote{G20's Energy Efficiency Action Plan: \url{https://www.iea-4e.org/projects/g20}}.
+%
+However, making effective use of standby requires software-related efforts in
+order to detect idle periods of activity in the device, identify peripherals
+that must remain functional, and apply appropriate sleep mode levels in the
+microcontroller.
 
-Given the projected scale of the IoT and the role of low-power standby towards
-energy efficiency, our research has the following goals:
-
-\begin{enumerate}
-\item Address energy efficiency through extensive use of standby.
-\item Target low-power, resource-constrained embedded architectures that form
-      the IoT.
-\item Provide standby mechanisms at the programming language level that scale
-      to all applications.
-\item Support transparent/non-intrusive standby mechanisms that reduce barriers
+Given the projected scale of the IoT, the role of low-power standby towards
+energy efficiency, and the posed software-related challenges, our research has
+the following goals:
+%
+%\begin{enumerate}
+(i)   address energy efficiency through meticulous use of standby;
+(ii)  target low-power, resource-constrained embedded architectures that form
+      the IoT;
+(iii) provide standby mechanisms at the programming language level that scale
+      to all applications; and
+(iv)  support transparent/non-intrusive standby mechanisms that reduce barriers
       of adoption.
-\end{enumerate}
+%\end{enumerate}
 
@@
 %
-We extend the synchronous programming language
-\CEU~\cite{ceu.sensys13,ceu.tecs17} with support for interrupt service routines
-(ISRs) and with a simple power management runtime (PMR).
+We extend the programming language \CEU~\cite{ceu.sensys13,ceu.tecs17} with
+support for interrupt service routines (ISRs) and with a simple power
+management runtime (PMR).
+%
+Our approach relies on the synchronous semantics of the language which enforces
+that reactions to the environment always reach an idle state amenable to
+standby.
 
@@
 In Section~\ref{sec.standby}, we discuss the software infrastructure that
-allows for unmmodified programs in \CEU to take advantage of standby
+allows for unmodified programs in \CEU to take advantage of standby
 automatically.

### 2018-02-20 LCTES/lctes.tex
Sugestões Noemi e Sztajnberg:   - colocar o modelo síncrono como um pré requisito   - incluir os desafios de software relacionados ao standby efetivo

--- 2018-02-20 LCTES/lctes.tex
@@
 %
-We extend the synchronous programming language \CEU with support for interrupt
-service routines and with a simple power management runtime.
+We extend the programming language \CEU with support for interrupt service
+routines and with a simple power management runtime.
 %
-We provide device drivers based on these primitives on top of which
+We also provide device drivers based on these primitives on top of which
 applications can be built to take advantage of standby automatically.
 %
-We also show that programs in \CEU can keep a sequential structure to lower the
-barrier of adoption, even when applications require non-trivial concurrent
-behavior.
+Our approach relies on the synchronous semantics of the language which enforces
+that reactions to the environment always reach an idle state amenable to
+standby.
+%
+In addition, we show that programs in \CEU can keep a sequential structure,
+even when applications require non-trivial concurrent behavior, to lower the
+barrier of adoption.
 \end{abstract}

### 2018-02-21 LCTES/body.tex
Revisões Noemi e Alexandre

@@
 %
-Our approach relies on the synchronous semantics of the language which enforces
-that reactions to the environment always reach an idle state amenable to
-standby.
+In contrast with other concurrency models (e.g., thread and actor based), the
+the synchronous semantics of \CEU guarantees that reactions to the environment
+always reach an idle state amenable to standby.
 
@@
 %
-Our approach relies on the synchronous semantics of the language which enforces
-that reactions to the environment always reach an idle state amenable to
-standby.
+Our approach relies on the synchronous semantics of the language which
+guarantees that reactions to the environment always reach an idle state
+amenable to standby.
 %

### 2018-02-21 LCTES/lctes.tex
Revisões Noemi e Alexandre

@@
 %
-We also provide device drivers based on these primitives on top of which
-applications can be built to take advantage of standby automatically.
+Based on these primitives, we also provide device drivers that allow
+applications to take advantage of standby automatically.
 %
-Our approach relies on the synchronous semantics of the language which enforces
-that reactions to the environment always reach an idle state amenable to
-standby.
+Our approach relies on the synchronous semantics of the language which
+guarantees that reactions to the environment always reach an idle state
+amenable to standby.
 %

### 2018-02-23 LCTES/body.tex
(>) Revisão LCTES.

@@
 %\begin{enumerate}
-(i)   address energy efficiency through meticulous use of standby;
+(i)   address energy efficiency through rigorous use of standby;
 (ii)  target low-power, resource-constrained embedded architectures that form
@@
 
-Our approach lies at the bottom of the software development
+Our proposal lies at the bottom of the software development
 layers---programming language mechanisms---meaning that \emph{all} applications
@@
 
-Each supported microcontroller requires bindings in C for the ISRs and PMR, and
-each peripheral requires a driver in \CEU.
-These are write-once code that are typically packaged and distributed in a
+In our approach, each supported microcontroller requires bindings in C for the
+ISRs and PMR, and each peripheral requires a driver in \CEU.
+These are write-once code that is typically packaged and distributed in a
 software development kit (SDK).
@@
 
-Suppose that we now want to immediately abort the loop in
-Figure~\ref{lst.direct}.a at any time, as soon as a radio message arrives.
-%
-Since the message might arrive concurrently with any of the blocking
-operations, we need to change the structure of the program.
-%
-Figure~\ref{lst.inversion}.a changes the blocking \code{delay} to the polling
-\code{millis}, which immediately returns the number of milliseconds since the
-reset.
-Now, we start by registering the current time (ln. 1--2) and, on each loop
-iteration, we recheck the time to see if one second has elapsed (ln. 7--9).
-Since these operations are non-blocking, we can intercalate their execution
-with checks for message arrivals (ln. 4--6).
-If the time is up, we start counting it again (ln. 10) before proceeding to the
-original operations in sequence (ln. 11--13).
-%
-The original structured style in Figure~\ref{lst.direct}.a has been drastically
-violated to accommodate concurrency in Figure~\ref{lst.inversion}.a.
-In the example, we only adapted the \code{delay} operation, but the other
-blocking operations (\code{analogRead} and \code{radioWrite}) would also need
-to be changed to achieve maximum concurrency.
-%
-Alternatively, we could resort to ISRs or implement an event-driven
-scheduler to handle the operations~\cite{wsn.nesc}, but ultimately, the
-program readability would still be compromised.
-
-The program in Figure~\ref{lst.inversion}.b in \CEU extends the one in
-Figure~\ref{lst.direct}.b to accommodate concurrency.
-%
-In contrast with the Arduino version, the original code in \CEU remains
-unmodified (Figure~\ref{lst.inversion}.b, ln. 4--9) and concurrency is achieved
-through the \code{par/or} construct, which creates two lines of execution and
-terminates when either of them terminates, aborting the other automatically.
-%
-This approach preserves the sequential, easy-to-read style while introducing
-concurrency seamlessly.
-
 \begin{figure}[t]
@@
 
+Suppose that we now want to immediately abort the loop in
+Figure~\ref{lst.direct}.a at any time, as soon as a radio message arrives.
+%
+Since the message might arrive concurrently with any of the blocking
+operations, we need to change the structure of the program.
+%
+Figure~\ref{lst.inversion}.a changes the blocking \code{delay} to the polling
+\code{millis}, which immediately returns the number of milliseconds since the
+reset.
+Now, we start by registering the current time (ln. 1--2) and, on each loop
+iteration, we recheck the time to see if one second has elapsed (ln. 7--9).
+Since these operations are non-blocking, we can intercalate their execution
+with checks for message arrivals (ln. 4--6).
+If the time is up, we start counting it again (ln. 10) before proceeding to the
+original operations in sequence (ln. 11--13).
+%
+The original structured style in Figure~\ref{lst.direct}.a has been drastically
+violated to accommodate concurrency in Figure~\ref{lst.inversion}.a.
+Furthermore, we only adapted the \code{delay} operation, but the other blocking
+operations (\code{analogRead} and \code{radioWrite}) would also need to be
+changed to achieve maximum concurrency.
+%
+Alternatively, we could resort to ISRs or implement an event-driven
+scheduler to handle the operations~\cite{wsn.nesc}, but ultimately, the
+program readability would still be compromised.
+
+The program in Figure~\ref{lst.inversion}.b in \CEU extends the one in
+Figure~\ref{lst.direct}.b to accommodate concurrency.
+%
+In contrast with the Arduino version, the original code in \CEU remains
+unmodified (Figure~\ref{lst.inversion}.b, ln. 4--9) and concurrency is achieved
+through the \code{par/or} construct, which creates two lines of execution and
+terminates when either of them terminates, aborting the other automatically.
+%
+This approach preserves the sequential, easy-to-read style while introducing
+concurrency seamlessly.
+
 \subsection*{Standby Considerations}
@@
 \item We included support for ISRs in \CEU to generate input events to the
-      program.
+      program and awake the microcontroller.
 \end{itemize}
+
+\begin{figure}[t]
+\begin{lstlisting}[numbers=left]
+// Exposed driver functionality
+
+output void ADC_REQUEST; // low-level request
+input  int  ADC_DONE;    // low-level response
+code AnalogRead (void) -> int; // high-level abstraction
+
+// Driver implementation
+
+output void ADC_REQUEST do
+  {
+    ADMUX = 0x40 | (A0 & 0x07); // selects channel A0
+    bitSet(ADCSRA, ADIE);       // enables interrupt
+    bitSet(ADCSRA, ADSC);       // starts the conversion
+  }
+end
+
+async/isr {ADC_vect_num} do
+  { bitClear(ADCSRA, ADIE); } // disables interrupt
+  var int value = {ADC}; // reads register with the value
+  emit ADC_DONE(value);
+end
+
+code AnalogRead (void) -> int do
+  {PM_SET(PM_ADC, 1);}
+  do finalize with
+      {PM_SET(PM_ADC, 0);}
+  end
+
+  emit ADC_REQUEST;
+  var int value = await ADC_DONE;
+
+  escape value;
+end
+\end{lstlisting}
+%\rule{8.4cm}{0.37pt}
+\caption{ \CEU driver for the \emph{ATmega328p} A/D converter.
+\label{lst.adc}
+}
+\end{figure}
 
@@
 (ln. 20) only takes effect on a subsequent reaction, when the synchronous part
-of the program is idle.
+of the program becomes idle.
 %
@@
 
-\begin{figure}[t]
-\begin{lstlisting}[numbers=left]
-// Exposed driver functionality
-
-output void ADC_REQUEST; // low-level request
-input  int  ADC_DONE;    // low-level response
-code AnalogRead (void) -> int; // high-level abstraction
-
-// Driver implementation
-
-output void ADC_REQUEST do
-  {
-    ADMUX = 0x40 | (A0 & 0x07);
-    bitSet(ADCSRA, ADIE); // enables interrupt
-    bitSet(ADCSRA, ADSC); // starts the conversion
-  }
-end
-
-async/isr {ADC_vect_num} do
-  { bitClear(ADCSRA, ADIE); } // disables interrupt
-  var int value = {ADC}; // reads register with the value
-  emit ADC_DONE(value);
-end
-
-code AnalogRead (void) -> int do
-  {PM_SET(PM_ADC, 1);}
-  do finalize with
-      {PM_SET(PM_ADC, 0);}
-  end
-
-  emit ADC_REQUEST;
-  var int value = await ADC_DONE;
-
-  escape value;
-end
-\end{lstlisting}
-%\rule{8.4cm}{0.37pt}
-\caption{ \CEU driver for the \emph{ATmega328p} A/D converter.
-\label{lst.adc}
-}
-\end{figure}
-
 The PMR also expects a platform-specific power management module to be able to
@@
 The code in Figure~\ref{lst.pm} implements the \code{pm\_sleep} function for
-the \emph{ATmega328p} microcontroller which the PMR calls when the program is
-idle.
+the \emph{ATmega328p} microcontroller which the PMR calls when the program
+becomes idle.
 %

### 2018-02-23 LCTES/lctes.tex
(>) Revisão LCTES.

@@
 %
-In addition, we show that programs in \CEU can keep a sequential structure,
-even when applications require non-trivial concurrent behavior, to lower the
-barrier of adoption.
+In addition, we show that programs in \CEU can keep a sequential syntactic
+structure, even when applications require non-trivial concurrent behavior, to
+lower the barrier of adoption.
 \end{abstract}

### 2018-03-18 LCTES/lctes18-reviews-16-rebuttal.txt
(>) LCTES

We would like to thank the reviewers for the thoughtful comments, questions,
and criticism.
We will accommodate all clarifications and answers in a final version of the
paper.
Each comment is addressed individually as follows.

References:

`[1]`: *Francisco Sant'Anna and others. 2013. Safe System-level Concurrency on
        Resource-Constrained Nodes. In Proceedings of ACM SenSys'13.*
`[2]`: *Adriano Branco and others. 2015. Terra: Flexibility and Safety in
        Wireless Sensor Networks. In ACM Transactions on Sensor Networks.*

Review #16A
===========

> I would appreciate getting a size estimate of the user code, runtime, driver,
> power management module as the cortex-m0 is very small in ROM/RAM.

All code is compiled together into a single binary image, but the runtime is in
the order of a few kilobytes (<5Kb).
Previous work `[1]` shows an overhead of around 10% for ROM and 5% for RAM for
complete applications in comparison to equivalent implementations in C (network
protocols, radio driver, etc.).

> I would also appreciate knowing the amount of power saved with this approach
> on some real life application.
> If so, may we get a bulk estimate of the amount of power saved?

Even though the examples in the paper use standby effectively, they are too
simple to draw any conclusions about power savings in real-world applications.
That said, our approach is optimal in the sense that the language ensures idle
states and the runtime can always figure out the most efficient sleeping mode
possible.
Nonetheless, this is a work in progress and we acknowledge the importance of a
quantitative evaluation in the conclusion of the paper as future work.
The evaluation will have to be quite extensive to cover typical passive/active
patterns in IoT applications.

Review #16B
===========

> It is not clear to me how the timing requirement can be met under this
> framework.

Timing requirements can be addressed in two levels of guarantees.
For soft real-time requirements, synchronous reactions (i.e., code outside
ISRs) are safer and more straightforward to use because they support rich
control abstractions (e.g., `par/or`), prevent race conditions, and are
guaranteed to terminate in bounded time.
However, since reactions run atomically and cannot interrupt each other, a
non-negligible delay in starting times is possible.
For hard real-time requirements, ISRs (i.e., code inside `async/isr`) are
asynchronous and can perform sensitive operations immediately (such as filling
a serial buffer) while synchronous code is active.

> Also, a quantitative analysis on the performance and energy for real
> applications is needed to show this work’s value in practice.

A quantitative analysis for real applications will be done in future work.
Please, see last comment in `Review #16A`.

Review #16C
===========

>    - no empirical evaluation or comparison to existing work
>    ...
>    - no discussion of related work

Empirical evaluation and comparison to existing work will be done in future
work.
Please, see last comment in `Review #16A`.

>    - no evaluation of the overhead imposed by the use of the language 
>   <...>
>   its unclear to me that synchronous semantics have a clear adoption
>   advantage given the cost/overhead of its implementation

The overhead is very little.
Please, see first comment in `Review #16A`.

> - doesn't this place an extra burden on programmers and a level of expertise
>   beyond that novices to control at the program level?
>   <...>
> - Can we not do something similar and simpler via libraries and wrappers
>   using existing low level languages?

The burden to deal with standby is mostly on the device drivers, which *"are
write-once code that is typically packaged and distributed in a software
development kit (SDK)."* `[Introduction]`
By transferring the work from the applications to the language level, the idea
is that novice programmers never have to deal with standby explicitly.
In general-purpose languages, by exposing libraries and wrappers, the
programmers would still have to call them explicitly, which is exactly what our
proposal intends to avoid.

Furthermore, we show in Figure 2 that, when introducing concurrency, the
structure of the program remains sequential and amenable to inference of the
appropriate sleep mode.
In comparison with Arduino, whose main goal is to lower the entry barrier for
embedded development, Céu also preserves the sequential structure for
concurrent applications.

> - the intro over does it a bit in my opinion with the emphasis on the G20
>   action plan.

OK. We will also discuss the motivations from an academic perspective.

Review #16D
===========

> While it is possible to integrate this approach with CEU, <...>, it is not
> clear whether the approach can be extended to regular programming languages.

Céu preserves a sequential programming structure for complex concurrent
applications.
Please, see third comment in `Review #16C`.

> The benefit of this approach in real systems has not been evaluated.

The evaluations of the benefits in real systems will be done in future work.
Please, see last comment in `Review #16A`.

Review #16E
===========

> Cons:
>  - Expressiveness of the DSL in representing real IoT app/driver semantics
>    not demonstrated
> <...> A major hurdle of a practical DSL is its expressiveness, <...>.
> I <...> would strongly encourage the authors to look for some real IoT
> scenarios to validate this design.

Although we have not evaluated the energy efficiency aspects (see last comment
in `Review #16A`), previous work [1,2] evaluates the expressiveness of Céu in
the context of Wireless Sensor Networks and discusses the development of
drivers, network protocols, and full applications in Céu.

### 2018-03-19 LCTES/lctes18-reviews-16-rebuttal.txt
(>) lctes

@@
         Resource-Constrained Nodes. In Proceedings of ACM SenSys'13.*
+
 `[2]`: *Adriano Branco and others. 2015. Terra: Flexibility and Safety in
@@
 non-negligible delay in starting times is possible.
-For hard real-time requirements, ISRs (i.e., code inside `async/isr`) are
-asynchronous and can perform sensitive operations immediately (such as filling
+For more strict requirements, ISRs (i.e., code inside `async/isr`) are
+asynchronous and can perform sensitive operations much faster (such as filling
 a serial buffer) while synchronous code is active.
@@
 >    - no empirical evaluation or comparison to existing work
->    ...
+>
 >    - no discussion of related work
@@
 >    - no evaluation of the overhead imposed by the use of the language 
->   <...>
+>
 >   its unclear to me that synchronous semantics have a clear adoption
@@
 >   beyond that novices to control at the program level?
->   <...>
+>
 > - Can we not do something similar and simpler via libraries and wrappers
@@
 >    not demonstrated
-> <...> A major hurdle of a practical DSL is its expressiveness, <...>.
+>
+> A major hurdle of a practical DSL is its expressiveness, <...>.
 > I <...> would strongly encourage the authors to look for some real IoT
@@
 Although we have not evaluated the energy efficiency aspects (see last comment
-in `Review #16A`), previous work [1,2] evaluates the expressiveness of Céu in
+in `Review #16A`), previous work `[1,2]` evaluates the expressiveness of Céu in
 the context of Wireless Sensor Networks and discusses the development of

### 2018-03-22 Submissao/pt.tex
(+) Versão em PT do projeto.

\documentclass[12pt,english]{amsart}
\usepackage[utf8]{inputenc}
%\usepackage[margin=0.5in]{geometry}
\usepackage[a4paper, margin=2cm]{geometry}
\linespread{1.5}

%\usepackage{fontspec}
%\setmainfont{Times New Roman}
\usepackage{mathptmx}

\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{lscape}
\usepackage{rotating}
\usepackage{epstopdf}

\usepackage{verbatim}
\usepackage{url}
\usepackage{xspace}
%\usepackage{wrapfig}

\usepackage{listings}
\usepackage{color}
    \definecolor{light}{gray}{0.97}
    \definecolor{dark}{gray}{0.30}
\lstset{
%columns=fullflexible,
%basicstyle=\ttfamily,
escapeinside={||},
    %mathescape=true,
    language=C, % choose the language of the code
    basicstyle=\fontfamily{pcr}\selectfont\scriptsize\color{black},
    keywordstyle=\color{black}\bfseries, % style for keywords
    numbers=none, % where to put the line-numbers
    numberstyle=\tiny, % the size of the fonts that are used for the line-numbers
    backgroundcolor=\color{light},
    showspaces=false, % show spaces adding particular underscores
    showstringspaces=false, % underline spaces within strings
    showtabs=false, % show tabs within strings adding particular underscores
    %frame=single, % adds a frame around the code
    tabsize=2, % sets default tabsize to 2 spaces
    %rulesepcolor=\color{gray}
    captionpos=b, % sets the caption-position to bottom
    breaklines=false, % sets automatic line breaking
    %breakatwhitespace=false,
    numbersep=2em,
    % C was used in the blocksworld example to refer to block C and nowhere else
    emph={par,or,hor,do,end,loop,code,await,pause,emit,input,event,call,with,%
          var,and,then,else,return,pure,deterministic,nohold,finalize,%
          class, every, FOREVER, this, spawn, in, pool, watching, until, 
          interface, each, abort, when, signal, PROC, CHAN, SIGNAL, PAR, not,
          bool, tag, escape, traverse,implementation,output,true,false,
          native,@const,@pure,@safe,define,public,private,none},
    emphstyle={\bfseries},
    commentstyle=\color{dark}\scriptsize,
    %xleftmargin=20pt,
    %xrightmargin=20pt,
    framesep=20pt,
    %upquote=true,
    %aboveskip={1.5\baselineskip},
}

%\newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
\newcommand{\code}[1] {{\small{\texttt{#1}}}}

\usepackage{enumitem}
\setlist{nolistsep}

% Auto-Standby
\title{Eficiência Energética para Software IoT em Larga Escala}

%\author{Anonymous Author}

\begin{document}

\date{}
\maketitle

\vspace{-1cm}
\begin{comment}
\abstract{
TODO
}
\end{comment}

%\newpage
\section{Narrativa Profissional}

Trabalho no campo de \emph{Linguagens de Programação} com o foco em 
\emph{Sistemas Reativos de Tempo Real}.
Sistemas reativos interagem continuamente com o mundo externo através de
sensores e atuadores (e.g., botões, LEDs e motores).
Quando essas interações têm que cumprir prazos estritos, o sistema é dito de
\emph{tempo real}.
A maioria das aplicações do dia-a-dia, tais como suítes de escritório e jogos
eletrônicos são reativas mas bastante tolerante a atrasos.
Em contraste, sistemas embarcados de tempo real, tais como aplicações médicas,
sistemas de aviação, e dispositivos de IoT, devem cumprir prazos mais estritos.

No início do meu Mestrado, me juntei ao grupo de desenvolvimento do
\emph{Ginga}, o padrão de software do Sistema Digital da TV
Brasileira~\cite{ncl.abnt}, que também se tornou um padrão
ITU/UN~\cite{ncl.itu}.
Durante esse período, trabalhei numa integração não intrusiva entre as duas
linguagens padronizadas NCL e Lua através de uma interface de programação
reativa.
O resultado desse trabalho foi publicado no \emph{ACM DocEng} de 2009.
Também escrevi o capítulo sobre desenvolvimento em Lua e NCL do livro de
referência do Ginga.

Em 2009, iniciei o programa de Doutorado com o objetivo de projetar uma nova
linguagem reativa desde o início, agora direcionada a sistemas embarcados com
recursos restritos.
Por um lado, essas plataformas são tipicamente seis ordem de magnitude menos
poderosas que computadores tradicionais em termos de memória e processamento.
Por outro lado, aplicações embarcadas necessitam de mais confiabilidade para
operar sem a intervenção humana por longos períodos de tempo.
Além disso, um subconjunto significativo de sistemas embarcados é implantado
em locais sem linhas de energia e devem funcionar com baterias.
Sendo assim, uma linguagem de programação direcionada a sistemas embarcados
restritos deve ser pequena, confiável e eficiente energeticamente.
Ao final de 2012, me tornei bolsista da \emph{SAAB Aerospace} para um período
de seis meses de pesquisa.
Juntei-me a um grupo de pesquisa em uma universidade da Suécia e reescrevemos
drivers e protocolos de rede de qualidade industrial.
Conseguimos uma redução considerável no tamanho do código fonte, com uso
similar de recursos em comparação com C.
Em 2013, publicamos um artigo com esses resultados no \emph{ACM SenSys}.

Após me graduar, contribuí com o projeto \emph{Cidades Inteligentes}, que
envolve 20 universidade e busca construir uma infraestrutura de IoT.
%
Em 2015, em cooperação com um aluno de Doutorado, publicamos um artigo no
\emph{ACM TOSN}, mostrando que podemos reprogramar dispositivos embarcados
remotamente usando menos energia do que o estado da arte.
%
Durante o mesmo período, publiquei um artigo no
\emph{Int'l Conference on Modularity}
propondo uma nova abstração de concorrência que gera programas mais modulares.
%
Também fui convidado a apresenta minha linguagem de pesquisa em um encontro
anual do \emph{Working Group on Language Design} grupo afiliado com a IFIP da
UNESCO.

Atualmente, como professor associado em uma universidade brasileira, continuo
a investigar como o projeto de linguagens pode afetar o desenvolvimento de
aplicações reativas.
Durante os últimos 4 anos, publiquei ou atuei no comitê do
\emph{Int'l Workshop on Reactive Languages}.
Em 2017, publiquei um artigo consolidando o projeto da minha linguagem de
pesquisa no \emph{ACM TECS}.

\section{Divulgação Científica: O Papel do Software IoT na Eficiência Energética}

De acordo com a Agência Internacional de Energia (IEA)~\cite{iea.data},
existiam em torno de 14 bilhões de dispositivos conectados tradicionais em 2013
(e.g., telefones TVs inteligentes).
Esse número deve crescer para 50 bilhões em 2020 com a proliferação de
dispositivos IoT (e.g., lâmpadas inteligentes e tecnologia vestível).
%
Dispositivos IoT e tradicionais já superam o número de pessoas no planeta por
um fator de dois, e o tráfego de dados é esperado crescer a uma taxa
exponencial nos próximos anos.
%
No entanto, a maior parte da energia desses aparelhos é consumida quando eles
estão em \emph{standby} (\emph{modo em espera}).
%
A emissão anual de $CO_2$ relacionada a standby é equivalente a de 1 milhão de
carros.
%
A projeção de crescimento de IoT, juntamente com o efeito surpreendente dos
efeitos de consumo de standby, fizeram com que a eficiência de standby para
dispositivos conectados fosse um dos seis pilares do \emph{Plano de Ação para
Eficiência Energética} do G20%
\footnote{G20's Energy Efficiency Action Plan: \url{https://www.iea-4e.org/projects/g20}}.

Outras organizações também reportaram sobre a importância da economia de
energia em dispositivos conectados.
%
Para o \emph{Internet Engineering Task Force (IETF)}, ``o gerenciamento
energético está se tornando um requisito adicional para redes devido a diversos
fatores que incluem o aumento dos custos de energia, o impacto ecológico para
a operação das redes, e a regulação de energia''~\cite{ietf.eman}.
%
Para o \emph{American Council for an Energy-Efficient Economy (ACEEE)},
``o potencial para a nova eficiência energética permanece enorme, (...) devemos
considerar uma abordagem sistêmica para escalar a eficiência energética.
(...) a eficiência inteligente é adaptativa, antecipatória, e
conectada''~\cite{aceee.1}.

Considerando iniciativas concretas, o grupo de trabalho
\emph{Electronic Devices and Networks} da IEA foca especificamente na questão
do standby em dispositivos conectados%
\footnote{EDNA initiative: \url{https://edna.iea-4e.org/}}.
A iniciativa da ACEEE em \emph{Intelligent Efficiency} promove uma abordagem
sistemática para otimizar o comportamento cooperativo de dispositivos de modo
a buscar ganhos de energia como um todo.
Ambos as abordagens, por dispositivo e sistêmica, envolvem soluções de
software, dado que a economia de energia é uma política dinâmica que depende
das demandas das aplicações e níveis de baterias em um determinados momentos.
%
Também existem padrões de baixo consume de energia para a infraestrutura de IoT
com diferentes demandas de alcance, velocidade e distribuição
física~\cite{iot.energy.2}.
Como exemplo, o \emph{Bluetooth Low-Energy (BLE)} é um substituto para o padrão
Bluetooth clássico e é projetado para baixas velocidades em redes pessoais
(PANs).
O \emph{6LoWPAN} adapta o padrão IPv6 para baixo consumo e em dispositivos de
processamento limitado.
%
Essas tecnologias permitem transmissões mais eficientes, suportam topologias
flexíveis e reduzem o tráfego consideravelmente.
Elas também possibilitam o uso de modos de standby mínimos.
%
No entanto, essas tecnologias exigem o uso de software para controlar os modos
ativos e de standby de maneira a construir uma IoT eficiente em termos de
energia.

%\newpage
\section{Resumo Científico}

O uso efetivo de standby terá papel fundamental na eficiência energética para
os 50 bilhões de dispositivos IoT esperados até 2020~\cite{iea.data}.
%
Este projeto de pesquisa visa endereçar os desafios de software, conforme
determinados pela IEA, em direção a uma IoT eficiente em termos de
energia~\cite{iea.data}:
    garantir que os dispositivos adotem os níveis mais econômicos possíveis de
    standby, e que os dispositivos permaneçam o maior período possível em
    standby.

Tendo em vista a escala projetada para a IoT e o papel do modo standby para a
eficiência energética, este projeto de pesquisa tem os seguintes objetivos:

\begin{enumerate}
    \item Endereçar a eficiência energética com o uso criterioso de standby.
    \item Focar em arquiteturas embarcadas restritas que formam a IoT.
    \item Prover mecanismos de standby no nível de linguagens de programação para
          escalar para todas as aplicações.
    \item Suportar mecanismos de standby transparentes e não intrusivos para
          reduzir as barreiras de adoção.
\end{enumerate}

Muitas linguagens, extensões e sistemas operacionais direcionados à energia
foram propostos recentemente.
%
Algumas propostas ajustam a QoS (qualidade de serviço) para reduzir o consumo~\cite{os.ecosystem,lang.green,lang.enerj,lang.greenweb}.
%
Outras propostas oferecem mecanismos para trocar o comportamento dependendo das
demandas e níveis de bateria da aplicação~\cite{lang.eon,lang.energytypes,lang.gradual,lang.ent}.
%
Nenhuma dessas iniciativas tira vantagem de modos de standby quando ociosas,
mas somente adaptam ou eliminam computações quando em modo ativo (não
satisfazendo o objetivo 1).

Tenho trabalhado no projeto de uma nova linguagem de programação reativa para
sistemas embarcados restritos pelos últimos 8 anos.
%
A linguagem é baseada no modelo de concorrência síncrono, que troca poder por
confiabilidade e possui um modelo de tempo mais simples que cobre a maioria dos
requisitos de aplicações IoT.
%
Nesse modelo, todas as reações ao mundo externo são computadas em tempo finito,
garantindo que as aplicações sempre chegam a um estado ocioso que é suscetível
ao modo standby.
%
Esperamos que aplicações existentes que não sejam cientes sobre o consumo de
energia irão beneficiar-se de economias na ordem de 50\% baseado nas
estimativas da IEA (considerando as tecnologias atuais de
standby~\cite{iea.data}) e também em trabalhos em ciência transparente de
energia~\cite{wsn.tos.2}.

\section{Projeto Científico}

\subsection{Primeiro Ano}
\label{sec.first}

Necessitaremos de um orçamento de R\$70k:
R\$60k para suporte a três estudantes, R\$5k para equipamentos, e
R\$5k para publicações e custos de viagem.

\subsubsection{Infraestrutura de Hardware para IoT}

Usaremos Arduinos~\cite{arduino} como a principal plataforma de hardware para
IoT.
A maioria dos Arduinos é baseado em microcontroladores de baixo consumo da
Atmel, tais como o \emph{ATmega328p}~\cite{arduino.atmega328p}, suportando
seis modos standby que podem reduzir o consumo para níveis baixíssimos.

No ambiente acadêmica, existe bastante pesquisa com o uso de Arduinos no
contexto de IoT~\cite{arduino.infra,arduino.health,arduino.home,arduino.energy}.
A popularidade do Arduino fará com que a nossa pesquisa seja mais acessível e
reproduzível para outros grupos.
Em educação, muitos cursos em universidades usam o Arduino~\cite{arduino.edu.1,arduino.edu.2,arduino.edu.3,arduino.edu.4}.
Nós temos usado o Arduino em um curso de graduação pelos últimos 3 anos, o que
permitirá avaliar os resultados com programadores de sistemas embarcados menos
experientes.
Na comunidade hobista, há uma abundância de software publicamente disponível
que poderemos adaptar para a nossa linguagem e avaliar os ganhos de eficiência.

\subsubsection{Infraestrutura de Software para IoT}
\label{sec.method.software}

A maneira mais típica em Arduino de interagir com o mundo externo é através de
\emph{polling}, que amostra um periférico externo para detectar mudanças de
estado.
Polling gasta ciclos de CPU e previne que o dispositivo entre em modo standby.
No Arduino, mesmo funcionalidades básicas, tais como temporizadores,
conversores A/D, e SPI, usam ciclos de polling que gastam energia em modo
ativo.

De modo a prover standby automático, as aplicações deve ser inteiramente
reativas a eventos.
%
Recentemente, adicionamos suporte a rotinas de interrupção (ISRs) como um
conceito primitivo na nossa linguagem, o que permitirá reconstruir a
infraestrutura de software IoT com ciência ao modo standby desde sua base.
%
Esse processo consistirá primordialmente de reescrever device drivers, que são
os pedaços de software que interagem diretamente com o hardware.
%
Essa abordagem não irá afetar a maneira como as aplicações são escritas em
níveis mais abstratos, que permanecerá similar às aplicações em Arduino.
No entanto, em vez de gastar ciclos da CPU em espera ativa, as aplicações
entrarão no modo de standby mais profundo possível em quanto estiverem ociosas.

\subsubsection{Aplicações IoT}
\label{sec.method.apps}

De modo a avaliar os ganhos de energia com a infraestrutura proposta,
precisaremos avaliar o consumo em aplicações realistas.
%
A comunidade do Arduino tem uma abundância de projetos open-source que podem
ser reescritos na nossa linguagem para tirar proveito do modo de standby
transparente.
%
Então, poderemos comparar as versões originais e reescritas em termos de
consumo para tirar conclusões sobre a efetividade do modo de standby
transparente.
%
Os cenários mais realistas de IoT usam comunicação por rádio extensivamente.
%
Nesse contexto, iremos avaliar desde protocolos ad-hoc simples até protocolos
mais complexos com ciência energética para ver até que extensão nossa proposta
contribuiria efetivamente em economias de energia.
%
No longo prazo, esperamos mostrar o valor para desenvolvedores reescreverem
suas aplicações na nossa linguagem para tirar proveito dos modos de standby
automáticos.
%
Nessa direção, avaliaremos o tempo tomado para reescrever as aplicações e os
ganhos reais de eficiência energética.

\subsubsection{\textbf{Contribuições Esperadas}}

O vocabulário dedicado a eventos oferecido pela linguagem aumenta o nível de
abstração dos programas para um nível mais próximo do domínio de IoT, provendo
mais segurança e expressividade para programadores.
Até onde sabemos, suporte para ISRs no nível de linguagem ainda não foi
tentado anteriormente.

Nossa proposta visa fazer com que todas as aplicações estejam sujeitas a modos
de standby transparentemente.
Sendo parte da infraestrutura de software, somente device drivers necessitarão
de gerenciamento explícito de energia, e todas as aplicações construídas sobre
eles se beneficiarão de eficiência energética automaticamente.

Nossa linguagem é um projeto de 8 anos e tem uma implementação open-source
madura que está disponível publicamente para downloads.
Com a proposta de adaptação ao contexto de eficiência energética para IoT, a
linguagem pode se tornar uma alternativa prática ao Arduino no curto prazo.

\subsection{Os Três Anos Seguintes}

\subsubsection{Ciência de Energia Sistemática}
\label{sec.method.systematic}

Ciência de energia sistemática considera a rede IoT como um todo, com
dispositivos cooperando para maximizar a eficiência de energia globalmente.
%
Nesse contexto, \emph{computação adaptativa}~\cite{adaptive} é a capacidade
da IoT se adaptar dinamicamente ao consumo de energia, baseando-se na demanda
das aplicações e níveis de bateria durante sua execução.
%
Fundamentalmente, o objetivo ainda é o de que cada dispositivo maximize o tempo
ocioso para ser mais suscetível a modos de standby.
Dessa forma, a pesquisa do primeiro ano é um pré-requisito para essa fase.
%
Nós continuaremos a investigar como mecanismos de linguagens podem propiciar
mais eficiência energética, mas agora em um contexto de computação adaptativa.

O método científico para energia sistemática será similar ao adotado para a
abordagem por dispositivo (conforme discutido na Seção~\ref{sec.method.apps}):
pegar protocolos e aplicações IoT cientes de energia e reescrevê-las usando
mecanismos não intrusivos, e usar os mesmo critérios de avaliação (i.e., tempo
de reescrita, estética do código, e consumo de energia).

\subsubsection{Arquiteturas e Aplicações de IoT Complexas}
\label{sec.method.complex}

A IoT também consiste de dispositivos mais tradicionais, tais como roteadores,
servidores, e smartphones.
%
Em 2016, existiam 3.9 bilhões de assinaturas telefônicas globalmente, e esse
número de alcançar 6.8 bilhões em 2022~\cite{ericsson.mobility}.

Smartphones têm restrições similares de consumo de bateria e também podem tirar
proveito das técnicas propostas para sistemas embarcados restritos.
%
De modo a transpor a barreira de dispositivos restritos para smartphones para a
IoT, percorreremos um caminho similar ao apresentado na Seção~\ref{sec.first}:
%
\begin{description}
\item[Infraestrutura de Hardware]
Usaremos o BeagleBone Black~\cite{bbb.manual}, que compartilha objetivos
similares ao do Arduino, provendo uma plataforma barata e aberta que é adequada
a aplicações mais ricas, tais como interfaces gráficas, multimídia, e jogos.
\item[Infraestrutura de Software]
De modo a garantir standby automático para aplicações, toda infraestrutura de
software, principalmente device drivers, terá que ser recriada usando ISRs em
nossa linguagem.
\item[Applicações]
Além de aplicações IoT, as aplicações de smartphone, tais como mensagens
instantâneas e navegação Web, também podem maximizar a eficiência energética
através do modo de standby.
Iremos reescrever desde aplicações simples, tais como um relógio gráfico, até
aplicações em rede mais complexas, tais como um navegador, para avaliar o
consumo de energia.
\end{description}

\subsubsection{\textbf{Contribuições Esperadas}}

Esperamos que a ciência de energia sistemática irá beneficiar-se dos mecanismos
não intrusivos da nossa linguagem, podendo ser implementada em um nível de
abstração maior, resultando em aplicações eficientes que requerem menos esforço
de programação.

A transição de telefones simples para smartphones resultou na degradação do
tempo de vida das baterias.
No entanto, a maior parte do tempo, os smartphones estão ociosos nos nossos
bolsos mas gastando energia.
Esperamos aumentar consideravelmente a autonomia das baterias mantendo toda a
funcionalidade de um smartphone moderno.

\newpage
\bibliographystyle{abbrv}
\bibliography{serra}
\end{document}

### 2018-04-11 LCTES/body.tex
(>) Revisões artigo LCTES

@@
 always reach an idle state amenable to standby.
+%
+Previous work~\cite{ceu.sensys13} evaluates the expressiveness of \CEU in the
+context of Wireless Sensor Networks and discusses the development of drivers,
+network protocols, and full applications in the language.
+In addition, \CEU incurs a small overhead of memory in comparison to C (around
+5-10\%), thus being a suitable alternative for constrained devices.
+The language runtime is in the order of a few kilobytes (less than 5Kb).
 
@@
 standby modes for the longest periods of time.
+
+Even though the examples in the paper use standby effectively, they are too
+simple to draw any conclusions about power savings in real-world applications.
+That said, our approach is optimal in the sense that the language ensures idle
+states and the runtime can always figure out the most efficient sleeping mode
+possible.
+Nonetheless, this is a work in progress and we acknowledge the importance of a
+quantitative evaluation in the conclusion of the paper as future work.
 
@@
 
+\section{Discussion}
+
+Note that the application of Figure~\ref{lst.inversion}.b relies on the driver
+of Figure~\ref{lst.adc} to achieve transparent standby.
+%
+In \CEU, the burden to deal with standby is transferred to the device drivers,
+which are write-once code written by specialists distributed with an SDK.
+By transferring the work from the applications to the language level, novice
+programmers never have to deal with standby explicitly.
+
+In contrast, general-purpose languages typically provide low-power libraries to
+deal with standby.
+However, programmers still have to call these libraries explicitly,
+characterizing a manual and error-prone mechanism.
+
+Furthermore, we show in Figure~\ref{lst.inversion}.b that, when introducing
+concurrency, the structure of the program remains sequential and amenable to
+inference of the appropriate sleep mode.
+In comparison with Arduino, whose main goal is to lower the entry barrier for
+embedded development, \CEU also preserves the sequential structure for
+concurrent applications.
+
 \section{Conclusion and Future Work}
@@
 \begin{acks}
-  The authors would like to thank Dr. Yuhua Li for providing the
-  MATLAB code of the \textit{BEPS} method.
-
-  The authors would also like to thank the anonymous referees for
-  their valuable comments and helpful suggestions. The work is
-  supported by the \grantsponsor{GS501100001809}{National Natural
-    Science Foundation of
-    China}{http://dx.doi.org/10.13039/501100001809} under Grant
-  No.:~\grantnum{GS501100001809}{61273304}
-  and~\grantnum[http://www.nnsf.cn/youngscientists]{GS501100001809}{Young
-    Scientists' Support Program}.
-
+The authors would like to thank Guilherme Simas for the initial explorations
+with the language extensions and development of some of the device drivers.
 \end{acks}

### 2018-04-12 LCTES/body.tex
(>) Revisões artigo LCTES

@@
 On the other hand, all reactions to the external environment are guaranteed to
-be computed in bounded time, ensuring that applications always reach an idle
-state amenable to standby mode.
+be computed in bounded time~\cite{ceu.sensys13}, ensuring that applications
+always reach an idle state amenable to standby mode.
 %
@@
 
-Note that the application of Figure~\ref{lst.inversion}.b relies on the driver
-of Figure~\ref{lst.adc} to achieve transparent standby.
+The application of Figure~\ref{lst.inversion}.b relies solely on the driver of
+Figure~\ref{lst.adc} to achieve transparent standby.
 %
@@
 programmers never have to deal with standby explicitly.
-
+%
 In contrast, general-purpose languages typically provide low-power libraries to
@@
 
-Furthermore, we show in Figure~\ref{lst.inversion}.b that, when introducing
-concurrency, the structure of the program remains sequential and amenable to
-inference of the appropriate sleep mode.
+In Figure~\ref{lst.inversion}.b that, when introducing concurrency, the
+structure of the program remains sequential and amenable to inference of the
+appropriate sleep mode.
 In comparison with Arduino, whose main goal is to lower the entry barrier for
@@
 concurrent applications.
+
+The synchronous model of \CEU provides logical parallelism for proper
+separation of concerns, while avoiding the hassle of explicit synchronization
+primitives (e.g., locks and mutexes).
+%
+Yet, asynchronous interrupts provide real-time responsiveness for
+time-sensitive operations closer to the hardware.
 

### 2018-04-14 LCTES/body.tex
(>) Revisões artigo LCTES

@@
 However, making effective use of standby requires software-related efforts in
-order to detect idle periods of activity in the device, identify peripherals
-that must remain functional, and apply appropriate sleep mode levels in the
+order to detect idle periods of activity in a device, identify peripherals
+that must remain functional, and apply appropriate sleep mode levels in its
 microcontroller.
@@
 In contrast with other concurrency models (e.g., thread and actor based), the
-the synchronous semantics of \CEU guarantees that reactions to the environment
+synchronous semantics of \CEU guarantees that reactions to the environment
 always reach an idle state amenable to standby.
@@
 5-10\%), thus being a suitable alternative for constrained devices.
-The language runtime is in the order of a few kilobytes (less than 5Kb).
+The language runtime is in the order of a few kilobytes only (less than 5Kb).
 
-In our approach, each supported microcontroller requires bindings in C for the
+In our approach, each supported microcontroller requires hooks in C for the
 ISRs and PMR, and each peripheral requires a driver in \CEU.
@@
 
-Even though the examples in the paper use standby effectively, they are too
-simple to draw any conclusions about power savings in real-world applications.
-That said, our approach is optimal in the sense that the language ensures idle
-states and the runtime can always figure out the most efficient sleeping mode
-possible.
-Nonetheless, this is a work in progress and we acknowledge the importance of a
-quantitative evaluation in the conclusion of the paper as future work.
+%Even though the examples in this paper use standby effectively, they are too
+%simple to draw any conclusions about power savings in real-world applications.
+%That said, our approach is optimal in the sense that the language ensures idle
+%states and the runtime can always figure out the most efficient sleeping mode
+%possible.
+%Nonetheless, this is a work in progress and we acknowledge the importance of a
+%quantitative evaluation in the conclusion of the paper as future work.
 
 In Section~\ref{sec.ceu}, we compare the structure of programs in \CEU and
-Arduino~\cite{arduino.book}, whose primary goal is to reduce the barrier of
-adoption for a non-technical audience (e.g., designers and artists).
+Arduino~\cite{arduino.book}, whose primary goal is to reduce the programming
+barrier of adoption for a non-technical audience (e.g., designers and artists).
 We show that we can keep the intended sequential reasoning of Arduino even when
@@
 Since the message might arrive concurrently with any of the blocking
-operations, we need to change the structure of the program.
+operations, we need to modify the structure of the program in Arduino.
 %
-Figure~\ref{lst.inversion}.a changes the blocking \code{delay} to the polling
+Figure~\ref{lst.inversion}.a replaces the blocking \code{delay} to the polling
 \code{millis}, which immediately returns the number of milliseconds since the
@@
 scheduler to handle the operations~\cite{wsn.nesc}, but ultimately, the
-program readability would still be compromised.
+program readability would still be compromised in the same way.
 
@@
 Hence, the language runtime can choose the most energy-efficient sleep mode
-that allows these peripherals to awake the microcontroller from associated
+that allows these two peripherals to awake the microcontroller from associated
 interrupts.
@@
 In order to empower the example in Figure~\ref{lst.inversion}.b with automatic
-standby, we developed some extensions to \CEU:
+standby, we have developed some extensions to \CEU as follows:
 %
@@
       in standby after each reaction to the environment.
-\item We provided operations for the drivers to indicate which interrupts can
+\item We provided operations for the drivers to indicate which interrupts might
       awake the program.
@@
 
-The driver exposes raw I/O events (ln. 3--4) that only deal with low-level
+The driver exposes raw I/O events (ln. 3--4) that will only deal with low-level
 port manipulation in the microcontroller.
@@
 %
-The output event \code{ADC\_REQUEST} (ln. 9--15) enables ADC interrupts and
-starts an analog-to-digital conversion asynchronously in the peripheral for
-the single channel \code{A0}.
+The output event \code{ADC\_REQUEST} actual implementation (ln. 9--15) enables
+ADC interrupts and starts an analog-to-digital conversion asynchronously in the
+peripheral for the single channel \code{A0}.
 In \CEU, any code in between \code{\{} and \code{\}} is treated as an inline
@@
 In \CEU, the burden to deal with standby is transferred to the device drivers,
-which are write-once code written by specialists distributed with an SDK.
-By transferring the work from the applications to the language level, novice
-programmers never have to deal with standby explicitly.
+which are write-once code written by specialists and distributed with an SDK.
+By transferring the work from the applications to the language level, novice or
+domain programmers never have to deal with standby explicitly.
 %
@@
 However, programmers still have to call these libraries explicitly,
-characterizing a manual and error-prone mechanism.
+characterizing a mechanism that is manual and error prone.
 
-In Figure~\ref{lst.inversion}.b that, when introducing concurrency, the
-structure of the program remains sequential and amenable to inference of the
-appropriate sleep mode.
+In Figure~\ref{lst.inversion}.b, when introducing concurrency, the structure of
+the program remains sequential and amenable to inference of the appropriate
+sleep mode.
 In comparison with Arduino, whose main goal is to lower the entry barrier for
@@
 
-The synchronous model of \CEU provides logical parallelism for proper
+The synchronous model of \CEU provides logical parallelism to enable proper
 separation of concerns, while avoiding the hassle of explicit synchronization
@@
 In this scenario, we can evaluate the time to rewrite, the resulting program
-structure, and the actual energy consumption efficiency.
+structure, and the actual energy efficiency.
 

### 2018-04-14 LCTES/lctes.tex
(>) Revisões artigo LCTES

--- 2018-02-21 LCTES/lctes.tex
@@
-\documentclass[sigplan, anonymous, review]{acmart}
+\documentclass[sigplan]{acmart}
 
@@
 
-\author{Ben Trovato}
-\authornote{Dr.~Trovato insisted his name be first.}
-\orcid{1234-5678-9012}
-\affiliation{%
-  \institution{Institute for Clarity in Documentation}
-  \streetaddress{P.O. Box 1212}
-  \city{Dublin}
-  \state{Ohio}
-  \postcode{43017-6221}
-}
-\email{trovato@corporation.com}
-
-\author{G.K.M. Tobin}
-\authornote{The secretary disavows any knowledge of this author's actions.}
-\affiliation{%
-  \institution{Institute for Clarity in Documentation}
-  \streetaddress{P.O. Box 1212}
-  \city{Dublin}
-  \state{Ohio}
-  \postcode{43017-6221}
-}
-\email{webmaster@marysville-ohio.com}
-
-\author{Lars Th{\o}rv{\"a}ld}
-\authornote{This author is the
-  one who did all the really hard work.}
-\affiliation{%
-  \institution{The Th{\o}rv{\"a}ld Group}
-  \streetaddress{1 Th{\o}rv{\"a}ld Circle}
-  \city{Hekla}
-  \country{Iceland}}
-\email{larst@affiliation.org}
-
-\author{Valerie B\'eranger}
-\affiliation{%
-  \institution{Inria Paris-Rocquencourt}
-  \city{Rocquencourt}
-  \country{France}
-}
-\author{Aparna Patel}
-\affiliation{%
- \institution{Rajiv Gandhi University}
- \streetaddress{Rono-Hills}
- \city{Doimukh}
- \state{Arunachal Pradesh}
- \country{India}}
-\author{Huifen Chan}
-\affiliation{%
-  \institution{Tsinghua University}
-  \streetaddress{30 Shuangqing Rd}
-  \city{Haidian Qu}
-  \state{Beijing Shi}
-  \country{China}}
-
-\author{Charles Palmer}
-\affiliation{%
-  \institution{Palmer Research Laboratories}
-  \streetaddress{8600 Datapoint Drive}
-  \city{San Antonio}
-  \state{Texas}
-  \postcode{78229}}
-\email{cpalmer@prl.com}
-
-\author{John Smith}
-\affiliation{\institution{The Th{\o}rv{\"a}ld Group}}
-\email{jsmith@affiliation.org}
-
-\author{Julius P.~Kumquat}
-\affiliation{\institution{The Kumquat Consortium}}
-\email{jpkumquat@consortium.net}
+\author[F.~Sant'Anna]{Francisco Sant'Anna}
+\email{francisco@ime.uerj.br}
+\affiliation{Rio de Janeiro State University, Brazil}
+%
+\author[A.~Sztajnberg]{Alexandre Sztajnberg}
+\email{alexszt@ime.uerj.br}
+\affiliation{Rio de Janeiro State University, Brazil}
+%
+\author[A.L..~de~Moura]{Ana L\'ucia de Moura}
+\email{amoura@inf.puc-rio.br}
+\affiliation{PUC-Rio, Brazil}
+%
+\author[N.~Rodrigues]{Noemi Rodrigues}
+\email{noemi@inf.puc-rio.br}
+\affiliation{PUC-Rio, Brazil}
 
 % The default list of authors is too long for headers.
-\renewcommand{\shortauthors}{B. Trovato et al.}
+%\renewcommand{\shortauthors}{B. Trovato et al.}
 
@@
 %
-In addition, we show that programs in \CEU can keep a sequential structure,
-even when applications require non-trivial concurrent behavior, to lower the
-barrier of adoption.
+In addition, to lower the programming barrier of adoption, we show that
+programs in \CEU can keep a sequential syntactic structure, even when
+applications require non-trivial concurrent behavior.
 \end{abstract}

### 2018-05-01 LCTES/body.tex
(>) LCTES: versão final

@@
 
+% TODO-FULL: voltar c/ artigo completo
+\begin{comment}
 According to the International Energy Agency (IEA), the number of
@@
 the following goals:
+\end{comment}
+%%%%
+According to the International Energy Agency (IEA), the number of
+network-connected devices is expected to reach 50 billion by 2020 with the
+expansion of the Internet of Things (IoT)~\cite{iea.data}.
+%
+Most of the energy to power these devices will be consumed in
+\emph{standby mode}, i.e., when they are neither transmitting or processing
+data.
+%
+However, making effective use of standby requires software-related efforts in
+order to detect idle periods of activity in a device, identify peripherals
+that must remain functional, and apply appropriate sleep mode levels in its
+microcontroller.
+Therefore, our research has the following goals:
+%%%%
 %
@@
 %
-Previous work~\cite{ceu.sensys13} evaluates the expressiveness of \CEU in the
-context of Wireless Sensor Networks and discusses the development of drivers,
-network protocols, and full applications in the language.
-In addition, \CEU incurs a small overhead of memory in comparison to C (around
-5-10\%), thus being a suitable alternative for constrained devices.
+Previous work~\cite{ceu.sensys13} demonstrates the expressiveness of \CEU in
+the context of Wireless Sensor Networks and discusses the development of
+drivers, network protocols, and full applications in the language.
+It also attests a small overhead of memory in comparison to C (around 5-10\%),
+thus being a suitable alternative for constrained devices.
 The language runtime is in the order of a few kilobytes only (less than 5Kb).
@@
 We developed a number of simple applications using these peripherals
-concurrently and could verify that the applications remain in the deepest
-standby modes for the longest periods of time.
+concurrently and could verify that they remain in the deepest standby modes for
+the longest periods of time.
 
@@
 Arduino~\cite{arduino.book}, whose primary goal is to reduce the programming
-barrier of adoption for a non-technical audience (e.g., designers and artists).
+barrier of adoption for a non-technical audience. % TODO-FULL (e.g., designers and artists).
 We show that we can keep the intended sequential reasoning of Arduino even when
@@
 With all the standby infrastructure set, the unmodified program of
-Figure~\ref{lst.inversion}.b will automatically take advantage of the deepest
+Figure~\ref{lst.inversion}.b automatically takes advantage of the deepest
 sleep modes for the longest periods of time possible.
@@
 
-In future work, in order to evaluate the gains in energy efficiency with the
-proposed infrastructure, we will evaluate the consumption of realistic
-applications.
+In future work, we will evaluate the consumption of realistic applications.
+%in order to evaluate the gains in energy efficiency with the
+%proposed infrastructure,
 %
@@
 with the language extensions and development of some of the device drivers.
+This work was supported by the Serrapilheira Institute (grant number
+Serra-1708-15612).
 \end{acks}
+
+\balance

### 2018-05-01 LCTES/lctes.tex
(>) LCTES: versão final

@@
 \documentclass[sigplan]{acmart}
+
+% TODO-FULL
+%\linespread{0.96}
+\setlength{\belowcaptionskip}{-10pt}
 
@@
 \usepackage{xspace}
+\usepackage{balance}
 
@@
 \lstset{
-  basicstyle=\ttfamily,
+  basicstyle=\small\ttfamily,
   captionpos=b,
@@
 %\setcopyright{acmlicensed}
-\setcopyright{rightsretained}
+%\setcopyright{rightsretained}
 %\setcopyright{usgov}
@@
 
-% DOI
-\acmDOI{10.475/123_4}
-
-% ISBN
-\acmISBN{123-4567-24-567/08/06}
-
-%Conference
-%\acmConference[WOODSTOCK'97]{ACM Woodstock conference}{July 1997}{El Paso, Texas USA}
-\acmConference[LCTES'18]{ACM SIGPLAN/SIGBED}{June 2018}{Philadelphia, USA}
-%\acmYear{1997}
-%\copyrightyear{2016}
-
-%\acmPrice{15.00}
-
 %\acmBadgeL[http://ctuning.org/ae/ppopp2016.html]{ae-logo}
 %\acmBadgeR[http://ctuning.org/ae/ppopp2016.html]{ae-logo}
+
+%%% If you see 'ACMUNKNOWN' in the 'setcopyright' statement below,
+%%% please first submit your publishing-rights agreement with ACM (follow link on submission page).
+%%% Then please update our instructions page and copy-and-paste the NEW commands into your article.
+%%% Please contact us in case of questions; allow up to 10 min for the system to propagate the information.
+%%%
+%%% The following is specific to LCTES'18 and the paper
+%%% 'Transparent Standby for Low-Power, Resource-Constrained Embedded Systems: A Programming Language-Based Approach (Short WIP Paper)'
+%%% by Francisco Sant'Anna, Alexandre Sztajnberg, and Noemi Rodrigues.
+%%%
+%\setcopyright{ACMUNKNOWN}
+\setcopyright{acmcopyright}
+\acmPrice{15.00}
+\acmDOI{10.1145/3211332.3211337}
+\acmYear{2018}
+\copyrightyear{2018}
+\acmISBN{978-1-4503-5803-3/18/06}
+\acmConference[LCTES'18]{19th ACM SIGPLAN/SIGBED Conference on Languages, Compilers, and Tools for Embedded Systems}{June 19--20, 2018}{Philadelphia, PA, USA}
 
@@
 %
-In addition, to lower the programming barrier of adoption, we show that
-programs in \CEU can keep a sequential syntactic structure, even when
+In addition, in order to lower the programming barrier of adoption, we show
+that programs in \CEU can keep a sequential syntactic structure, even when
 applications require non-trivial concurrent behavior.

### 2018-05-29 LCTES/lctes.tex
(*) fix previous commit
(>) LCTES

@@
-\documentclass[sigplan]{acmart}
+\documentclass[sigplan,screen]{acmart}
 
@@
 
-%%% If you see 'ACMUNKNOWN' in the 'setcopyright' statement below,
-%%% please first submit your publishing-rights agreement with ACM (follow link on submission page).
-%%% Then please update our instructions page and copy-and-paste the NEW commands into your article.
-%%% Please contact us in case of questions; allow up to 10 min for the system to propagate the information.
-%%%
 %%% The following is specific to LCTES'18 and the paper
@@
 %%%
-%\setcopyright{ACMUNKNOWN}
 \setcopyright{acmcopyright}
@@
 %\title{A Language-Based Approach for Standby in Embedded Systems}
-\title{WIP: Transparent Standby for Low-Power, Resource-Constrained Embedded Systems}
-\subtitle{A Programming Language-Based Approach}
+\title[Transparent Standby for Low-Power, Resource-Constrained ...]{Transparent Standby for Low-Power, Resource-Constrained Embedded Systems}
+\subtitle{A Programming Language-Based Approach (Short WIP Paper)}
 
@@
 \author[F.~Sant'Anna]{Francisco Sant'Anna}
+\affiliation{Rio de Janeiro State University, Brazil}
 \email{francisco@ime.uerj.br}
-\affiliation{Rio de Janeiro State University, Brazil}
 %
 \author[A.~Sztajnberg]{Alexandre Sztajnberg}
+\affiliation{Rio de Janeiro State University, Brazil}
 \email{alexszt@ime.uerj.br}
-\affiliation{Rio de Janeiro State University, Brazil}
 %
-\author[A.L..~de~Moura]{Ana L\'ucia de Moura}
+\author[A.L.~de~Moura]{Ana L\'ucia de Moura}
+\affiliation{PUC-Rio, Brazil}
 \email{amoura@inf.puc-rio.br}
-\affiliation{PUC-Rio, Brazil}
 %
 \author[N.~Rodrigues]{Noemi Rodrigues}
+\affiliation{PUC-Rio, Brazil}
 \email{noemi@inf.puc-rio.br}
-\affiliation{PUC-Rio, Brazil}
 

### 2018-07-12 Submissao/mid.tex
(+) mid evaluation

\documentclass[12pt,english]{amsart}
%\usepackage[margin=0.5in]{geometry}
\usepackage[a4paper, margin=2cm]{geometry}
\linespread{1.5}

%\usepackage{fontspec}
%\setmainfont{Times New Roman}
\usepackage{mathptmx}

\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{lscape}
\usepackage{rotating}
\usepackage{epstopdf}

\usepackage{verbatim}
\usepackage{url}
\usepackage{xspace}

\usepackage{listings}
\usepackage{color}
    \definecolor{light}{gray}{0.97}
    \definecolor{dark}{gray}{0.30}
\lstset{
%columns=fullflexible,
%basicstyle=\ttfamily,
escapeinside={||},
    %mathescape=true,
    language=C, % choose the language of the code
    basicstyle=\fontfamily{pcr}\selectfont\scriptsize\color{black},
    keywordstyle=\color{black}\bfseries, % style for keywords
    numbers=none, % where to put the line-numbers
    numberstyle=\tiny, % the size of the fonts that are used for the line-numbers
    backgroundcolor=\color{light},
    showspaces=false, % show spaces adding particular underscores
    showstringspaces=false, % underline spaces within strings
    showtabs=false, % show tabs within strings adding particular underscores
    %frame=single, % adds a frame around the code
    tabsize=2, % sets default tabsize to 2 spaces
    %rulesepcolor=\color{gray}
    captionpos=b, % sets the caption-position to bottom
    breaklines=false, % sets automatic line breaking
    %breakatwhitespace=false,
    numbersep=2em,
    % C was used in the blocksworld example to refer to block C and nowhere else
    emph={par,or,hor,do,end,loop,code,await,pause,emit,input,event,call,with,%
          var,and,then,else,return,pure,deterministic,nohold,finalize,%
          class, every, FOREVER, this, spawn, in, pool, watching, until, 
          interface, each, abort, when, signal, PROC, CHAN, SIGNAL, PAR, not,
          bool, tag, escape, traverse,implementation,output,true,false,
          native,@const,@pure,@safe,define,public,private,none},
    emphstyle={\bfseries},
    commentstyle=\color{dark}\scriptsize,
    %xleftmargin=20pt,
    %xrightmargin=20pt,
    framesep=20pt,
    %upquote=true,
    %aboveskip={1.5\baselineskip},
}

\newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
\newcommand{\code}[1] {{\small{\texttt{#1}}}}

\usepackage{enumitem}
\setlist{nolistsep}

% Auto-Standby
\title{Energy Efficiency for IoT Software in the Large \\ Progress Report}

\author{Francisco Sant'Anna, UERJ}

\begin{document}

%\date{\today}
\maketitle

\vspace{-0.5cm}
%\newpage

\section{Original Proposal}

Considering the projected scale of the IoT for the next decade and the role of
low-power standby towards energy efficiency~\cite{iea.data}, this research
project has the following goals:

\begin{enumerate}
    \item Address energy efficiency through rigorous use of standby.
    \item Target constrained embedded architectures that form the IoT.
    \item Provide standby mechanisms at the programming language level that
          scale to all applications.
    \item Support transparent/non-intrusive standby mechanisms that reduce
          barriers of adoption.
\end{enumerate}

This proposal lies at the bottom of the software development
layers---transparent programming language mechanisms---meaning that \emph{all}
applications would take advantage of low-power standby modes automatically,
without extra programming efforts.
%
We expect that existing energy-unaware applications will benefit from savings
in the order of 50\%.
%
We will require a budget of R\$70k:
R\$60k to support three students, R\$5k for equipment, and
R\$5k for publication and travel costs.

The bulk of this proposal is grounded on the design of \CEU~\cite{ceu.tecs17},
a new reactive programming language targeting resource-constrained embedded
systems, which I have been working for the past 8 years.
%
The language is grounded on the synchronous concurrency model, which trades
power for reliability and has a simpler model of time that suits most
requirements of IoT applications.
%
In this model, all reactions to the external world are guaranteed to be
computed in bounded time, ensuring that applications always reach an idle state
amenable to standby mode.

\section{Timeline of Main Events}

\begin{description}
\item[Jul/17]
    Early explorations with the core idea of the project.
\item[Sep/17]
    Submission of the proposal to Serrapilheira.
\item[Dec/17]
    Grant award notification from Serrapilheira.
\item[Jan/18]
    Submission of two student scholarships to \emph{Google Summer of Code (GSoC'18)}%
    \footnote{\emph{GSoC'18}: \url{https://summerofcode.withgoogle.com/}}.
\item[Fev/18]
    First fully-working prototype of \CEU with support for interrupts and a
    power manager.
\item[Fev/18]
    Submission to \emph{LCTES'18} %
    \footnote {
        \emph{19th Annual ACM SIGPLAN / SIGBED Conference on Languages,
            Compilers, and Tools for Embedded Systems}:
        \url{https://conf.researchr.org/track/LCTES-2018/LCTES-2018-papers}.
    }
    of a work-in-progress paper entitled \emph{``Transparent Standby for
    Low-Power, Resource-Constrained Embedded Systems: A Programming
    Language-Based Approach''}~\cite{ceu.lctes18.short} and a full paper entitled \emph{``A
    Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous
    Programming Language \CEU''}~\cite{ceu.lctes18}.
\item[Mar/18]
    Admission into the graduate program of Electrical Engineering at UERJ.
    \\
    Invitation to be program chair of \emph{REBLS'18} %
    \footnote {
        \emph{5th ACM SIGPLAN Workshop on Reactive and Event-based Languages \& Systems}:
        \url{https://2018.splashcon.org/track/rebls-2018-papers}.
    }.
\item[Mar-May/18]
    Engineering efforts on the prototype: refactoring, optimizations, more
    drivers, multiple platforms (AVR \& SAMD).
\item[Apr/18]
    Grant agreement signature with Serrapilheira.
    \\
    Grant award notification from Google for the two \emph{GSoC'18} students%
    \footnote{\emph{LabLua@GSoC'18}: \url{https://summerofcode.withgoogle.com/organizations/5641514328260608/}}:
    Anny (from UERJ) will work on a integrated programming environment for \CEU.
    Naveen (from India) will work on new device drivers.
\item[Apr/18]
    First full-application experiments with significant energy savings ranging
    from 20-99\%.
\item[Jun/18]
    Admission of Anna (from PUC-Rio) as an intern to work on new IoT
    applications.
\item[Jun/18]
    Trip to present the two papers on \emph{LCTES'18}.
\end{description}

\section{Progress Report}

Since the grant notification at the end of last year, our main goal was to
reach, as fast as possible, a working prototype of our core research idea.
%
I worked towards a new version of our programming language \CEU that would
provide automatic standby for all applications:
    a programmer writes an IoT application unaware of standby or energy
    constraints and the language puts the device to sleep whenever possible.

In parallel, I worked with colleagues from PUC-Rio on a formal semantics of
\CEU and a proof that all valid programs will react to an external event in
bounded time, with bounded memory, and deterministic behavior.
This proof is an important result because it shows that applications always
reach an idle state amenable to standby.
The formalization of \CEU already appeared on my PhD thesis~\cite{ceu.phd}, but
the proofs are new results from this year.

By the end of February was the submission deadline for \emph{LCTES'18}, the
main annual venue that links programming languages and embedded systems.
Fortunately, we had enough time to submit a full paper with the
proofs~\cite{ceu.lctes18} and a work-in-progress paper describing our approach
for transparent standby at the language level~\cite{ceu.lctes18.short}, both
of which were accepted for publication.
%
Even though we did not have a comprehensive evaluation of energy savings, the
conceptual framework and working prototype raised enough interest in the
community for discussion on the conference.

\begin{figure}
\begin{minipage}{0.35\textwidth}
\includegraphics[width=\linewidth]{results3}
\end{minipage}
\begin{minipage}{0.33\textwidth}
\includegraphics[width=\linewidth]{blink_c}
\end{minipage}
\begin{minipage}{0.28\textwidth}
\includegraphics[width=\linewidth]{blink_ceu}
\end{minipage}
\caption{ Energy consumption for 5 scenarios.
          ``Blink'' in C/Arduino and \CEU.
\label{lst.eval}
}
\end{figure}

At this moment, we have more concrete evaluation scenarios that show
significant reduction in energy consumption due to standby in the
microcontroller and other peripherals.
%
Figure~\ref{lst.eval} shows the energy consumption of five applications
originally in C/Arduino which we rewrote in \CEU.
%
The first case illustrates how a naive ``Empty'' application in C/Arduino
remains 100\% of the time awake while the same application in \CEU sleeps in
the most efficient standby mode.
%
The ``Blink'' example shows that \CEU preserves the sequential, easy-to-read
style of C/Arduino, while consuming much less energy.
This is due to the assumptions \CEU, based on its tractable semantics, can
extract from the program to be sure that it can sleep on the calls to
\code{await}.
%
The ``Radio'' example is another interesting case because the radio driver in
\CEU identifies that the radio peripheral can enter in sleep mode to consume
much less energy (toggling between active and passive radio modes).

Considering the second phase to come and the interactions with Serrapilheira
since the beginning of the first phase and, it became more and more clear that
the grant evaluation criteria will be mostly based on long-term relevance of
the research and that it will encourage more fundamental aspects.
%
In this direction, we chose to invest considerable efforts on the foundations
of the language, highlighting its distinguished semantic aspects that enable
transparent standby.
The paper with the termination proof is the first result in this front, and we
want to continue in this direction by employing automatic verification tools,
such as TODO.
More concretely, we want to automate the manual proofs we currently have and
also answer program specification questions such as "Can we prove that this
application will turn off the radio 10 minutes after receiving a certain
message?"
%
Another education
%
At the beginning of this year I was invited to be the program of
\emph{REBLS'18} which is
The REBLS fundamentals of reactive languages
foundational models of reactive languages

Specification correctness

Dafny is a verification-aware programming language
    - program verifier

LiquidHaskell
 logical predicates that let you enforce critical properties at compile time
    - Avoid Infinite Loops
    - Enforce Correctness Properties
    - Prove Laws by Writing Code

REBLS is about the future of our area

How IoT development will affect the new generation and how our undergraduate
courses are ready to .

- new results
- education: IoT and Cyber physical strong on graduate, not 
    - lack of accessible languages and concurrency programming environments
    - specially energy efficiency

- incluir o standby de componentes na avaliacao
- mostrar o que nao dá pra fazer com TinyOS
    - como seria o caso de par/or em que o finalize remove o periferico da lista de possiveis awakes?
    - e o caso de componentes que são inteiramente desligados
        - caso de um par/or que mata o
            - radio listening
            - o radio inteiramente
        - como desligar dentro de finalize coisas que são desligadas via emit/await?

questions such as, can I prove that this particular program will

My initial budget plan was to spend around R\$5k per month with research staff
(80\% of the budget).
In particular, we wanted a professional embedded system engineer early in the
project full time on the heavy-lifting work on drivers and platforms to serve
as basis for for developing IoT applications.
%
However, since the grant became available only on April, I could not hire such
professional.
%
Nonetheless, I could dedicate most of my time until May on the engineering
aspects of the project, since the core language mechanisms were developed
between December and February.
At the end, the outcome was actually favorable since I overestimated the
importance of an experienced domain-specific engineer at an early stage.
The saved money was redirected to the international trip which was not
initially planned.

Another difficulty that we had to deal was the lack of students to work in the
area of embedded systems, even when offering above-average remuneration.
%
Furthermore, as a new professor in my University I was not associated with any
graduate program until mid March, when I joined the Electrical Engineering
program in the field of \emph{Networks \& Distributed Systems}.
However, we currently only offer Master degrees and most of the students have
part-time commitment with the program.
%
The scholarships from Google for \emph{GSoC'18} raise more interest on students
and we could hire an undergraduate student that is working

part time

(Doctorate degrees

the University currently has no Doctorate program with

On April we were notified about two scholarships

%
My current plan is

Another

porting 
SAMD
From the very beginning,

the concept

describing the

all programs

Céu is a synchronous programming language for embedded soft real-time systems. It focuses on control-flow safety features, such as safe shared-memory concurrency and safe abortion of lines of execution, while enforcing memory bounded, deterministic, and terminating reactions to the environment.
In this work, we present a small-step structural operational semantics for Céu

 and a proof that reactions have the properties enumerated above: that for a given arbitrary timeline of input events, multiple executions of the same program always react in bounded time and arrive at the same final finite memory state.

By the end of February

- Theoretical
- Core
- Engineering
- Experiments

\begin{comment}
- Include a working title and the words "Progress Report" at the top of the page.
- Use section headings in the report to simplify both the writing and reading process.
- Open the report with a "Scope and Purpose" section, where you give a condensed version of your future report’s introduction and objective.
- Always include a section entitled, for example, "Progress," which summarizes the work’s pace and progress and explains any snafus, dilemmas, or setbacks.
- Always include a section entitled, for example, "Remaining Work," which honestly assesses the work that must still be completed. Think right on the page in this section, posing questions, speculating meaningfully, exploring your options.
- Always include a section that projects the expected results. Commit to a schedule for obtaining those results if possible.
- If necessary, include a section in which you directly solicit advice from your teacher or advisor. Be forthright and professional about the nature of the advice you need.
- Keep your paragraphs short and focused—just a few paragraphs per section, typically.
- Your tone can often be straightforward and familiar—therefore, as a rule, you can use "I" and "you" freely—but do not lapse into informality.
- Avoid being overly optimistic, pessimistic, apologetic, cocky, or self-deprecating.
\end{comment}

\bibliographystyle{abbrv}
\bibliography{serra,my}
\end{document}

### 2018-08-11 Submissao/mid.tex
(>) mid evaluation

@@
 for transparent standby at the language level~\cite{ceu.lctes18.short}, both
-of which were accepted for publication.
+of which were accepted to the conference.
 %
